### 2023-05-26 manuscript/5pillars.Rmd
transfer 5pillars from https://github.com/markziemann/enrichment_recipe

---
title: "The five pillars of computational reproducibility: Bioinformatics and beyond"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: 5pillars.bib
csl: briefings-in-bioinformatics.csl
---

Mark Ziemann

Deakin University, Geelong, Australia, School of Life and Environmental Sciences.

ORCID: 0000-0002-7688-6974

Correspondance: m.ziemann@deakin.edu.au

Source: https://github.com/markziemann/5pillars/blob/main/manuscript/5pillars.Rmd

## Summary

Computational reproducibility is a simple premise in theory, but in practice it is difficult to
achieve.
Building upon past efforts and proposals to maximise reproducibility and rigour in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) code version control and sharing, (2) compute environment control,
(3) literate programming, (4) documentation and (5) FAIR data sharing.
These best practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

## Introduction

In research, computational reproducibility is the ability to use the materials from a past study
(such as data, code and documentation) to regenerate the outputs including figures and tables to
confirm the study's findings @Goodman2016-wi.
Reproducibility is only the first step towards overall reliability, for example a study may be
reproducible, but suffer from analytical problems that invalidate the results.
If a study is reproducible, then at least these analytical issues can be more easily identified and
rectified.
Therefore, reproducibility is a necessary stepping-stone that in theory should enhance the overall
reliability of computational research including replicability and robustness.
Replicability being defined as the ability to repeat the entire study and come up with similar overall
results.
Robustness refers to the ability for the findings to be generalisable to other situations
@Nosek2022-zc.

Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
research directions.
Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015-bs; @Begley2012-di].
Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
2 of 18 articles could be reproduced (11%) @Ioannidis2009-hv, bringing into question the reliability of
those studies.
Workshops conducted by NIH intramural researchers in 2018/2019 sought to reproduce five bioinformatics
studies and could not reproduce *any*, citing missing data, software and documentation
@Zaringhalam2020-ms.
A recent systamatic analysis of Jupyter notebooks in biomedical articles had similar observations, with
only 245/4169 notebooks (5.9%) giving similar results compared to the original, with irreproducibility
blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
earlier case study @Baggerly2010-nr.
Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
in practice this is rarely achieved.

A number of guides recommending enhanced computational reproducibility practices have emerged
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw].
This article, inspired by the tutorial by Peikert and BrandMaier @Peikert2021-eq, attempts to unify
these concepts into five key principles called the five pillars of reproducible computational
research (**Figure 1**).
The intended audience is bioinformatics data analysts (not tool developers), however the
principles described here could equally apply to other domains of study.

```{r,5pillars,fig.cap="Figure 1. The five pillars of computational reproducibility.",echo=FALSE,fig.height=6}

library("RColorBrewer")
cols <- brewer.pal(5, "Set2")
cols2 <- adjustcolor( cols, alpha.f = 0.7)
plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
polygon(c(-0.2,5,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)
#mtext("Five pillars of reproducible computational research", cex=1.5)

labs <- c("Code version\ncontrol &\npersistent\n sharing",
  "Compute \nenvironment \ncontrol",
  "Literate\nprogramming",
  "Documentation",
  "FAIR &\npersistent\ndata sharing")

text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
text( 5 , 0.2 , lab="End-to-end code coverage")
#text(5,7,lab="Code testing")
text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)

```

## End-to-end code coverage

End-to-end code coverage is the foundation for reproducibility as the codebase should contain all
instructions for completing a data analysis task.
This removes the need for any manual steps, which are not only time-consuming but also prone to errors.
Without end-to-end code coverage, most of the other best practices described here are not possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
somewhat overused in the life sciences.

Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.
Related to this, the code and data need to be "linked", which means the code is aware of the location
of the data and how to get it.
The code should be able to fetch the data from a publicly accessible location and conduct the downstream
processing and analysis without having to acquire the data some other way.

A caveat for end-to-end code coverage is that sometimes manual data cleaning is unavoidable.
In those cases, retaining raw and cleaned data along with a cleaning protocol is recommended.
Then the computational workflow can begin with the cleaned dataset.

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard languages like R/R Markdown, Python/Jupyter notebook or the Unix Shell.
Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
`snakemake` [@Koster2012-sq; @Molder2021-mp], `drake` [@Michael_Landau2018-kk] or
`nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to fail mid-way,
for example at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't need to be repeated.
After fixing the issue with code at step 8, re-running the analysis would pick up at step 8 again,
which saves labour and compute time.

## Literate programming

Literate programming combines "chunks" of analytical code with human readable text @Knuth1984-di.
After compilation, the resulting output document contains the code together with computational results
such as figures and tables along with contextualising explanations and narratives.

The Sweave project, conceived in 2001, was designed to give LaTeX documents embedded R code chunks and
was envisaged as a way to generate R documentation materials and generate statistical analysis reports
@Leisch2002-yc.
Around 2015, R Markdown emerged as a more popular alternative, as formatting content is simpler with
Markdown as compared to LaTeX, saving time.
An example R Markdown script and output report is given in **Figure 2**.

![Figure 2. An example R Markdown script (left) and output document (right) @RStudio2016-fw.](../img/markdownchunk.png)

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
Similarly, the [Streamlit](https://streamlit.io/) tool, which was originally invented for creating
Python-based web dashboards but is equally capable of generating data science reports
[e.g.: @Nantasenamat2023-wr].

In 2022, the company behind RStudio (Posit) released Quarto, the conceptual successor of R Markdown,
but with enhanced support for other computer languages like Python, Julia, and Observable JavaScript
@Bauer2023-eh.

Whether R Markdown, Jupyter or Quarto, literate programming offers some substantial benefits over
alternative approaches:

* The provenance of any result can be demonstrated.
The resulting document produced contains the code executed, together with the results (e.g.: a chart).
This is in contrast to a data analysis report assembled in a word processor, which due to the many
copy-paste operations required might contain errors, or version mismatches.

* It saves time.
For the case where a report needs to be run routinely, this would involve significant copy-paste to
be assembled "manually" with a word processor.
Using a literate script would mean only minimal changes are required between iterations @Perkel2022-af.

* It accommodates extensive documentation.
This allows the analyst to include rich descriptions of scientific works.
For example, a report may contain background, methods, results, discussion and references.
Not only text, but various objects can be included like links to other resources,
tables, images, videos, etc.
This means it is possible to author an entire journal articles using literate programming.

* Outputs are arranged.
When using a regular script, an analyst might make dozens of charts which are written to the
disk with names like "chart1.png" or "model8.svg".
When the number of outputs is large, it becomes unwieldy and hard to place which plot corresponds
to which part of the script.
By embedding the outputs such as charts and tables in a document in the sequence that they were
generated, it helps the reader understand the logical steps taken in an analysis.

* Amenable to version control.
Version control is a useful best practice in software development (discussed below).

* Output reports are free from code errors.
The output document is only rendered when the entire script has been compiled without errors.
In contrast, a regular script might generate some outputs and then encounter an error,
so we are not sure whether the whole script is free of errors and completed successfully.
Therefore, it is good practice to routinely execute these scripts during the development process,
and not rely too heavily on the interactive execution of individual chunks or lines.

* Flexible output formats.
These include PDF, DOC/DOCX and HTML, the latter having some notable benefits
including better usability for mobile devices like phones and tablets, and the ability to
support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
Outputs include HTML documents, but also slideshows and even ebooks.

These features make literate programming a quantum leap for science communication in a range of
situations.
Whether this is sharing a data analysis report, giving a presentation at a meeting,
writing a research article or self-publishing an ebook, literate programming provides the
ability to construct transparent data narratives with clear provenance in a conveniently shareable
form.

Literate programming also works neatly with the concept of the "executable paper", the idea that
the data analysis underlying an entire study can be reproduced with one or a few commands
[@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
A typical genomics/bioinformatics study could involve one large literate script, or be broken down into
smaller scripts, where each one contains the code for generating a part of the article.
A multi-script approach would benefit from a master script that executes each component.
This further makes the job of reproducibility easier.

## Code version control and persistent sharing

A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A set of files under version control is called a "repository" and represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs, but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}

cols2 <- adjustcolor( cols, alpha.f = 0.4)

plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
polygon(c(-0.6,3,3,-0.6),c(0.8,0.8,5.2,5.2), col=cols2[1], border = NA)
polygon(c(-0.2,2.8,2.8,-0.2),c(1,1,2,2), col=cols[1], border = NA)
polygon(c(-0.2,2.8,2.8,-0.2),c(4,4,5,5), col=cols[1], border = NA)

polygon(c(3.1,6.7,6.7,3.1),c(0.8,0.8,5.2,5.2), col=cols2[2], border = NA)
polygon(c(3.3,6.5,6.5,3.3),c(1,1,2,2), col=cols[2], border = NA)
polygon(c(3.3,6.5,6.5,3.3),c(4,4,5,5), col=cols[2], border = NA)

polygon(c(6.8,10.2,10.2,6.8),c(0.8,0.8,5.2,5.2), col=cols2[3], border = NA)
polygon(c(7,10,10,7),c(1,1,2,2), col=cols[3], border = NA)
polygon(c(7,10,10,7),c(4,4,5,5), col=cols[3], border = NA)

polygon(c(2.5,7.3,7.3,2.5),c(7,7,8.5,8.5), col=cols[4], border = NA)

labs <- c("Developer 1",
  "Local Repo",
  "Working Files",
  "Developer 2",
  "Local Repo",
  "Working Files",
  "Developer 3",
  "Local Repository",
  "Working Files")

text( c(1.2,1.2,1.2,4.9,4.9,4.9,8.5,8.5,8.5), c(0,1.5,4.5,0,1.5,4.5,0,1.5,4.5), labels=labs )
text(4.9,7.7,labels="Main Repository",cex=1.5)

arrows( 1, 2, 1, 4 ,lwd=3)
arrows( 1.4, 4, 1.4, 2 ,lwd=3)
arrows( 4.7, 2, 4.7, 4 ,lwd=3)
arrows( 5.1, 4, 5.1, 2 ,lwd=3)
arrows( 8.3, 2, 8.3, 4 ,lwd=3)
arrows( 8.7, 4, 8.7, 2 ,lwd=3)
arrows( 1, 5, 2.5, 7 ,lwd=3)
arrows( 2.9, 7 , 1.4, 5 ,lwd=3)
arrows( 4.7, 5, 4.7, 7 ,lwd=3)
arrows( 5.1, 7 , 5.1, 5 ,lwd=3)
arrows( 7.3, 7, 8.7, 5 ,lwd=3)
arrows( 8.3, 5, 6.9, 7 ,lwd=3)

text(c(0.7,4.4,8) ,c(3,3,3), labels="commit",srt=90)
text(c(1.7,5.4,9) ,c(3,3,3), labels="update",srt=-90)
text(c(1.5) ,c(6.2), labels="push",srt=50)
text(c(2.4) ,c(5.9), labels="pull",srt=50)
text(c(4.4) ,c(6), labels="push",srt=90)
text(c(5.4) ,c(6), labels="pull",srt=-90)
text(c(7.3) ,c(5.9), labels="push",srt=-60)
text(c(8.2) ,c(6.2), labels="pull",srt=-60)

```

Computational researchers can benefit significantly by adopting DVCS into their routine
practices.
Below I outline some of the benefits.

* Keeps a complete history of all code changes over time.
We always know which version of the script is the most up to date.
We can inspect and execute the code and reproduce the analysis at any particular point in the past.
Commit messages are retained, which gives a record of the motivations, purpose and person behind each
modification.

* Helps collaboration and project management.
Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
from team members who may be working asynchronously in different time zones and on different servers.
It also makes the user's work searchable, which is convenient when code snippets from a project
need to be reused for a new project.

* Helps resolve issues.
These centralised platforms also enable discussions of code issues such as bugs and feature requests.
In bioinformatics teams, issues can be used to track the progress of a project through its milestones
and allow team members to raise potential issues with the code.
This also allows the scientific community to raise an issue with the software maintainers.
For example, this feature is used extensively during the peer review process for the *Journal of
Open-Source Software* @Editorial2018-kt.

* Supports documentation best practices.
Books, educational resources and other data types are well served using DVCS (see below).

* Makes works easily shareable.
Work is easier to reproduce because the source folder contains a complete workflow with code linked
to data and metadata, so it doesn't need any modifications before reproduction.
It is relatively easy to add a software license with a DVCS like `GitHub`, so consumers can understand
the conditions of using it.

* Protect against code loss.
Computers sometimes fail and we sometimes delete or overwrite important files.
If code is lost and unrecoverable it can be a significant cost to replace.
Although DVCSs are not a back-up solution, they do add a layer of protection to the
code.
Preserving a copy of the repository on each team members' computer in addition to the central
repository means the code can be easily recovered if any one of those computers fails.

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution and
might not survive the "ten-year challenge" of code reproducibility @Perkel2020-io.
To achieve that, the code needs to be deposited to a long-term repository.
Zenodo and Figshare repositories are both good options for long-term archiving of software, and the
link to these should be provided in the respective journal article/pre-print.

`git` is typically used at the command line, however it is also integrated into Rstudio, JupyterLab and
VS Code, and there are also several `git` clients with graphical interfaces which better allow inspection
of code changes.

## Compute environment control

Most popular software undergoes regular updates to patch bugs and add new features.
Bioinformatics software is no different, and it is well known that such changes have the
potential to affect results @Seyednasrollah2015-cc.
This is why it is best practice to report the exact version of all programs used in an analysis
(and packages therein) and even make archival copies for future reference @Sandve2013-lk.

In R, such reporting is possible using the `sessionInfo()` command,
while for Python this is possible using the `session_info` package.
Using literate programming and sharing output documents ensures that a record of this important
information is made available.

Although regular software updates are overall a good thing, it poses a problem for future
reproducibility.
A researcher trying to reproduce a ten-year-old study in R v3.0 could have a headache, as they would need
to roll back their R version, and possibly their operating system as well, as R and other languages
require certain system libraries for low-level routines.

To avoid this, a virtual machine (VM) could be used to run a system-in-a-system.
This means that the "host" machine can run another "guest" operating system with the right R version,
without needing to change the host R version.
While this provides good reproducibility, setting up an environment with a ten-year-old operating system,
R and packages would take a few hours to accomplish.
Researchers could take a snapshot of their VM system and share it to help reproducibility and
auditability, however the size of these images is relatively large due to the fact it contains the OS,
software stack and project data.
Moreover, the performance of computation in the guest system is typically slower than when run directly
on the host.

Another potential solution to this is to use a package/environment management system
(e.g.: [Conda](https://conda.org/) or [Guix](https://guix.gnu.org/)).
These allow users to create, manage, and deploy software packages,
dependencies, and environments across different computing platforms @Perkel2023-cv.
Conda was initially developed to solve the problem of package dependency management in Python,
but it now supports many other programming languages, including R, C/C++, Java, and others.
Conda allows researchers to create isolated environments with specific versions of
packages, so users can have different versions of R or Python in different environments
on the same host system.
Conda environments have limitations around portability and consistency, as such environments
are tied to specific operating systems and hardware architecture.
Guix on the other hand, gives system-level reproducibility, but has a smaller community and ecosystem
of supported packages.

Containers are an attempt to resolve some of the downsides of VMs and package managers.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system.
In the example shown in **Figure 4**, the six containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
containerisation system for web development and research.
Due to its early adoption, Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images called DockerHub.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp].

![Figure 4. Architecture of containers (left) compared to VMs (right) @noauthor_2021-lt.](../img/docker-containerized-and-vm-transparent-bg.webp)

We are also watching with interest the early development of WebAssembly (Wasm) based approaches for
compute environment control.
The ability to reproduce operating systems, programming languages and workflows in the browser
opens up intriguing possibilities for more widespread reproduction and auditing without the need to
install any software at all.
For example, it is now possible to embed a Wasm-enabled R terminal inside a Quarto HTML document, which
enables in-browser reproducibility of R workflows @Balamuta2023-ma.

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
As they are a recipe, these are small files which are easily shared in the project code repository.
Docker images are larger and cannot be shared in the code repository.
They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
A better option for bioinformatics workflows would be to use the BioContainers registry
@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
*Global Alliance for Genomics and Health* that will underpin its longevity.
Still, these are not considered permanent archival services, so it is advised to deposit the image
used in an article to a suitable persistent long-term archive (discussed below).
The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
added to the project source code repository.

## FAIR and persistent data sharing

Without data sharing, computational research is not reproducible nor auditable.
Lack of data sharing is one of the key reasons why research is irreproducible
[@Ioannidis2009-hv, @Archmiller2020-er ].
Data sharing is also one of the key features of "open science", which is characterised as science
that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].
Indeed, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
In addition to facilitating reproduction and auditing, sharing enables reuse in other contexts.
Data reuse increases efficiency, as it prevents redundant research expenditure and facilitates new
research ideas that were previously impossible [@Wilkinson2016-qq].
In terms of research rigour, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
In light of this, a research article without supporting data and code is much like a press release or
advertisement where the claims made cannot be verified as true or not.

Although it is common to see journal articles with "Data available upon reasonable request", this is less
than ideal.
A systematic study of such data availability statements found that the data was successfully provided
in just 6.8% of requests [@Gabelica2022-ci].
In the rare cases that data is shared in accordance with the data availability statement, it poses
a burden in terms of labour to lodge and respond to such requests, and if the data is large, may incur
an additional cost for hosting/transferring these large files.

The field of genomics has a long history of data sharing @Brazma2001-yk, which is supported by
policy leadership by funding bodies
[@National_Health_and_Medical_Research_Council2019-yx;
@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
(eg [@Packer2018-ru; @Hanson2011-ub]), although there are some challenges.

Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.
This is a well-founded concern for identifiable data, however once identifying features are removed,
it is nearly impossible to reidentify genomic data @Venkatesaramani2021-tn unless the attackers have
preexisting genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
Therefore, sharing deidentified genomic data and metadata is likely to carry little risk to participant
privacy.
Still, much human genomic data and metadata resides in restricted databases due to the scope of
data sharing that was agreed to in the consent process.
For example, the consent form may only allow sharing of data to researchers investigating certain
diseases, which means a vetting process must be undertaken before sharing data with external
investigators.
In the interests of reuse and reproducibility, future studies could consider a consent form that allows
more broad sharing of deidentified data.

In order to maximise the value of shared data for reuse and reproducibility, it needs to be Findable,
Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
These principles are summarised in **Figure 5**.

![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb.](../img/gr2.jpg)

High throughput biology has already built a number of data repositories for specific data types
conforming to FAIR principles (e.g.: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
However other domains lack well known specialised organised data repositories, such as evolutionary
biology, ecology, cell biology, social sciences and computing.
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
repository that accepts data from these fields [@Pampel2013-ep].
If no specialised repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered.
Moreover, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

Key recommendations include:

1. Depositing data to a specialised repository if possible, otherwise, a general-purpose repository.

2. Avoid commodity cloud storage as these are impermanent.

3. Avoid large supplementary data files accompanying journal articles as these are less findable and
accessible.

4. Provide detailed metadata.

5. Preferably archive and share raw data and use existing standards for the discipline.

6. Use file formats that are compatible with many different types of software; e.g.: comma-separated
values (CSV) files are preferred over Excel files.

Once the quality of the data set has been established researchers may consider depositing it
to a data repository early, before publicisation, as it has some advantages.
Most data repositories have an optional 12-month embargo period so researchers can share publicly
at a later date once they're happy with the overall project.
The data repository acts as an additional backup to the host institutions' own in case of data loss
or calamity.
By assuming the data and metadata can be sourced from a public location, the workflow developers ensure
that the code they share won't suffer from "file not found errors" during reproduction.
This ensures code and data are linked, which has been a priority for reproducibility [@Peng2011-ia],
and a source of problems for shared code [@Samuel2022-ha].

In addition to the experimental data, reference data that is critical to a project should be archived.
For example, in the field of genomics, gene function annotation sets are relied upon for a common
procedure called "pathway analysis" @Khatri2012-ye.
These annotation sets are regularly updated, but versions are rarely reported [@Wijesooriya2022-ly]
and finding historical data versions is sometimes difficult.
In order to guarantee future reproducibility, snapshots of those reference data should be archived
and shared if the licence permits it.

## Documentation

Documentation is the glue that binds a data science project together.
The manuscript/published article is the central artefact that outlines the research project and
links to the supporting materials.
From a reproducibility standpoint, the methods section is the most critical part.
It should be detailed enough so that other researchers can understand and replicate the
experiments/analysis and yield similar results.
The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
comprehensive methodological reporting in the life sciences [@Mellor2022-up].

Researchers should consider depositing their thorough laboratory/*in silico* protocols as separate
outputs to relevant repositories such as [protocols.io](https://www.protocols.io/),
[RIO Journal](https://riojournal.com/) or [Zenodo](https://zenodo.org/), minting a Digital Object
Identifier (DOI) that can be cited in the manuscript.
This is particularly helpful when there are strict word limits on articles.

The manuscript should have clear links to supporting materials including datasets, software code and
other resources like computational environments.
The code repository should have a detailed `README` file which plays a critical role in reproducibility.
It should outline the purpose of the code/software/overall project and how it relates to the
manuscript/article.
For example, some articles may rely on more than one repository for various parts of the research
project, so these need to be explained.

The `README` should outline exactly what is required in order to reproduce the
analysis, including the requirements and the instructions to reproduce.
Typically it is written in the markdown format, which should be familiar
Hardware requirements such as RAM, CPU architecture, GPU needs, etc need to be outlined.
Software requirements, such as operating system, container engine, workflow manager,
etc, also need to be described.
The exact instructions for reproduction should be clearly outlined.
This should include what the output objects of the analysis are, and instructions on how to access the
results and what should be expected.
Literate programming allows for thorough documentation of workflow methods and results which makes
it more accessible for reproducers to comprehend the workflow details, which is not possible with
standard code comments [@Rule2019-eh].
These instructions should be tested whenever changes are made to the codebase.

As outlined above, one of the goals is to reduce the complexity of reproduction, and using a master
script is a simple way to do this.
It also makes the README documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu linux install [@Ziemann2023-nm].
By minimising the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.

In addition, the `README` should outline the contents of the repository, how users can contribute to
the codebase and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.

The need for documentation extends to the other supporting information.
The deposited data set needs thorough descriptive metadata to provide context.
The container image and code snapshot should be deposited to a persistent repository with
a thorough description of their purpose.
Both the data and container image descriptions need to contain links to the DOI of the manuscript,
so that machines or individuals browsing these resources can understand the context of these objects
relative to the overall research project.

## Limitations and challenges

Due to the constantly developing software landscape, this review is not designed to give strong
recommendations on the exact combination of tools or resources to use in computational research,
but to set out key principles and highlight some useful solutions that users can incorporate
into their toolkit to improve the overall reproducibility of their work.

When these principles are not integrated correctly and checked, problems readily emerge
[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
This is why regular code testing is another recommended best practice, especially after making updates
to the code or data [@Rule2019-eh; @The_Turing_Way_Community2019-ig].
It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
scientific community would fare reproducing such a workflow.

Moreover, some have suggested adapting software development
principles of unit tests and continuous integration/continuous delivery.
This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
made to a project's code base [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
If the workflow executes successfully and tests are passed, the repository can be decorated with a
"badge" to inform users of the lack of errors in the code.
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves.
Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
"temple" of reproducible research.

The key technological advances enabling the five pillars approach have been around for about a decade,
yet progress on integrating these into computational research publications has been minimal.
Part of the problem is the incentive structures in academia, with funding bodies and journals are
focused on novel research directions rather than confirming previous findings
[@Kaelin2017-id; @Munafo2017-ty].
Indeed the number of systematic replication studies in bioinformatics remains small
(e.g.: [@Ioannidis2009-hv; ; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
problem actually is.

A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help, but the most scalable solution would be to
provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.

## Acknowledgements

I am grateful to the individuals who provided feedback on an earlier draft of the concept including
Dr Aziz Khan (Stanford University),
Dr Sriganesh Srihari (QIMR Berghofer Medical Research Institute),
Dr Barbara Zdrazil (EMBL-EBI)

* Pierre Poulain @pierrepo

* Irene López Santiago @IreneLopSci

## Bibliography

### 2023-05-26 main.sh
add main script

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+git clone https://github.com/markziemann/5pillars.git
+
+cd 5pillars
+
+Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer"))'
+
+Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'

### 2023-05-30 manuscript/5pillars.Rmd
adding collabs
link decay addressed #1

@@
 that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].
-Indeed, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
 In addition to facilitating reproduction and auditing, sharing enables reuse in other contexts.
@@
 
-2. Avoid commodity cloud storage as these are impermanent.
+2. Avoid commodity cloud storage as these are impermanent and susceptible to link decay @Hennessey2013-tj.
 
@@
 
-4. Provide detailed metadata.
+4. Preferably archive and share raw data and use existing standards for the discipline.
 
-5. Preferably archive and share raw data and use existing standards for the discipline.
+5. Use file formats that are compatible with many different types of software; e.g.: comma-separated
+values (CSV) files are preferred over Excel files.
 
-6. Use file formats that are compatible with many different types of software; e.g.: comma-separated
-values (CSV) files are preferred over Excel files.
+6. Provide detailed metadata; e.g.: sample descriptions that match the manuscript; describe the columns
+in tabular data.
 
@@
 
-* Pierre Poulain @pierrepo
+* Pierre Poulain @pierrepo (elevate to author)
 

### 2023-05-30 manuscript/5pillars.Rmd
new automated build script

@@
 title: "The five pillars of computational reproducibility: Bioinformatics and beyond"
-author: "Mark Ziemann"
+author: "Mark Ziemann, Samuel S. Minot, Pierre Poulain, Anusuiya Bora"
 date: "`r Sys.Date()`"
@@
 
-Mark Ziemann
+Mark Ziemann<sup>1*</sup>, Samuel S. Minot<sup>2</sup>, Pierre Poulain<sup>3</sup>, Anusuiya Bora<sup>1</sup>
 
-Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
+1. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.
 
-ORCID: 0000-0002-7688-6974
+2. Fred Hutchinson Cancer Center, Seattle, USA.
+
+3. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
 
@@
 To achieve that, the code needs to be deposited to a long-term repository.
-Zenodo and Figshare repositories are both good options for long-term archiving of software, and the
+
+[Zenodo](https://zenodo.org/) and [Software Heritage](https://www.softwareheritage.org/) repositories
+are both good options for long-term archiving of software, and the
 link to these should be provided in the respective journal article/pre-print.
@@
 
-* Pierre Poulain @pierrepo (elevate to author)
-
 * Irene López Santiago @IreneLopSci

### 2023-05-30 test_build.sh

@@ -1,0 +1,3 @@
+docker run -it rocker/rstudio:4.3 bash -c "wget https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
+docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html .
+cp 5pillars.html ~/public/5pillars/

### 2023-05-30 manuscript/5pillars.Rmd
#6 resolved

@@
 to roll back their R version, and possibly their operating system as well, as R and other languages
-require certain system libraries for low-level routines.
+require certain system dependencies for low-level routines @Vallet2022-no.
 
@@
 focused on novel research directions rather than confirming previous findings
-[@Kaelin2017-id; @Munafo2017-ty].
+[@Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
 Indeed the number of systematic replication studies in bioinformatics remains small
-(e.g.: [@Ioannidis2009-hv; ; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
+(e.g.: [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
 problem actually is.

### 2023-05-30 manuscript/5pillars.Rmd
MyST and JupyterLite

@@
 R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
+The MyST{NB} (short for Markedly Structured Text notebook) is built on top of Jupyter and has some
+notable advantages around flexibility of content that make it an attractive choice for an executable
+paper authoring [@DuPre2022-bk].
+
 Similarly, the [Streamlit](https://streamlit.io/) tool, which was originally invented for creating
@@
 install any software at all.
-For example, it is now possible to embed a Wasm-enabled R terminal inside a Quarto HTML document, which
-enables in-browser reproducibility of R workflows @Balamuta2023-ma.
+Two notable examples of this include JupyterLite, a version of JupyterLab in early development that
+runs entirely in the browser @Tuloup2021-le, and WebR-enabled interactive code blocks in Quarto HTML
+documents @Balamuta2023-ma.
 

### 2023-05-30 manuscript/5pillars.Rmd
more refs

@@
 [ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
-@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw].
-This article, inspired by the tutorial by Peikert and BrandMaier @Peikert2021-eq, attempts to unify
-these concepts into five key principles called the five pillars of reproducible computational
-research (**Figure 1**).
+@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw;
+@DuPre2022-bk].
+Inspired by these, this article attempts to unify the main principles into the
+*five pillars of reproducible computational research* (**Figure 1**).
 The intended audience is bioinformatics data analysts (not tool developers), however the
@@
 In light of this, a research article without supporting data and code is much like a press release or
-advertisement where the claims made cannot be verified as true or not.
+advertisement where the claims made cannot be verified as true or not [@Donoho2010-yz, @DuPre2022-bk].
 
@@
 Dr Sriganesh Srihari (QIMR Berghofer Medical Research Institute),
-Dr Barbara Zdrazil (EMBL-EBI)
-
-* Irene López Santiago @IreneLopSci
+Dr Barbara Zdrazil (EMBL-EBI) and Dr Irene López Santiago (EMBL-EBI).
 

### 2023-05-30 manuscript/5pillars.Rmd
acknowledgements

@@
 paper authoring [@DuPre2022-bk].
-
-Similarly, the [Streamlit](https://streamlit.io/) tool, which was originally invented for creating
-Python-based web dashboards but is equally capable of generating data science reports
-[e.g.: @Nantasenamat2023-wr].
 
@@
 Indeed the number of systematic replication studies in bioinformatics remains small
-(e.g.: [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
+[@Zaringhalam2020-ms ; @Ioannidis2009-hv; @Samuel2022-ha] so we don't have much of a grasp as to how severe the
 problem actually is.
@@
 
-I am grateful to the individuals who provided feedback on an earlier draft of the concept including
+We are grateful to the individuals who provided feedback on an earlier draft of the concept including
 Dr Aziz Khan (Stanford University),
 Dr Sriganesh Srihari (QIMR Berghofer Medical Research Institute),
-Dr Barbara Zdrazil (EMBL-EBI) and Dr Irene López Santiago (EMBL-EBI).
+Dr Barbara Zdrazil (EMBL-EBI) and
+Dr Irene López Santiago (EMBL-EBI).
+
+This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
+platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+
+## Availability of materials
+
+Code repository: https://github.com/markziemann/5pillars
 

### 2023-06-01 manuscript/5pillars.Rmd
more text editing aroung guix

@@
 
-Another potential solution to this is to use a package/environment management system
-(e.g.: [Conda](https://conda.org/) or [Guix](https://guix.gnu.org/)).
-These allow users to create, manage, and deploy software packages,
-dependencies, and environments across different computing platforms @Perkel2023-cv.
-Conda was initially developed to solve the problem of package dependency management in Python,
-but it now supports many other programming languages, including R, C/C++, Java, and others.
-Conda allows researchers to create isolated environments with specific versions of
-packages, so users can have different versions of R or Python in different environments
-on the same host system.
-Conda environments have limitations around portability and consistency, as such environments
-are tied to specific operating systems and hardware architecture.
-Guix on the other hand, gives system-level reproducibility, but has a smaller community and ecosystem
-of supported packages.
-
 Containers are an attempt to resolve some of the downsides of VMs and package managers.
 Container images can be thought of as similar to VMs, but are more lightweight as they
-share parts of the host operating system.
+share parts of the host operating system [@Nust2020-cr].
 In the example shown in **Figure 4**, the six containerised applications share the same operating
@@
 
+Another potential solution to this problem is to use a package/environment management system such as
+[Conda](https://conda.org/) or [Guix](https://guix.gnu.org/).
+These allow users to create, manage, and deploy software packages,
+dependencies, and environments across different computing platforms @Perkel2023-cv.
+Conda was initially developed to solve the problem of package dependency management in Python,
+but it now supports many other programming languages, including R, C/C++, Java, and others.
+Conda allows researchers to create isolated environments with specific versions of
+packages, so users can have different versions of R or Python in different environments
+on the same host system.
+Conda environments have limitations around portability and consistency, as such environments
+are tied to specific operating systems and hardware architecture.
+
+Guix is described as a "functional package manager" and has the unique benefit of giving
+system-level build reproducibility [@Vallet2022-no].
+This is a notable advantage, as Docker images are not guaranteed to build reproducibly in the future
+due to link decay.
+Moreover, Guix can build and output environments as Docker-compatible images which solves the
+provenance problem of such environments [@GNU_Project_Volunteers2022-bt].
+These features are being used in the bioinformatics sphere to create highly reproducible pipelines
+that work equally well on personal as well as shared computer systems @Wurmus2018-lv.
+Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
+tutorials, illustrated by the complete absence of published Guix protocols in protocols.io, as compared
+to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).
+
 We are also watching with interest the early development of WebAssembly (Wasm) based approaches for
@@
 make the environment.
-As they are a recipe, these are small files which are easily shared in the project code repository.
-Docker images are larger and cannot be shared in the code repository.
-They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
+Similarly, Guix environments rely on two files: `channels.scm` and `manifest.scm`.
+Docker images are built using a `Dockerfile` instruction set.
+These are small files which are easily shared in the project code repository.
+To enable rapid reproduction, making Docker images available is recommended.
+As these are often very large they cannot be shared in the code repository; rather
+they are commonly deposited to a Docker image registry such as DockerHub.
+One must keep in mind that availability of images is dependent on the
 commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
-A better option for bioinformatics workflows would be to use the BioContainers registry
+To ensure long-term availability, it would be advised to deposit images to the BioContainers registry
 @Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
@@
 
+## Conclusion
+
+Reproducibility is an essential feature of reliable computational research yet current practices
+are falling short of this goal.
+Our recommendations for workflow-based research articles are summarised as:
+
+0. Encapsulate the entire workflow with end-to-end code, executable with a master script.
+
+1. Share the code repository and deposit a snapshot to an archive.
+
+2. Share a controlled environment; options include a Docker or virtual machine image or
+configuration for Conda or Guix.
+
+3. Use literate scripting tools like R Markdown or Jupyter notebooks to generate all outputs,
+such as charts and tables.
+
+4. Document the workflow sufficiently for reproduction, especially the repository `README`.
+
+5. Share research data and metadata according to FAIR principles.
+
 ## Acknowledgements
 
-We are grateful to the individuals who provided feedback on an earlier draft of the concept including
+We are grateful to individuals who provided feedback on an earlier draft of the concept including
 Dr Aziz Khan (Stanford University),

### 2023-06-01 test_build.sh
README and test_build

@@ -1,3 +1,6 @@
+#!/bin/bash
 docker run -it rocker/rstudio:4.3 bash -c "wget https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
-docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html .
-cp 5pillars.html ~/public/5pillars/
+docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html . \
+  && echo Build successful \
+  || echo Error Build Failed
+[[ $(hostname) == "ziemann-01" ]] && cp 5pillars.html ~/public/5pillars/

### 2023-06-01 manuscript/5pillars.Rmd
covletter

@@
 @Zaringhalam2020-ms.
-A recent systamatic analysis of Jupyter notebooks in biomedical articles had similar observations, with
+A recent systematic analysis of Jupyter notebooks in biomedical articles had similar observations, with
 only 245/4169 notebooks (5.9%) giving similar results compared to the original, with irreproducibility
 blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
+A similar survey of R scripts in the Harvard Dataverse repository found slightly better results,
+with 26% of scripts completing without errors [@Trisovic2022-dq].
 Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
@@
 @Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
-(eg [@Packer2018-ru; @Hanson2011-ub]), although there are some challenges.
+(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]), although there are some challenges.
 

### 2023-06-01 manuscript/5pillars.Rmd
add VM example ref

@@
 Researchers could take a snapshot of their VM system and share it to help reproducibility and
-auditability, however the size of these images is relatively large due to the fact it contains the OS,
-software stack and project data.
+auditability @Pasquier2018-nj, however the size of these images is relatively large due to the fact
+it contains the OS, software stack and project data.
 Moreover, the performance of computation in the guest system is typically slower than when run directly

### 2023-06-01 manuscript/5pillars.Rmd
typos fixed

@@
 rectified.
-Therefore, reproducibility is a necessary stepping-stone that in theory should enhance the overall
+Therefore, reproducibility is a necessary stepping-stone that should enhance the overall
 reliability of computational research including replicability and robustness.
@@
 
-A number of guides recommending enhanced computational reproducibility practices have emerged
+A number of guides recommending enhanced computational reproducibility practices have been developed
 [ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@@
 `nextflow` [@Di_Tommaso2017-yo].
-The advantage of such tools is that if the analysis were to fail mid-way,
+The advantage of such tools is that if the analysis were to terminate due to a hardware issue mid-way,
 for example at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't need to be repeated.
-After fixing the issue with code at step 8, re-running the analysis would pick up at step 8 again,
+After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
 which saves labour and compute time.
@@
 
-Whether R Markdown, Jupyter or Quarto, literate programming offers some substantial benefits over
+Whether R Markdown, Jupyter or other system, literate programming offers some substantial benefits over
 alternative approaches:
@@
 tables, images, videos, etc.
-This means it is possible to author an entire journal articles using literate programming.
+This means it is possible to author an entire journal article using literate programming.
 
@@
 Due to its early adoption, Docker now has a large community of users, extensive documentation and
-a vast collection of pre-built container images called DockerHub.
+a vast collection of pre-built container images in the DockerHub registry.
 Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
@@
 used in an article to a suitable persistent long-term archive (discussed below).
-The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
-added to the project source code repository.
 
@@
 research ideas that were previously impossible [@Wilkinson2016-qq].
-In terms of research rigour, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
+In terms of research rigour, inspecting raw data can uncover inadvertent errors and research
+integrity problems [@Miyakawa2020-yj].
 In light of this, a research article without supporting data and code is much like a press release or
@@
 analysis, including the requirements and the instructions to reproduce.
-Typically it is written in the markdown format, which should be familiar
+Typically it is written in the Markdown format, which should be familiar to those using R Markdown or
+Jupyter notebooks.
 Hardware requirements such as RAM, CPU architecture, GPU needs, etc need to be outlined.
@@
 script is a simple way to do this.
-It also makes the README documentation much simpler.
+It also makes the `README` documentation much simpler.
 For example, the pathway analysis workflow we previously developed could be reproduced and inspected
-with just six commands, starting from a new Ubuntu linux install [@Ziemann2023-nm].
+with just six commands, starting from a new Ubuntu Linux install [@Ziemann2023-nm].
 By minimising the complexity of reproduction and writing the instructions clearly in the `README`,
@@
 
-Due to the constantly developing software landscape, this review is not designed to give strong
+Due to the constantly developing software landscape, this review is not intended to give strong
 recommendations on the exact combination of tools or resources to use in computational research,
@@
 yet progress on integrating these into computational research publications has been minimal.
-Part of the problem is the incentive structures in academia, with funding bodies and journals are
+Part of the problem is the incentive structures in academia, with funding bodies and journals more
 focused on novel research directions rather than confirming previous findings
@@
 Indeed the number of systematic replication studies in bioinformatics remains small
-[@Zaringhalam2020-ms ; @Ioannidis2009-hv; @Samuel2022-ha] so we don't have much of a grasp as to how severe the
-problem actually is.
+[@Zaringhalam2020-ms ; @Ioannidis2009-hv; @Samuel2022-ha], so we don't have much of a grasp as to how
+severe the problem actually is.
 

### 2023-06-01 manuscript/5pillars.Rmd
#7 and #11

@@
 @DuPre2022-bk].
-Inspired by these, this article attempts to unify the main principles into the
+Inspired by these principles and the three pillars of open science framework, (open data, code
+and papers) [@Gorgolewski2016-hv], this article presents the
 *five pillars of reproducible computational research* (**Figure 1**).
@@
 Guix is described as a "functional package manager" and has the unique benefit of giving
-system-level build reproducibility [@Vallet2022-no].
-This is a notable advantage, as Docker images are not guaranteed to build reproducibly in the future
-due to link decay.
+bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-rx].
+This is a notable advantage, as Docker images are not verifiable nor guaranteed to build reproducibly
+in the future due to link decay.
 Moreover, Guix can build and output environments as Docker-compatible images which solves the
@@
 These features are being used in the bioinformatics sphere to create highly reproducible pipelines
-that work equally well on personal as well as shared computer systems @Wurmus2018-lv.
+that work equally well on personal as well as shared high performance computers
+[@Wurmus2018-lv; @Courtes2022-rx].
 Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
@@
 Data sharing is also one of the key features of "open science", which is characterised as science
-that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].
+that is collaborative, transparent, accessible and inclusive [@McKiernan2016-vu; @noauthor_2023-uq].
 In addition to facilitating reproduction and auditing, sharing enables reuse in other contexts.
@@
 Indeed the number of systematic replication studies in bioinformatics remains small
-[@Zaringhalam2020-ms ; @Ioannidis2009-hv; @Samuel2022-ha], so we don't have much of a grasp as to how
-severe the problem actually is.
+[@Zaringhalam2020-ms ; @Ioannidis2009-hv; @Samuel2022-ha], so the severity of the problem isn't
+well understood.
 

### 2023-06-02 manuscript/5pillars.Rmd
#11 resolved?

@@
 @DuPre2022-bk].
-Inspired by these principles and the three pillars of open science framework, (open data, code
-and papers) [@Gorgolewski2016-hv], this article presents the
+Inspired by these principles, this article presents the
 *five pillars of reproducible computational research* (**Figure 1**).
+There are clear parallels with the established the three pillars of open science framework,
+(open data, code and papers) [@Gorgolewski2016-hv].
+Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
+transparent reporting and the role of computational environments.
 The intended audience is bioinformatics data analysts (not tool developers), however the

### 2023-06-02 manuscript/5pillars.Rmd
Close #3
ack

@@
 
-Mark Ziemann<sup>1*</sup>, Samuel S. Minot<sup>2</sup>, Pierre Poulain<sup>3</sup>, Anusuiya Bora<sup>1</sup>
+Mark Ziemann<sup>1*</sup>, Samuel S. Minot<sup>2</sup>, Pierre Poulain<sup>3</sup>,
+Anusuiya Bora<sup>1</sup>
+
+<details><summary>Author information (*click to expand*)</summary><p>
 
@@
 
+| Author | ORCID |
+| --- | --- |
+| Mark Ziemann | 0000-0002-7688-6974 |
+| Samuel S. Minot | 0000-0003-1639-3905 |
+| Pierre Poulain | 0000-0003-4177-3619 |
+| Anusuiya Bora | 0009-0006-2908-1352 |
+
+</details>
+</p>
 Source: https://github.com/markziemann/5pillars/blob/main/manuscript/5pillars.Rmd
@@
 Dr Irene López Santiago (EMBL-EBI).
+Special thanks to Dr Altuna Akalin (Max Delbrück Center, Germany) and
+Dr Simon Tournier (Paris Cité University, France) for advice on Guix.
 

### 2023-06-05 manuscript/5pillars.Rmd
#12 some changes around end-to-end processes and webtools

@@
 text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
-text( 5 , 0.2 , lab="End-to-end code coverage")
+text( 5 , 0.2 , lab="End-to-end automated process")
 #text(5,7,lab="Code testing")
-text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
+#text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
 
@@
 
-## End-to-end code coverage
+## End-to-end automated process
 
-End-to-end code coverage is the foundation for reproducibility as the codebase should contain all
-instructions for completing a data analysis task.
-This removes the need for any manual steps, which are not only time-consuming but also prone to errors.
-Without end-to-end code coverage, most of the other best practices described here are not possible.
+For a bioinformatics project to be reproducible, it needs to be enshrined by code and include all
+instructions, from inspecting the raw data to generating the outputs that form the conclusions of the
+study.
+This automated process removes the need for any manual steps, which are not only time-consuming but
+also prone to errors.
+Without an end-to-end automated process, most of the other best practices described here are not
+possible.
 Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
-which would be difficult for graphical tools like spreadsheets.
+which would be difficult for graphical tools like spreadsheets or webtools.
 Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
 somewhat overused in the life sciences.
-
+While webtools are valuable for data exploration, there are worries that they undermine
+reproducibility for the sake of convenience [@Wijesooriya2022-ly].
 Transferring data between compute platforms is also discouraged.
@@
 data transfer steps which are inefficient and error-prone.
-Related to this, the code and data need to be "linked", which means the code is aware of the location
-of the data and how to get it.
+
+To facilitate an end-to-end automated process, the code and data need to be "linked", which means the
+code is aware of the location of the data and how to get it automatically @Peng2011-ia.
 The code should be able to fetch the data from a publicly accessible location and conduct the downstream
@@
 
-A caveat for end-to-end code coverage is that sometimes manual data cleaning is unavoidable.
+A caveat for end-to-end automation is that sometimes manual data cleaning is unavoidable.
 In those cases, retaining raw and cleaned data along with a cleaning protocol is recommended.
@@
 
-0. Encapsulate the entire workflow with end-to-end code, executable with a master script.
+0. Encapsulate the entire workflow in an end-to-end automated process, executable with a master script.
 

### 2023-06-06 manuscript/5pillars.Rmd

@@
 code to consumers.
-There are many such DVCSs, but `git` has emerged as the most popular
+There are many such DVCSs (subversion, git, bazaar, mercurial...), but `git` has emerged as the most popular
 solution due to its many powerful features, speed/efficiency and large community and ecosystem.
@@
 
-Although DVCSs assist with code sharing, they are not considered a long-term archiving solution and
-might not survive the "ten-year challenge" of code reproducibility @Perkel2020-io.
-To achieve that, the code needs to be deposited to a long-term repository.
-
-[Zenodo](https://zenodo.org/) and [Software Heritage](https://www.softwareheritage.org/) repositories
-are both good options for long-term archiving of software, and the
-link to these should be provided in the respective journal article/pre-print.
+Although DVCSs assist with code sharing, they are not considered a long-term archiving solution, 
+as recalled by the complete shutdown of the Google Code platform in 2015 
+and the end of Mercurial support by the Bitbucket platform in 2020. 
+The code needs to be deposited to a long-term repository.
+[Zenodo](https://zenodo.org/) and [Figshare](https://figshare.com/) are often used 
+to store the content of a repository but do not retain the history of the development of the source code.
+[Software Heritage](https://www.softwareheritage.org/) @dicosmo2017 is the universal software archive 
+which mission is to collect, preserve and share source code. Software Heritage provides 
+an intrinsic persistent identifier called SWHID that allows to cite the archived source code 
+in the respective journal article/preprint. The archival of the source code in Software Heritage
+can be performed manually or automated by a webhook within a continuous integration.
 
@@
 In R, such reporting is possible using the `sessionInfo()` command,
-while for Python this is possible using the `session_info` package.
+while for Python this is possible using the `session_info` or `watermark` packages.
 Using literate programming and sharing output documents ensures that a record of this important
@@
 [@Wurmus2018-lv; @Courtes2022-rx]. 
+Guix is probably one of the best solution so far for the "ten-year challenge" of code reproducibility @Perkel2020-io.
 Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
@@
 to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).
+Eventually, it's also worth mentionning that the Guix project uses Software Heritage
+as a fallback system to provide a package source code
+if the download of the source code from its original location failed.
 

### 2023-06-06 manuscript/5pillars.Rmd
#13 Have added excel date info and refs

@@
 Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
-somewhat overused in the life sciences.
+somewhat overused and misused in the life sciences, leading to problems like inadvertant conversion
+of gene names to dates in databases and supplementary Excel files [@Zeeberg2004-js;
+@Ziemann2016-pv; @Abeysooriya2021-uz].
 While webtools are valuable for data exploration, there are worries that they undermine

### 2023-06-06 guides/practical_guides.Rmd
practical guides

---
title: "Practical guides for implementing 'Five Pillar' principles"
author: "Mark Ziemann, Samuel S. Minot, Pierre Poulain, Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: practical_guides.bib
csl: plos-computational-biology.csl
---

Mark Ziemann<sup>1*</sup>, Samuel S. Minot<sup>2</sup>, Pierre Poulain<sup>3</sup>,
Anusuiya Bora<sup>1</sup>

<details><summary>Author information (*click to expand*)</summary><p>

1. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

2. Fred Hutchinson Cancer Center, Seattle, USA.

3. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.

Correspondance: m.ziemann@deakin.edu.au

| Author | ORCID |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Samuel S. Minot | 0000-0003-1639-3905 |
| Pierre Poulain | 0000-0003-4177-3619 |
| Anusuiya Bora | 0009-0006-2908-1352 |

</details>
</p>
Source: https://github.com/markziemann/5pillars/blob/main/manuscript/5pillars.Rmd

## Practical guides

The five pillar approach is a synthesis of over a decade of learnings around computational
reproducibility, and contains six key recommendaions and many best practices.
These may be overwhelming for any data scientists, especially beginners and novices,
so here we have tried to put together some recommended resources to enable practitioners to put these
principles into practice.
One of the questions practitioners may have is *"Where do I start?"* and
*"In what order should I learn/master these principles?"*.
The following sections have been arranged in order, and have been submitted to the
[Internet Archive](https://archive.org/) for posterity.

### Getting started with data analysis in R and Python

Learn the basics of your scripting language (R/Python/shell) so that you can do useful analyses
with it. Learn by solving problems.

* Intro to R and Rstudio

* Intro to Python and iPython IDE

* Intro to vscode

### Literate programming with R Markdown, Jupyter and Quarto

Convert those scripts into literate scripts and document them with introduction, methods,
code comments, results observations and interpretations.
Use JupyterLab (Jupyter notebook) or Rstudio (Rmarkdown) development environments to provide
graphical interfaces to the programming languages.

* https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074

* https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/

* https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html

* https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html

* https://quarto.org/docs/computations/r.html

* https://quarto.org/docs/computations/python.html

* https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/

* https://www.dataquest.io/blog/jupyter-notebook-tutorial/

* https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af

### Practical guides for `git` and GitHub

Learn how integrate version control into your development environment.

* https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/

* https://www.freecodecamp.org/news/introduction-to-git-and-github/

* https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/

* https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/

* https://andgarc.github.io/basic-git-jupyter-lesson/index.html

* https://blog.reviewnb.com/github-jupyter-notebook/

* https://andgarc.github.io/basic-git-jupyter-lesson/index.html

### Practical guides to data sharing

Learn about the FAIR principles and make the research data available in a public repository.

### Extend the scripts to make them end-to-end processes

Try to extend the scripts to make them end-to-end processes, by linking to raw data and outputting
whole figures or research articles.

### Practical guides for Conda, Guix and Docker

Control the environment.
Record the version of programming languages and packages.
Become familiar and experiment with Conda, Guix and Docker.
Select the option that works best for your project.

### Practical guides for documenting computational research

Document the process of reproduction as if you were an external researcher.
Test the instructions.

### Moving from "dev" to "prod": testing and continuous analysis

Test the reproduction process as if you were an external researcher.

### 2023-06-07 guides/practical_guides.Rmd
added more practical guides

@@
 </p>
-Source: https://github.com/markziemann/5pillars/blob/main/manuscript/5pillars.Rmd
 
-## Practical guides
+Source: https://github.com/markziemann/5pillars/blob/main/manuscript/practical_guides.Rmd
 
@@
 
-### Getting started with data analysis in R and Python
+## Getting started with data analysis in R and Python
 
@@
 
-* Intro to R and Rstudio
+### Intro to R and Rstudio
 
-* Intro to Python and iPython IDE
+* [Ebook: R for data science](https://r4ds.had.co.nz/)
 
-* Intro to vscode
+* [Video: Rstudio for the Total Beginner](https://www.youtube.com/watch?v=FIrsOBy5k58)
 
-### Literate programming with R Markdown, Jupyter and Quarto
+* [Ebook: YaRrr! The Pirate’s Guide to R](https://bookdown.org/ndphillips/YaRrr/)
+
+### Shell and linux systems
+
+* [Ebook: Bioinformatics Data Skills](https://vincebuffalo.com/book/)
+
+* [Book: Unix Shell Programming (ISBN:0672324903)](https://www.amazon.com/Unix-Shell-Programming-Stephen-Kochan/dp/0672324903)
+
+### Intro to Python
+
+* [Tutorial: Beginner's Guide to Python](https://wiki.python.org/moin/BeginnersGuide)
+
+* [Ebook: The Absolute Beginner's Guide to Python Programming](https://doi.org/10.1007/978-1-4842-8716-3)
+
+* [Video: Python Tutorial for Absolute Beginners](https://www.youtube.com/watch?v=Z1Yd7upQsXY&t=19s)
+
+### Intro to VS Code
+
+* [Tutorial and video: Getting Started with Visual Studio Code](https://code.visualstudio.com/docs/introvideos/basics)
+
+## Literate programming with R Markdown, Jupyter and Quarto
 
@@
 
-* https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074
+* [Tutorial: Beginner Guide to Markdown](https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074)
 
-* https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/
+* [Tutorial: Getting Started with R Markdown — Guide and Cheatsheet](https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/)
 
-* https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html
+* [Ebook Chapter: Reproducible Workflow with R Markdown](https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html)
 
-* https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html
+* [Tutorial: Reproducibility and R Markdown](https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html)
 
-* https://quarto.org/docs/computations/r.html
+* [Quick guide: R based literate scripts with Quarto](https://quarto.org/docs/computations/r.html)
 
-* https://quarto.org/docs/computations/python.html
+* [Quick guide: Python based literate scripts with Quarto](https://quarto.org/docs/computations/python.html)
 
-* https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/
+* [Tutorial: Jupyter/IPython Notebook Quick Start Guide](https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/)
 
-* https://www.dataquest.io/blog/jupyter-notebook-tutorial/
+* [Tutorial: How to Use Jupyter Notebook](https://www.dataquest.io/blog/jupyter-notebook-tutorial/)
 
-* https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af
+* [Tutorial: The Ultimate Beginner’s Guide to Jupyter Notebooks](https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af)
 
-### Practical guides for `git` and GitHub
+## Practical guides for `git` and GitHub
 
@@
 
-* https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/
+* [Tutorial: GitHub and Git for Beginners](https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/)
 
-* https://www.freecodecamp.org/news/introduction-to-git-and-github/
+* [Tutorial: How to Use Git and GitHub – Introduction for Beginners](https://www.freecodecamp.org/news/introduction-to-git-and-github/)
 
-* https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/
+* [Tutorial: How to use Git with R and RStudio](https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/)
 
-* https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/
+* [Tutorial: How to link and use Git with RStudio](https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/)
 
-* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
+* [Tutorial: Introduction to Using Git in Jupyter](https://andgarc.github.io/basic-git-jupyter-lesson/index.html)
 
-* https://blog.reviewnb.com/github-jupyter-notebook/
+* [Tutorial: How to use Git / GitHub with Jupyter Notebook](https://blog.reviewnb.com/github-jupyter-notebook/)
 
-* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
+* [Tutorial: Introduction to Using Git in Jupyter](https://andgarc.github.io/basic-git-jupyter-lesson/index.html)
 
-### Practical guides to data sharing
+## Practical guides to data sharing
 
@@
 
-### Extend the scripts to make them end-to-end processes
+* [Guide: Dryad good data practices](https://datadryad.org/stash/best_practices)
+
+* [Guide: Best practice for managing your outputs on Figshare](https://help.figshare.com/article/best-practice-for-managing-your-outputs-on-figshare)
+
+* [Guide: Introduction to data sharing](https://zenodo.org/badge/DOI/10.5281/zenodo.3518692.svg)
+
+* [Guide: FAIR Data (according to Australian Research Data Commons)](https://ardc.edu.au/resource/fair-data/)
+
+## Extend the scripts to make them end-to-end processes
 
@@
 
-### Practical guides for Conda, Guix and Docker
+* [Ebook: The Practice of Reproducible Research - Case Studies and Lessons from the Data-Intensive Sciences](http://www.practicereproducibleresearch.org/)
+
+* [Ebook: The Turing Way - Guide for Reproducible Research](https://the-turing-way.netlify.app/reproducible-research/reproducible-research.html#)
+
+* [Ebook: Building reproducible analytical pipelines with R](https://raps-with-r.dev/)
+
+* [Tutorial: A short introduction to drake (An R based workflow manager)](https://data.nozav.org/posts/2019-a-short-introduction-to-drake/)
+
+* [Tutorial: An example Snakemake workflow](https://snakemake.readthedocs.io/en/stable/tutorial/basics.html)
+
+* [Tutorial compendium: Learn Nextflow in 2023](https://snakemake.readthedocs.io/en/stable/tutorial/basics.html)
+
+## Practical guides for Conda, Guix and Docker
 
@@
 
-### Practical guides for documenting computational research
+* [Tutorial: Getting started with conda](https://conda.io/projects/conda/en/latest/user-guide/getting-started.html)
+
+* [Video tutorial: Anaconda Beginners Guide for Linux and Windows](https://www.youtube.com/watch?v=MUZtVEDKXsk)
+
+* [Tutorial: Running RStudio Server with Docker](https://davetang.org/muse/2021/04/24/running-rstudio-server-with-docker/)
+
+* [Tutorial: Creating a Docker Image for your Application](https://www.stereolabs.com/docs/docker/creating-your-image/)
+
+* [Documentation: Getting started with GNU Guix](https://guix.gnu.org/manual/en/html_node/Getting-Started.html)
+
+## Practical guides for documenting computational research
 
@@
 
-### Moving from "dev" to "prod": testing and continuous analysis
+* [Guide: How to write an Awesome README](https://towardsdatascience.com/how-to-write-an-awesome-readme-68bf4be91f8b)
+
+* [Example: README for "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker"](https://github.com/aaronpeikert/reproducible-research/blob/master/README.md)
+
+## Moving from "dev" to "prod": testing and continuous analysis
 
 Test the reproduction process as if you were an external researcher.
+
+* [Ebook: Building reproducible analytical pipelines with R](https://raps-with-r.dev/)
+
+* [Ebook: The Turing Way - Guide for Reproducible Research](https://the-turing-way.netlify.app/reproducible-research/reproducible-research.html#)
+
+* [Tutorial: Continuous Analysis](https://greenelab.github.io/continuous_analysis/)

### 2023-06-07 manuscript/5pillars.Rmd
finishing touches

@@
 way to reduce the number of commands required for reproduction.
-This can be coded in standard languages like R/R Markdown, Python/Jupyter notebook or the Unix Shell.
+This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
+Unix Shell.
+Using free and open source software ensures that the research is accessible to the greatest audience,
+as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
+Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
+the scientific community.
 Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
@@
 Researchers could take a snapshot of their VM system and share it to help reproducibility and
-auditability @Pasquier2018-nj, however the size of these images is relatively large due to the fact
+auditability [ @Hurley2015-hk ;@Pasquier2018-nj], however the size of these images is relatively large due to the fact
 it contains the OS, software stack and project data.
@@
 experiments/analysis and yield similar results.
+Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
+their reproducibility [@Simoneau2021-kp].
 The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
-comprehensive methodological reporting in the life sciences [@Mellor2022-up].
+comprehensive methodological reporting in the life sciences [@Mellor2022-up], and "Minimum Information
+About a Bioinformatics Investigation" guidelines describe good reporting practices for computational
+biology research [@Tan2010-is].
 
@@
 yet progress on integrating these into computational research publications has been minimal.
-Part of the problem is the incentive structures in academia, with funding bodies and journals more
-focused on novel research directions rather than confirming previous findings
-[@Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
+Part of the problem is the incentive structures in academia, with funding bodies and journals
+largely focused on novel research directions rather than confirming previous findings
+[@Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
 Indeed the number of systematic replication studies in bioinformatics remains small
@@
 More such protocols are needed to make extreme reproducibility mainstream.
+A curated set of tutorials is provided in the Supplement.
 

### 2023-06-08 guides/practical_guides.Rmd
#15 Added Github actions
spelling, editing, BiB formatting

@@
 
-### Intro to R and Rstudio
+### Introduction to R and Rstudio
 
@@
 
-### Shell and linux systems
+### Introduction to Unix and Shell programming
 
@@
 
-### Intro to Python
+### Introduction to Python
 
@@
 
-### Intro to VS Code
+### Introduction to VS Code
 
@@
 
-Test the reproduction process as if you were an external researcher.
+Continuous analysis involves automatic execution and testing of code, which is important when changes
+are made to code or data.
+Results generated in HTML are automatically updated.
+Badges on the repository indicate the status of the codebase in terms of completing with or without
+errors.
 
@@
 * [Tutorial: Continuous Analysis](https://greenelab.github.io/continuous_analysis/)
+
+* [Documentation: GitHub Actions](https://docs.github.com/en/actions)

### 2023-06-08 manuscript/5pillars.Rmd
spelling, editing, BiB formatting

@@
 achieve.
-Building upon past efforts and proposals to maximise reproducibility and rigour in
+Building upon past efforts and proposals to maximize reproducibility and rigor in
 bioinformatics, we present a framework called the five pillars of reproducible computational
@@
 *five pillars of reproducible computational research* (**Figure 1**).
-There are clear parallels with the established the three pillars of open science framework,
+There are clear parallels with the established three pillars of the open science framework,
 (open data, code and papers) [@Gorgolewski2016-hv].
@@
 Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
-somewhat overused and misused in the life sciences, leading to problems like inadvertant conversion
+somewhat overused and misused in the life sciences, leading to problems like inadvertent conversion
 of gene names to dates in databases and supplementary Excel files [@Zeeberg2004-js;
@@
 After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
-which saves labour and compute time.
+which saves labor and compute time.
 
@@
 After compilation, the resulting output document contains the code together with computational results
-such as figures and tables along with contextualising explanations and narratives.
+such as figures and tables along with contextualizing explanations and narratives.
 
@@
 * Helps collaboration and project management.
-Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
+Using a centralized code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
 from team members who may be working asynchronously in different time zones and on different servers.
@@
 * Helps resolve issues.
-These centralised platforms also enable discussions of code issues such as bugs and feature requests.
+These centralized platforms also enable discussions of code issues such as bugs and feature requests.
 In bioinformatics teams, issues can be used to track the progress of a project through its milestones
@@
 [Software Heritage](https://www.softwareheritage.org/) @dicosmo2017 is the universal software archive 
-which mission is to collect, preserve and share source code. Software Heritage provides 
+whose mission is to collect, preserve and share source code. Software Heritage provides 
 an intrinsic persistent identifier called SWHID that allows to cite the archived source code 
@@
 [@Wurmus2018-lv; @Courtes2022-rx].
-Guix is probably one of the best solution so far for the "ten-year challenge" of code reproducibility @Perkel2020-io.
+Guix is among the best solutions for the "ten-year challenge" of code reproducibility @Perkel2020-io.
 Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
@@
 to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).
-Eventually, it's also worth mentionning that the Guix project uses Software Heritage
+Lastly, it's worth mentioning that the Guix project uses Software Heritage
 as a fallback system to provide a package source code
-if the download of the source code from its original location failed.
+if the download from its original location failed.
 
@@
 [@Ioannidis2009-hv, @Archmiller2020-er ].
-Data sharing is also one of the key features of "open science", which is characterised as science
+Data sharing is also one of the key features of "open science", which is characterized as science
 that is collaborative, transparent, accessible and inclusive [@McKiernan2016-vu; @noauthor_2023-uq].
@@
 research ideas that were previously impossible [@Wilkinson2016-qq].
-In terms of research rigour, inspecting raw data can uncover inadvertent errors and research
+In terms of research rigor, inspecting raw data can uncover inadvertent errors and research
 integrity problems [@Miyakawa2020-yj].
@@
 In the rare cases that data is shared in accordance with the data availability statement, it poses
-a burden in terms of labour to lodge and respond to such requests, and if the data is large, may incur
+a burden in terms of labor to lodge and respond to such requests, and if the data is large, may incur
 an additional cost for hosting/transferring these large files.
@@
 This is a well-founded concern for identifiable data, however once identifying features are removed,
-it is nearly impossible to reidentify genomic data @Venkatesaramani2021-tn unless the attackers have
-preexisting genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
-Therefore, sharing deidentified genomic data and metadata is likely to carry little risk to participant
+it is nearly impossible to re-identify genomic data @Venkatesaramani2021-tn unless the attackers have
+pre-existing genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
+Therefore, sharing de-identified genomic data and metadata is likely to carry little risk to participant
 privacy.
@@
 
-In order to maximise the value of shared data for reuse and reproducibility, it needs to be Findable,
+In order to maximize the value of shared data for reuse and reproducibility, it needs to be Findable,
 Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
-These principles are summarised in **Figure 5**.
+These principles are summarized in **Figure 5**.
 
@@
 conforming to FAIR principles (e.g.: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
-However other domains lack well known specialised organised data repositories, such as evolutionary
+However other domains lack well known specialized organized data repositories, such as evolutionary
 biology, ecology, cell biology, social sciences and computing.
@@
 repository that accepts data from these fields [@Pampel2013-ep].
-If no specialised repository exists, then a general-purpose repository such as
+If no specialized repository exists, then a general-purpose repository such as
 [Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
@@
 
-1. Depositing data to a specialised repository if possible, otherwise, a general-purpose repository.
+1. Depositing data to a specialized repository if possible, otherwise, a general-purpose repository.
 
@@
 Once the quality of the data set has been established researchers may consider depositing it
-to a data repository early, before publicisation, as it has some advantages.
+to a data repository early, before publicization, as it has some advantages.
 Most data repositories have an optional 12-month embargo period so researchers can share publicly
@@
 In order to guarantee future reproducibility, snapshots of those reference data should be archived
-and shared if the licence permits it.
+and shared if the license permits it.
 
@@
 Documentation is the glue that binds a data science project together.
-The manuscript/published article is the central artefact that outlines the research project and
+The manuscript/published article is the central artifact that outlines the research project and
 links to the supporting materials.
@@
 with just six commands, starting from a new Ubuntu Linux install [@Ziemann2023-nm].
-By minimising the complexity of reproduction and writing the instructions clearly in the `README`,
+By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
 reproducibility is made possible to a larger segment of the computational research community,
@@
 More such protocols are needed to make extreme reproducibility mainstream.
-A curated set of tutorials is provided in the Supplement.
+A compendium of publicly available learning materials around the five pillar concepts are
+provided in the supplementary information.
 
@@
 are falling short of this goal.
-Our recommendations for workflow-based research articles are summarised as:
+Our recommendations for workflow-based research articles are summarized as:
 
@@
 5. Share research data and metadata according to FAIR principles.
+
+## Key points
+
+* Irreproducibility of bioinformatics studies remains a significant problem.
+
+* We present the *five pillars* framework, a set of best practices that enable extremely reproducible
+workflows.
+
+* Widespread adoption of these principles will enhance research reliability and speed translation of
+basic research to tangible benefits.
 
@@
 
+## Supplementary information
+
+[Supplementary materials 1: A curated compendium of publicly available educational materials related
+to five pillar concepts.](https://ziemann-lab.net/public/5pillars/practical_guides.html)
+
+## Author biographies
+
+**Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
+His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
+of research rigour.
+
+**Samuel S. Minot** works as a Staff Scientist at Fred Hutchinson Cancer Center within the Vaccine and
+Infectious Disease Division.
+His research is focused on human microbiota associations with human health and disease.
+
+**Pierre Poulain** is an Associate Professor at Université Paris Cité, Institut Jacques Monod, France.
+Research directions include proteomics, bioinformatics and education/pedagogy.
+
+**Anusuiya Bora** is a PhD student at Deakin University School of Life and Environmental Sciences.
+Her research project involves improving the rigor of commonly used bioinformatics tools.
+
 ## Bibliography

### 2023-06-08 manuscript/5pillars.Rmd
close #17 references checked - using plos csl as the bib csl is bad

@@
 bibliography: 5pillars.bib
-csl: briefings-in-bioinformatics.csl
+csl: plos-computational-biology.csl
 ---
@@
 Much has been written on how irreproducibility has reached crisis levels in psychology and life
-sciences [@Baker2015-bs; @Begley2012-di].
+sciences [@Baker2015; @Begley2012-di].
 Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
@@
 study.
-This automated process removes the need for any manual steps, which are not only time-consuming but
-also prone to errors.
+This automated process removes the need for any manual steps, which are time-consuming and
+prone to errors.
 Without an end-to-end automated process, most of the other best practices described here are not
@@
 Unix Shell.
-Using free and open source software ensures that the research is accessible to the greatest audience,
-as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
+Using free and open source software ensures that the research is accessible to the greatest audience
+@Garijo2013-zl, as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
 Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
@@
 `nextflow` [@Di_Tommaso2017-yo].
-The advantage of such tools is that if the analysis were to terminate due to a hardware issue mid-way,
-for example at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't need to be repeated.
+The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
+for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
+need to be repeated.
 After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
@@
 
-![Figure 2. An example R Markdown script (left) and output document (right) @RStudio2016-fw.](../img/markdownchunk.png)
+![Figure 2. An example R Markdown script (left) and output document (right) @Grolemund2017-le.](../img/markdownchunk.png)
 
@@
 The MyST{NB} (short for Markedly Structured Text notebook) is built on top of Jupyter and has some
-notable advantages around flexibility of content that make it an attractive choice for an executable
-paper authoring [@DuPre2022-bk].
+notable advantages around flexibility of content that make it an attractive choice for executable
+article authoring [@DuPre2022-bk].
 
@@
 code to consumers.
-There are many such DVCSs (subversion, git, bazaar, mercurial...), but `git` has emerged as the most popular
+There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
 solution due to its many powerful features, speed/efficiency and large community and ecosystem.
@@
 to data and metadata, so it doesn't need any modifications before reproduction.
-It is relatively easy to add a software license with a DVCS like `GitHub`, so consumers can understand
+It is relatively easy to add a software license with a DVCS like GitHub, so consumers can understand
 the conditions of using it.
@@
 
-Containers are an attempt to resolve some of the downsides of VMs and package managers.
+Containers are an attempt to resolve some of the downsides of VMs.
 Container images can be thought of as similar to VMs, but are more lightweight as they
@@
 Guix is described as a "functional package manager" and has the unique benefit of giving
-bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-rx].
+bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-zz].
 This is a notable advantage, as Docker images are not verifiable nor guaranteed to build reproducibly
@@
 that work equally well on personal as well as shared high performance computers
-[@Wurmus2018-lv; @Courtes2022-rx].
+[@Wurmus2018-lv; @Courtes2022-zz].
 Guix is among the best solutions for the "ten-year challenge" of code reproducibility @Perkel2020-io.
@@
 (eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]), although there are some challenges.
-
 Some researchers would rather keep data for their private use.
@@
 In the interests of reuse and reproducibility, future studies could consider a consent form that allows
-more broad sharing of deidentified data.
+more broad sharing of de-identified data.
 
@@
 
-6. Provide detailed metadata; e.g.: sample descriptions that match the manuscript; describe the columns
+6. Provide detailed metadata; e.g.: sample descriptions that match the article; describe the columns
 in tabular data.
@@
 Documentation is the glue that binds a data science project together.
-The manuscript/published article is the central artifact that outlines the research project and
+The published article is the central artifact that outlines the research project and
 links to the supporting materials.
@@
 Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
-their reproducibility [@Simoneau2021-kp].
+their reproducibility [@Garijo2013-zl; @Simoneau2021-kp].
+Commonly missing information includes versions of software and packages as well as any parameter setting
+and configuration files.
 The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
@@
 [RIO Journal](https://riojournal.com/) or [Zenodo](https://zenodo.org/), minting a Digital Object
-Identifier (DOI) that can be cited in the manuscript.
+Identifier (DOI) that can be cited in the article.
 This is particularly helpful when there are strict word limits on articles.
 
-The manuscript should have clear links to supporting materials including datasets, software code and
+The article should have clear links to supporting materials including datasets, software code and
 other resources like computational environments.
@@
 It should outline the purpose of the code/software/overall project and how it relates to the
-manuscript/article.
+article.
 For example, some articles may rely on more than one repository for various parts of the research
@@
 Hardware requirements such as RAM, CPU architecture, GPU needs, etc need to be outlined.
-Software requirements, such as operating system, container engine, workflow manager,
+Software requirements, such as operating system, dependencies, container engine, workflow manager,
 etc, also need to be described.
@@
 For example, the pathway analysis workflow we previously developed could be reproduced and inspected
-with just six commands, starting from a new Ubuntu Linux install [@Ziemann2023-nm].
+with just six commands, starting from a new Ubuntu Linux installation [@Ziemann2023-nm].
 By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
@@
 a thorough description of their purpose.
-Both the data and container image descriptions need to contain links to the DOI of the manuscript,
+Both the data and container image descriptions need to contain links to the DOI of the article,
 so that machines or individuals browsing these resources can understand the context of these objects
@@
 yet progress on integrating these into computational research publications has been minimal.
-Part of the problem is the incentive structures in academia, with funding bodies and journals
-largely focused on novel research directions rather than confirming previous findings
-[@Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
 Indeed the number of systematic replication studies in bioinformatics remains small
@@
 well understood.
+Part of the problem is the incentive structures in academia, with funding bodies and journals
+largely focused on novel research directions rather than confirming previous findings
+[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
 
@@
 are falling short of this goal.
-Our recommendations for workflow-based research articles are summarized as:
+Our recommendations for computational researchers are summarized as:
 
@@
 His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
-of research rigour.
+of research rigor.
 
@@
 **Anusuiya Bora** is a PhD student at Deakin University School of Life and Environmental Sciences.
-Her research project involves improving the rigor of commonly used bioinformatics tools.
+Her research project involves improving the reproducibility of commonly used bioinformatics tools.
 

### 2023-06-10 manuscript/5pillars_docx.Rmd
Sam moved to acknowledgement - confirmed via email

--- 2023-06-08 manuscript/5pillars.Rmd
@@
 title: "The five pillars of computational reproducibility: Bioinformatics and beyond"
-author: "Mark Ziemann, Samuel S. Minot, Pierre Poulain, Anusuiya Bora"
+author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
 date: "`r Sys.Date()`"
-output:
-  rmdformats::robobook:
-  html_document:
-    toc: true
-    toc_float: true
+output: word_document
 code_folding: hide
-    fig_width: 5
-    fig_height: 5
+fig_width: 7
+fig_height: 6
 bibliography: 5pillars.bib
@@
 
-Mark Ziemann<sup>1*</sup>, Samuel S. Minot<sup>2</sup>, Pierre Poulain<sup>3</sup>,
-Anusuiya Bora<sup>1</sup>
-
-<details><summary>Author information (*click to expand*)</summary><p>
+Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>, Anusuiya Bora<sup>1</sup>
 
@@
 
-2. Fred Hutchinson Cancer Center, Seattle, USA.
+2. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
 
-3. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
-
-Correspondance: m.ziemann@deakin.edu.au
+* Correspondance: m.ziemann@deakin.edu.au
 
@@
 | Mark Ziemann | 0000-0002-7688-6974 |
-| Samuel S. Minot | 0000-0003-1639-3905 |
 | Pierre Poulain | 0000-0003-4177-3619 |
@@
 
-</details>
-</p>
 Source: https://github.com/markziemann/5pillars/blob/main/manuscript/5pillars.Rmd
 
-## Summary
+<br>
+
+## Abstract
 
@@
 domains.
+
+**Keywords:** Reproducibility; computational research; bioinformatics; research best practices.
+
+<br>
 
@@
 
+<br>
+
 ## End-to-end automated process
@@
 which saves labor and compute time.
+
+<br>
 
@@
 This further makes the job of reproducibility easier.
+
+<br>
 
@@
 
+<br>
+
 ## Compute environment control
@@
 
+<br>
+
 ## FAIR and persistent data sharing
@@
 
+<br>
+
 ## Documentation
@@
 
+<br>
+
 ## Limitations and challenges
@@
 
+<br>
+
 ## Conclusion
@@
 
+<br>
+
 ## Key points
@@
 
+<br>
+
 ## Acknowledgements
@@
 Dr Simon Tournier (Paris Cité University, France) for advice on Guix.
-
+We appreciate advice on Python notebooks from Dr Samuel S. Minot (Fred Hutchinson Cancer Center).
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
 platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+
+<br>
 
@@
 
+<br>
+
 ## Supplementary information
@@
 to five pillar concepts.](https://ziemann-lab.net/public/5pillars/practical_guides.html)
+
+<br>
 
@@
 
-**Samuel S. Minot** works as a Staff Scientist at Fred Hutchinson Cancer Center within the Vaccine and
-Infectious Disease Division.
-His research is focused on human microbiota associations with human health and disease.
-
 **Pierre Poulain** is an Associate Professor at Université Paris Cité, Institut Jacques Monod, France.
@@
 
+<br>
+
 ## Bibliography

### 2023-06-10 manuscript/5pillars.Rmd
keywords added
Add reference citations for Project Jupyter

@@
 title: "The five pillars of computational reproducibility: Bioinformatics and beyond"
-author: "Mark Ziemann, Samuel S. Minot, Pierre Poulain, Anusuiya Bora"
+author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
 date: "`r Sys.Date()`"
@@
 
+<br>
+
 ## Literate programming
@@
 Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
-R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
+R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
 The MyST{NB} (short for Markedly Structured Text notebook) is built on top of Jupyter and has some
@@
 Dr Simon Tournier (Paris Cité University, France) for advice on Guix.
-
+We appreciate advice on Python notebooks from Samuel S. Minot (Fred Hutchinson Cancer Center).
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
 platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+
+<br>
 

### 2023-06-10 manuscript/5pillars.Rmd
Add example of paper and Jupyter notebook for Zika virus analysis

@@
 R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
+A noteworthy attempt to use litterate programming in Jupyter notebooks to aim for reproducibility 
+has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
+In this work, authors provided along the traditionnal research paper a 
+[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb) 
+that performed and documented the entire analysis procedure.
 The MyST{NB} (short for Markedly Structured Text notebook) is built on top of Jupyter and has some

### 2023-06-10 manuscript/5pillars.Rmd
Rephrase Myst-NB

@@
 that performed and documented the entire analysis procedure.
-The MyST{NB} (short for Markedly Structured Text notebook) is built on top of Jupyter and has some
-notable advantages around flexibility of content that make it an attractive choice for executable
+The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
+and has some notable advantages around flexibility of content that make it an attractive choice for executable
 article authoring [@DuPre2022-bk].

### 2023-06-10 manuscript/5pillars.Rmd
Add documentation and missing references for Software Heritage

@@
 to store the content of a repository but do not retain the history of the development of the source code.
-[Software Heritage](https://www.softwareheritage.org/) @dicosmo2017 is the universal software archive
+[Software Heritage](https://www.softwareheritage.org/) [@dicosmo2017; abramatic2018] is the universal software archive
 whose mission is to collect, preserve and share source code. Software Heritage provides
 an intrinsic persistent identifier called SWHID that allows to cite the archived source code
-in the respective journal article/preprint. The archival of the source code in Software Heritage
-can be performed manually or automated by a webhook within a continuous integration.
+in the respective journal article/preprint [@cosmo2020]. The archival of the source code in Software Heritage
+can be performed [manually](https://archive.softwareheritage.org/save/) 
+or automated by a [webhook](https://docs.softwareheritage.org/user/save_code_now/webhooks/index.html) 
+within continuous integration.
 

### 2023-06-10 manuscript/5pillars.Rmd
Introduce the idea of 'cumulative science'

@@
 rectified.
-Therefore, reproducibility is a necessary stepping-stone that should enhance the overall
-reliability of computational research including replicability and robustness.
+Therefore, reproducibility is a necessary stepping-stone in the framework of cumulative science.
+Reproducibility should enhance the overall reliability of computational research 
+including replicability and robustness.
 Replicability being defined as the ability to repeat the entire study and come up with similar overall

### 2023-06-10 manuscript/5pillars.Rmd
Add comment of the use of DVCS

@@
 
+Finally, although DVCSs come from the world of software engineering, 
+their use has become a standard in computational biology, as witnessed by the numerous guides 
+and best practices on the subject [@ram2013; @blischak2016; @perez-riverol2016].
+
 <br>
@@
 promise of reproducibility.
-While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
+While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
 containerisation system for web development and research.
@@
 High throughput biology has already built a number of data repositories for specific data types
-conforming to FAIR principles (e.g.: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
+conforming to FAIR principles (e.g.: GEO, SRA, ENA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
 However other domains lack well known specialized organized data repositories, such as evolutionary

### 2023-06-10 manuscript/5pillars.Rmd
Add warning about non-moderated data repositories

@@
 However other domains lack well known specialized organized data repositories, such as evolutionary
-biology, ecology, cell biology, social sciences and computing.
+biology, ecology, cell biology, molecular simulation, social sciences and computing.
 The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
@@
 [Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
-should be considered.
-Moreover, general recommendations have been proposed to enhance data FAIRness in ecology, but which
+should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
+and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
+To this aime, general recommendations have been proposed to enhance data FAIRness in ecology, but which
 can also be applied to other disciplines [@Roche2015-aq].

### 2023-06-10 manuscript/5pillars.Rmd
Add Nekrutenko paper on reproducibility

@@
 Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
-their reproducibility [@Garijo2013-zl; @Simoneau2021-kp].
+their reproducibility [@Nekrutenko2012; @Garijo2013-zl; @Simoneau2021-kp].
 Commonly missing information includes versions of software and packages as well as any parameter setting

### 2023-06-10 manuscript/5pillars.Rmd
Complete key points and bio

@@
 
-* Irreproducibility of bioinformatics studies remains a significant problem.
+* Irreproducibility of bioinformatics studies remains a significant and still relevant problem.
 
@@
 **Pierre Poulain** is an Associate Professor at Université Paris Cité, Institut Jacques Monod, France.
-Research directions include proteomics, bioinformatics and education/pedagogy.
+Research directions include bioinformatics, large scale data analysis and education/pedagogy.
 

### 2023-06-11 guides/practical_guides.Rmd
Update practical_guides.Rmd

@@
 
+* [Video Tutorial: Git and GitHub for Beginners - Crash Course](https://www.youtube.com/watch?v=RGOj5yH7evk)
+
+* [Video Tutorial: Git basics - 4 short video tutorials](https://git-scm.com/videos)
+
+* [Quick Guide: Git for humans](https://speakerdeck.com/alicebartlett/git-for-humans)
+
 * [Tutorial: How to use Git with R and RStudio](https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/)
@@
 * [Tutorial: How to link and use Git with RStudio](https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/)
+
+* [Ebook:Happy Git and GitHub for the useR](https://happygitwithr.com/index.html)
+
+* [Ebook: Reproducible workflow and version control with Git and Github](https://jules32.github.io/2016-07-12-Oxford/git/)
 
@@
 * [Tutorial: Introduction to Using Git in Jupyter](https://andgarc.github.io/basic-git-jupyter-lesson/index.html)
+
+* [Ebook: Version Control with Git](https://swcarpentry.github.io/git-novice/)
+
+* [Tutorial: Using Git source control in VS Code](https://code.visualstudio.com/docs/sourcecontrol/overview)
 
@@
 
+* [Video Tutorial: Reproducible Data Analysis with Snakemake](https://www.youtube.com/watch?v=hPrXcUUp70Y)
+
+* [Tutorial: Understanding Snakemake](https://vincebuffalo.com/blog/2020/03/04/understanding-snakemake.html)
+
 * [Tutorial: An example Snakemake workflow](https://snakemake.readthedocs.io/en/stable/tutorial/basics.html)
+
+* [Quick Guide: Using snakemake to do simple wildcard operations on many, many, many files](http://ivory.idyll.org/blog/2021-snakemake-simple-operations.html)
 

### 2023-06-12 manuscript/5pillars.Rmd
fix typos

@@
 R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
-A noteworthy attempt to use litterate programming in Jupyter notebooks to aim for reproducibility 
+A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility 
 has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
@@
 practices.
-Below I outline some of the benefits.
+Below we outline some of the benefits.
 
@@
 and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
-To this aime, general recommendations have been proposed to enhance data FAIRness in ecology, but which
+To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
 can also be applied to other disciplines [@Roche2015-aq].

### 2023-06-12 manuscript/5pillars.Rmd
funding
fixed quirk with refs 10-20

@@
 [ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
-@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw;
-@DuPre2022-bk].
+@Gruning2018-sg; @Akalin2018-eg; @Brito2020-ld; @Peikert2021-eq;
+@Rodrigues2023-fw; @DuPre2022-bk].
 Inspired by these principles, this article presents the
@@
 
+## Funding
+
+The authors received no specific funding for this work.
+
+<br>
+
 ## Availability of materials

### 2023-06-12 guides/practical_guides.Rmd
guides; drake -> targets

@@
 title: "Practical guides for implementing 'Five Pillar' principles"
-author: "Mark Ziemann, Samuel S. Minot, Pierre Poulain, Anusuiya Bora"
+author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
 date: "`r Sys.Date()`"
@@
 
-Mark Ziemann<sup>1*</sup>, Samuel S. Minot<sup>2</sup>, Pierre Poulain<sup>3</sup>,
+Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>,
 Anusuiya Bora<sup>1</sup>
 
-<details><summary>Author information (*click to expand*)</summary><p>
+Author information
 
@@
 
-2. Fred Hutchinson Cancer Center, Seattle, USA.
-
-3. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
+2. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
 
@@
 
-Source: https://github.com/markziemann/5pillars/blob/main/manuscript/practical_guides.Rmd
+Source: https://github.com/markziemann/5pillars/blob/main/guides/practical_guides.Rmd
 
@@
 
+<br>
+
 ## Getting started with data analysis in R and Python
@@
 
+<br>
+
 ### Introduction to R and Rstudio
+
+* [Video: Rstudio for the Total Beginner](https://www.youtube.com/watch?v=FIrsOBy5k58)
+
+* [Video: Running Basic Statistical Analysis in R](https://www.youtube.com/watch?v=I6FJo8x1wZE)
 
@@
 
-* [Video: Rstudio for the Total Beginner](https://www.youtube.com/watch?v=FIrsOBy5k58)
+* [Ebook: YaRrr! The Pirate’s Guide to R](https://bookdown.org/ndphillips/YaRrr/)
 
-* [Ebook: YaRrr! The Pirate’s Guide to R](https://bookdown.org/ndphillips/YaRrr/)
+<br>
 
@@
 * [Book: Unix Shell Programming (ISBN:0672324903)](https://www.amazon.com/Unix-Shell-Programming-Stephen-Kochan/dp/0672324903)
+
+<br>
 
@@
 
+<br>
+
 ### Introduction to VS Code
@@
 * [Tutorial and video: Getting Started with Visual Studio Code](https://code.visualstudio.com/docs/introvideos/basics)
+
+<br>
 
@@
 
+<br>
+
 ## Practical guides for `git` and GitHub
@@
 
-* [Ebook:Happy Git and GitHub for the useR](https://happygitwithr.com/index.html)
+* [Ebook: Happy Git and GitHub for the useR](https://happygitwithr.com/index.html)
 
@@
 
-* [Tutorial: Introduction to Using Git in Jupyter](https://andgarc.github.io/basic-git-jupyter-lesson/index.html)
-
 * [Ebook: Version Control with Git](https://swcarpentry.github.io/git-novice/)
@@
 * [Tutorial: Using Git source control in VS Code](https://code.visualstudio.com/docs/sourcecontrol/overview)
+
+<br>
 
@@
 
-* [Guide: Introduction to data sharing](https://zenodo.org/badge/DOI/10.5281/zenodo.3518692.svg)
+* [Guide: Introduction to data sharing](https://doi.org/10.5281/zenodo.3518692)
 
 * [Guide: FAIR Data (according to Australian Research Data Commons)](https://ardc.edu.au/resource/fair-data/)
+
+* [Documentation: FAIRsharing.org](https://fairsharing.org/)
+
+<br>
 
@@
 
-* [Tutorial: A short introduction to drake (An R based workflow manager)](https://data.nozav.org/posts/2019-a-short-introduction-to-drake/)
+* [Documentation: The {targets} R package user manual (workflow manager for R)](https://books.ropensci.org/targets/)
 
@@
 * [Tutorial compendium: Learn Nextflow in 2023](https://snakemake.readthedocs.io/en/stable/tutorial/basics.html)
+
+<br>
 
@@
 
+<br>
+
 ## Practical guides for documenting computational research
@@
 
+* [Ebook: Bioinformatics Data Skills Chapter 2](https://vincebuffalo.com/book/)
+
 * [Example: README for "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker"](https://github.com/aaronpeikert/reproducible-research/blob/master/README.md)
+
+<br>
 

### 2023-06-12 manuscript/5pillars.Rmd
guides; drake -> targets

@@
 Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
-`snakemake` [@Koster2012-sq; @Molder2021-mp], `drake` [@Michael_Landau2018-kk] or
+`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2012] or
 `nextflow` [@Di_Tommaso2017-yo].

### 2023-06-12 manuscript/5pillars.Rmd
correct landau reference

@@
 Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
-`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2012] or
+`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
 `nextflow` [@Di_Tommaso2017-yo].

### 2023-06-12 manuscript/5pillars.Rmd
documentation edits
fix citation

@@
 other resources like computational environments.
+
 The code repository should have a detailed `README` file which plays a critical role in reproducibility.
@@
 Indeed the number of systematic replication studies in bioinformatics remains small
-[@Zaringhalam2020-ms ; @Ioannidis2009-hv; @Samuel2022-ha], so the severity of the problem isn't
+[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
 well understood.

### 2023-06-12 manuscript/5pillars.Rmd
removed the docker image because it belongs to docker inc

@@
 share parts of the host operating system [@Nust2020-cr].
-In the example shown in **Figure 4**, the six containerised applications share the same operating
+In the example shown in **Figure 4**, the five containerised applications share the same operating
 system, while the three VM applications each involve their own operating system which incurs a
@@
 
-![Figure 4. Architecture of containers (left) compared to VMs (right) @noauthor_2021-lt.](../img/docker-containerized-and-vm-transparent-bg.webp)
+```{r,containers,fig.cap="Figure 4. Architecture of containers (left) compared to VMs (right). Adapted from @noauthor_2021-lt.",echo=FALSE,fig.height=6}
 
+library("RColorBrewer")
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 12), ylim=c(0, 8))
+
+#border
+polygon(c(0,12,12,0),c(0,0,8,8),border="gray")
+
+polygon(c(0,5.5,5.5,0),c(0,0,0.9,0.9), col=cols2[1], border = NA)
+polygon(c(0,5.5,5.5,0),c(1,1,1.9,1.9), col=cols2[2], border = NA)
+polygon(c(0,5.5,5.5,0),c(2,2,2.9,2.9), col=cols2[3], border = NA)
+
+polygon(c(0,1,1,0),c(3,3,5,5), col=cols2[4], border = NA)
+polygon(c(1.1,2.1,2.1,1.1),c(3,3,5,5), col=cols2[4], border = NA)
+polygon(c(2.2,3.2,3.2,2.2),c(3,3,5,5), col=cols2[4], border = NA)
+polygon(c(3.3,4.3,4.3,3.3),c(3,3,5,5), col=cols2[4], border = NA)
+polygon(c(4.4,5.4,5.4,4.4),c(3,3,5,5), col=cols2[4], border = NA)
+
+labs1 <- c("Hardware","Host Operating System","Container Engine")
+text(c(2.7,2.7,2.7),c(0.4,1.4,2.4),labels=labs1)
+
+labs2 <- c("App 1","App 2","App 3","App 4","App 5")
+text(c(0.5,1.6,2.7,3.8,4.9) ,c(4,4,4,4,4), labels=labs2,srt=90)
+
+lines(c(0,0,5.5,5.5),c(5.2,5.5,5.5,5.2)  )
+text(2.7,5.7,labels="Containerized applications")
+
+polygon(c(6,11.5,11.5,6),c(0,0,0.9,0.9), col=cols2[1], border = NA)
+polygon(c(6,11.5,11.5,6),c(1,1,1.9,1.9), col=cols2[3], border = NA)
+
+labs3 <- c("Hardware","Hypervisor")
+text(c(8.5,8.5),c(0.4,1.4),labels=labs3)
+
+polygon(c(6,7.8,7.8,6),c(2,2,4.5,4.5), col=cols2[2], border = NA)
+polygon(c(8,9.6,9.6,8),c(2,2,4.5,4.5), col=cols2[2], border = NA)
+polygon(c(9.8,11.5,11.5,9.8),c(2,2,4.5,4.5), col=cols2[2], border = NA)
+
+text(c(6.85,8.8,10.65),c(3.5,3.5,3.5),labels="Guest\nOperating\nSystem")
+
+polygon(c(6,7.8,7.8,6),c(4.5,4.5,6,6), col=cols2[4], border = NA)
+polygon(c(8,9.6,9.6,8),c(4.5,4.5,6,6), col=cols2[4], border = NA)
+polygon(c(9.8,11.5,11.5,9.8),c(4.5,4.5,6,6), col=cols2[4], border = NA)
+
+text(c(6.85,8.8,10.65),c(5,5,5),labels=labs2[1:3])
+
+lines(c(6,6,11.5,11.5),c(6.2,6.5,6.5,6.2)  )
+text(8.5,6.7,labels="Virtualized applications")
+
+```
 Another potential solution to this problem is to use a package/environment management system such as

### 2023-06-13 manuscript/5pillars.Rmd
#18 fixing some details

@@
 
-Computational reproducibility is a simple premise in theory, but in practice it is difficult to
-achieve.
+Computational reproducibility is a simple premise in theory, but is difficult to
+achieve in practice.
 Building upon past efforts and proposals to maximize reproducibility and rigor in
@@
 has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
-In this work, authors provided along the traditionnal research paper a 
+In this work, authors provided along the traditional research paper a
 [Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
@@
 to store the content of a repository but do not retain the history of the development of the source code.
-[Software Heritage](https://www.softwareheritage.org/) [@dicosmo2017; abramatic2018] is the universal software archive
-whose mission is to collect, preserve and share source code. Software Heritage provides
-an intrinsic persistent identifier called SWHID that allows to cite the archived source code
-in the respective journal article/preprint [@cosmo2020]. The archival of the source code in Software Heritage
-can be performed [manually](https://archive.softwareheritage.org/save/)
-or automated by a [webhook](https://docs.softwareheritage.org/user/save_code_now/webhooks/index.html)
-within continuous integration.
+[Software Heritage](https://www.softwareheritage.org/) [@dicosmo2017; @abramatic2018] is the universal
+software archive whose mission is to collect, preserve and share source code.
+Software Heritage provides an intrinsic persistent identifier called SWHID that allows to cite the
+archived source code in the respective journal article/preprint [@cosmo2020].
+Archiving source code in Software Heritage can be performed
+[manually](https://archive.softwareheritage.org/save/) or automatically using a
+[webhook](https://docs.softwareheritage.org/user/save_code_now/webhooks/index.html) within continuous
+integration.
 
@@
 plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 12), ylim=c(0, 8))
-
-#border
-polygon(c(0,12,12,0),c(0,0,8,8),border="gray")
 
@@
 A lack of technical computer skills in many research groups is another culprit.
-Specific workshop-based training in these areas could help, but the most scalable solution would be to
-provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
+Specific workshop-based training in these areas could help [@Williams2023-ic].
+The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
+[@Wilson2014-mh].
+However, the most scalable solution would be to provide comprehensive beginner-friendly step-by-step
+guides for frequently used routines.
 For example, our group has just released such a protocol that walks novice users through the process

### 2023-06-15 manuscript/5pillars.Rmd
comments from martin

@@
 labs <- c("Developer 1",
-  "Local Repo",
+  "Local Repository",
   "Working Files",
   "Developer 2",
-  "Local Repo",
+  "Local Repository",
   "Working Files",
@@
 * Supports documentation best practices.
-Books, educational resources and other data types are well served using DVCS (see below).
+Books, tutorials, protocols and other educational resources are well served using DVCS (see below).
 
@@
 
-* Protect against code loss.
+* Protects against code loss.
 Computers sometimes fail and we sometimes delete or overwrite important files.
@@
 containerisation system for web development and research.
-Due to its early adoption, Docker now has a large community of users, extensive documentation and
+Docker now has a large community of users, extensive documentation and
 a vast collection of pre-built container images in the DockerHub registry.
@@
 6. Provide detailed metadata; e.g.: sample descriptions that match the article; describe the columns
-in tabular data.
+in tabular data (i.e.: data dictionary).
 
@@
 
-* Widespread adoption of these principles will enhance research reliability and speed translation of
+* Widespread adoption of these principles will enhance research reliability and will speed translation of
 basic research to tangible benefits.
@@
 We appreciate advice on Python notebooks from Samuel S. Minot (Fred Hutchinson Cancer Center).
+We are grateful to Dr Martin O'Hely (Deakin University) for critical comments on the manuscript.
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research

### 2023-06-15 manuscript/5pillars.Rmd
#22 set seed issue addressed

@@
 Then the computational workflow can begin with the cleaned dataset.
+
+Another issue is that perfect reproducibility isn't possible in cases where some element of randomness
+is requrired by the underlying algorithms.
+Such algorithms are extensively used in molecular simulation, machine learning, permutation-based
+statistical tests and certain data projection methods, namely t-distributed stochastic neighbour
+embedding [@Van_der_Maaten2008-zn] and uniform manifold approximation and projection [@McInnes2018-of],
+which are popular for visualising high-dimensional omics data [@Becht2018-ca].
+To make such workflows deterministic, the pseudo-random number generator can be initialized with a fixed
+value (sometimes called "setting the seed")[@Sandve2013-lk].
+However this needs to be done with care to ensure the results do not misrepresent the bulk of
+iterations (see [@Benureau2017-uo]).
 

### 2023-06-15 manuscript/5pillars.Rmd
fixed dvcs figure

@@
 labs <- c("Developer 1",
+  "Working Files",
   "Local Repository",
+  "Developer 2",
   "Working Files",
-  "Developer 2",
   "Local Repository",
+  "Developer 3",
   "Working Files",
-  "Developer 3",
-  "Local Repository",
-  "Working Files")
+  "Local Repository")
 

### 2023-06-16 manuscript/5pillars.Rmd
added t-SNE and UMAP

@@
 statistical tests and certain data projection methods, namely t-distributed stochastic neighbour
-embedding [@Van_der_Maaten2008-zn] and uniform manifold approximation and projection [@McInnes2018-of],
+embedding (t-SNE) [@Van_der_Maaten2008-zn] and uniform manifold approximation and projection UMAP [@McInnes2018-of],
 which are popular for visualising high-dimensional omics data [@Becht2018-ca].

### 2023-06-16 manuscript/5pillars.Rmd
close #22

--- 2023-06-15 manuscript/5pillars.Rmd
@@
 Another issue is that perfect reproducibility isn't possible in cases where some element of randomness
-is requrired by the underlying algorithms.
+is required by the underlying algorithms.
 Such algorithms are extensively used in molecular simulation, machine learning, permutation-based
-statistical tests and certain data projection methods, namely t-distributed stochastic neighbour
-embedding [@Van_der_Maaten2008-zn] and uniform manifold approximation and projection [@McInnes2018-of],
-which are popular for visualising high-dimensional omics data [@Becht2018-ca].
+statistical tests and certain data projection methods, namely t-distributed stochastic neighbor
+embedding (t-SNE) [@Van_der_Maaten2008-zn] and uniform manifold approximation and projection UMAP [@McInnes2018-of],
+which are popular for visualizing high-dimensional omics data [@Becht2018-ca].
 To make such workflows deterministic, the pseudo-random number generator can be initialized with a fixed

### 2023-06-16 main.sh
fix libxt issue

@@ -4,6 +4,8 @@ git clone https://github.com/markziemann/5pillars.git
 
 cd 5pillars
 
+apt update && apt install libxt-dev -y
+
 Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer"))'
 
 Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'

### 2023-06-16 main.sh
remove apt cmd

@@ -4,8 +4,6 @@ git clone https://github.com/markziemann/5pillars.git
 
 cd 5pillars
 
-apt update && apt install libxt-dev -y
-
 Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer"))'
 
 Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'

### 2023-08-10 manuscript/rebuttal.Rmd
#25 close. Have fixed figure 1 so the order of pillars is same as the text

Dear Editor,

Many thanks for your consideration of our manuscript.
Below is our point-by-point response to reviewer comments.

### Reviewer 1

> In this manuscript, the authors present a framework that computational researchers should follow to maximize the reproducibility of their computational research. The five pillars that they identified are 1) code version control and sharing, 2) compute environment control,  3) literate programming, 4) documentation and 5) FAIR data sharing. They give a fair amount of detail regarding each pillar. This paper contains no original research and is more of an opinion piece. With that in mind, if BiB is interested in this topic, I recommend publishing. The authors make very reasonable suggestions. My only request for edits is to rearrange the section structure of the manuscript to be in line with the numbering of the pillars. Right now it is quite confusing.

Response: We thank the reviewer for their positive feedback.
With regards to the ordering of the five pillars, we have altered Figure 1 so that the pillars now
appear in the same order as they are detailed in sections following.

### 2023-08-10 manuscript/5pillars.Rmd
#25 close. Have fixed figure 1 so the order of pillars is same as the text

@@
 title: "The five pillars of computational reproducibility: Bioinformatics and beyond"
-author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
 date: "`r Sys.Date()`"
@@
 
-Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>, Anusuiya Bora<sup>1</sup>
+<strong>Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>, Anusuiya Bora<sup>1</sup></strong>
 
@@
 
-labs <- c("Code version\ncontrol &\npersistent\n sharing",
+labs <- c("Literate\nprogramming",
+  "Code version\ncontrol &\npersistent\n sharing",
   "Compute \nenvironment \ncontrol",
-  "Literate\nprogramming",
-  "Documentation",
-  "FAIR &\npersistent\ndata sharing")
+  "FAIR &\npersistent\ndata sharing",
+  "Documentation")
 

### 2023-08-11 manuscript/rebuttal.Rmd
working on #26

@@
+---
+title: "Response to reviewer comments (The five pillars)"
+author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
+date: "`r Sys.Date()`"
+output: word_document
+code_folding: hide
+fig_width: 7
+fig_height: 6
+bibliography: 5pillars.bib
+csl: plos-computational-biology.csl
+---
+
 Dear Editor,
@@
 appear in the same order as they are detailed in sections following.
+
+### Reviewer 2
+
+> Comments to the Author: The article offers a comprehensive perspective on reproducibility in computational research, delivering clear and logically backed recommendations.
+It is indeed a very interesting topic that needs our attention.
+Nevertheless, there are areas that could benefit from further development and clarity:
+
+Response: We appreciate reviewer 2's positive feedback and thoughtful suggestions.
+
+> 1.
+Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case studies or practical examples that demonstrate the effectiveness of your recommendations.
+Highlighting specific instances where these best practices have enhanced reproducibility or solved particular issues could substantiate your points more effectively.
+
+Consider case study of enrichment analysis.
+
+> 2.
+Expanding on Current Shortcomings: The article mentions that current practices are not meeting the goal of reproducibility.
+An expanded discussion on the ramifications of these shortcomings on scientific progress could accentuate the urgency and significance of addressing this issue.
+
+After the paragraph ending in "... yet as we see, in practice this is rarely achieved." it would be
+good to discuss the ramifications of such shortcomings on scientific progress.
+This will highlight the urgency and significance of the proposed measures.
+
+> 3.
+Focus on Implementation Challenges: While the article acknowledges the importance of certain practices, an increased focus on potential obstacles researchers may face when trying to implement these suggestions, as well as strategies to overcome these challenges, would be insightful.
+
+Just before the paragraph "A lack of technical computer skills in many research groups is another culprit."
+We should explicitly discuss implementation challenges.
+Part of the preceeding paragraph can also be amended.
+
+> 4.
+Clear Structure: Consider organizing the article into clearly defined sections such as 'Introduction', 'Background', 'Recommendations', 'Challenges', and 'Conclusion'.
+This could significantly enhance readability and comprehensibility.
+
+The subheading structure has been amended in line with these recommendations.
+
+* Abstract
+
+* Introduction
+
+* Recommendations
+
+  * End-to-end automated process
+
+  * Literate programming
+
+  * Code version control and persistent sharing
+
+  * Compute environment control
+
+  * FAIR and persistent data sharing
+
+  * Documentation
+
+* Challenges
+
+* Conclusion
+
+* Key points
+
+> 5.
+Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial.
+Instead of merely summarizing the recommendations, it would be useful to indicate the potential trajectory of computational research if these recommendations are adopted, as well as highlighting the potential risks if they aren't.
+
+Response: We have completely rewritten the conclusion.
+Instead of reiterating the five pillars, we focus on the motivation for adoption:
+
+>Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
+>of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
+>These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
+>significant expertise to analyze and distill key observations.
+>The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
+>Such reuse reduces waste and enhances the cost-effectiveness of research.
+>The transparency provided by sharing data, code and documentation makes it easier for problems to be
+>identified, so that erroneous findings are less likely to pollute the scientific literature.
+>Researchers, their institutions, publishers and funders each have a major role to play in fighting the
+>reproducibility crisis by encouraging highly reproducible research practices.

### 2023-08-11 manuscript/5pillars.Rmd
working on #26

@@
 
-<strong>Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>, Anusuiya Bora<sup>1</sup></strong>
+<strong>Mark Ziemann<sup>1,2*</sup>, Pierre Poulain<sup>3</sup>, Anusuiya Bora<sup>1</sup></strong>
 
@@
 
-2. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
+2. Burnet Institute, Melbourne, Australia
+
+3. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
 
@@
 
-Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
-research directions.
 Much has been written on how irreproducibility has reached crisis levels in psychology and life
@@
 in practice this is rarely achieved.
+
+(TODO: discuss the ramifications of such shortcomings on scientific progress.
+This will highlight the urgency and significance of the proposed measures.)
+
+(Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
+research directions.)
+
+## Recommendations
 
@@
 
-## End-to-end automated process
+### End-to-end automated process
 
@@
 
-## Literate programming
+### Literate programming
 
@@
 
-## Code version control and persistent sharing
+### Code version control and persistent sharing
 
@@
 
-## Compute environment control
+### Compute environment control
 
@@
 
-## FAIR and persistent data sharing
+### FAIR and persistent data sharing
 
@@
 
-## Documentation
+### Documentation
 
@@
 
-## Limitations and challenges
+## Challenges
 
@@
 
+(TODO: reviewer 2 point 3: focus on implementation challenges and strategies to overcome.
+Expand this section articulating the difficulties with making gold standard reproducibility the norm.
+Detail some strategies that can be used.
+Case in point: spreadsheet causing gene name errors
+
+1. described in 2004
+
+2. followed by large scale screens in 2016,2022
+
+3. HUGO finally takes action
+
+4. Public awareness
+
+5. Numbers of affected articles diminishes
+)
+
 A lack of technical computer skills in many research groups is another culprit.
@@
 
-Reproducibility is an essential feature of reliable computational research yet current practices
-are falling short of this goal.
-Our recommendations for computational researchers are summarized as:
-
-0. Encapsulate the entire workflow in an end-to-end automated process, executable with a master script.
-
-1. Share the code repository and deposit a snapshot to an archive.
-
-2. Share a controlled environment; options include a Docker or virtual machine image or
-configuration for Conda or Guix.
-
-3. Use literate scripting tools like R Markdown or Jupyter notebooks to generate all outputs,
-such as charts and tables.
-
-4. Document the workflow sufficiently for reproduction, especially the repository `README`.
-
-5. Share research data and metadata according to FAIR principles.
+Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
+of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
+These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
+significant expertise to analyze and distill key observations.
+The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
+Such reuse reduces waste and enhances the cost-effectiveness of research.
+The transparency provided by sharing data, code and documentation makes it easier for problems to be
+identified, so that erroneous findings are less likely to pollute the scientific literature.
+Researchers, their institutions, publishers and funders each have a major role to play in fighting the
+reproducibility crisis by encouraging highly reproducible research practices.
 
@@
 **Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
+He has recently been recruited to Burnet Institute as Head of Bioinformatics.
 His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects

### 2023-08-11 manuscript/rebuttal.Rmd
potti case study

@@
 
-> In this manuscript, the authors present a framework that computational researchers should follow to maximize the reproducibility of their computational research.
-The five pillars that they identified are 1) code version control and sharing, 2) compute environment control,  3) literate programming, 4) documentation and 5) FAIR data sharing.
-They give a fair amount of detail regarding each pillar.
-This paper contains no original research and is more of an opinion piece.
-With that in mind, if BiB is interested in this topic, I recommend publishing.
-The authors make very reasonable suggestions.
-My only request for edits is to rearrange the section structure of the manuscript to be in line with the numbering of the pillars.
-Right now it is quite confusing.
+> In this manuscript, the authors present a framework that computational researchers should follow to
+> maximize the reproducibility of their computational research. The five pillars that they identified
+> are 1) code version control and sharing, 2) compute environment control,  3) literate programming,
+> 4) documentation and 5) FAIR data sharing. They give a fair amount of detail regarding each pillar.
+> This paper contains no original research and is more of an opinion piece. With that in mind, if BiB
+> is interested in this topic, I recommend publishing. The authors make very reasonable suggestions.
+> My only request for edits is to rearrange the section structure of the manuscript to be in line with
+> the numbering of the pillars. Right now it is quite confusing.
 
 Response: We thank the reviewer for their positive feedback.
-With regards to the ordering of the five pillars, we have altered Figure 1 so that the pillars now
+With regards to the ordering of the five pillars, we have amended Figure 1 so that the pillars now
 appear in the same order as they are detailed in sections following.
@@
 
-> Comments to the Author: The article offers a comprehensive perspective on reproducibility in computational research, delivering clear and logically backed recommendations.
-It is indeed a very interesting topic that needs our attention.
-Nevertheless, there are areas that could benefit from further development and clarity:
+> Comments to the Author: The article offers a comprehensive perspective on reproducibility in
+> computational research, delivering clear and logically backed recommendations. It is indeed a very
+> interesting topic that needs our attention. Nevertheless, there are areas that could benefit from
+> further development and clarity:
 
@@
 
-> 1.
-Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case studies or practical examples that demonstrate the effectiveness of your recommendations.
-Highlighting specific instances where these best practices have enhanced reproducibility or solved particular issues could substantiate your points more effectively.
+> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
+> studies or practical examples that demonstrate the effectiveness of your recommendations.
+> Highlighting specific instances where these best practices have enhanced reproducibility or solved
+> particular issues could substantiate your points more effectively.
 
@@
 
-> 2.
-Expanding on Current Shortcomings: The article mentions that current practices are not meeting the goal of reproducibility.
-An expanded discussion on the ramifications of these shortcomings on scientific progress could accentuate the urgency and significance of addressing this issue.
-
-After the paragraph ending in "... yet as we see, in practice this is rarely achieved." it would be
-good to discuss the ramifications of such shortcomings on scientific progress.
-This will highlight the urgency and significance of the proposed measures.
+> 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the
+> goal of reproducibility. An expanded discussion on the ramifications of these shortcomings on
+> scientific progress could accentuate the urgency and significance of addressing this issue.
 
@@
 
-> 5.
-Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial.
-Instead of merely summarizing the recommendations, it would be useful to indicate the potential trajectory of computational research if these recommendations are adopted, as well as highlighting the potential risks if they aren't.
+> 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of
+> merely summarizing the recommendations, it would be useful to indicate the potential trajectory of
+> computational research if these recommendations are adopted, as well as highlighting the potential
+> risks if they aren't.
 

### 2023-08-11 manuscript/5pillars.Rmd
potti case study

@@
 with 26% of scripts completing without errors [@Trisovic2022-dq].
-Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
-earlier case study @Baggerly2010-nr.
-Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
-in practice this is rarely achieved.
 
-(TODO: discuss the ramifications of such shortcomings on scientific progress.
-This will highlight the urgency and significance of the proposed measures.)
+The ramifications of unreliable research include misleading the community, wasting research funds,
+slowing scientific progress, eroding public confidence in science and tarnishing the reputation
+of associated institutions and colleagues.
+In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.
 
-(Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
-research directions.)
+For example in 2006, an article entitled "Genomic signatures to guide the use of chemotherapeutics"
+generated a great deal of interest as it was an early application of high throughput transcriptomics
+in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
+After identifying some unusual features of the patient group, Baggerly and Coombes attempted
+reproduction of some of the key findings @Baggerly2010-nr.
+Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
+techniques to piece together how the dataset was originally analysed.
+Their investigations found a litany of issues.
+Firstly, the labeling of patients in the test set as "responders" or
+"non-responders" had been reversed in the original.
+Secondly, some of the patients were included more than once (some up to four times) in the analysis,
+which is likely to lead to major distortions in results.
+Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
+both sensitive and resistant.
+Additional errors include two cases of reported results for the wrong drug.
+Baggerly and Coombes highlight that such mistakes can inadvertently occur when using spreadsheets, and
+they note that poor documentation and lack of code contributed to a lack of transparency and poor data
+hygiene.
+The article underwente two corrigenda, but was untimately retracted in 2011, as the authors were not
+able to reproduce the findings themselves due to "corruption of several validation data sets"
+@Potti2011-ut.
+As the array findings were the basis for clinical trials where patients were allocated to treatments,
+the flawed data analysis may have had damaging consequences to patients given the wrong drug in the
+period 2007-2010, before Duke terminated the trials and suspended the lead author, Dr Anil Potti, who
+later resigned.
+Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
+harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
+
+In order for bioinformatic-driven findings to be tranalated to the clinic, they need to be
+
+as demonstrated by an earlier case study @Baggerly2010-nr. <more detail here>
+
+These consequences highlight the urgency for which we must improve stadards for computational research
 
@@
 
+https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-021-05875-3#Sec2
+
 <br>

### 2023-08-13 manuscript/rebuttal.Rmd
progress on rev2

@@
 
-Response: We thank the reviewer for their positive feedback.
+Response: We thank reviewer 1 for their positive feedback.
 With regards to the ordering of the five pillars, we have amended Figure 1 so that the pillars now
@@
 
-Response: We appreciate reviewer 2's positive feedback and thoughtful suggestions.
+Response: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.
 
@@
 
-Consider case study of enrichment analysis.
+To demonstrate the importance of computational reprodcibility, we elaborate on the
+"forensic bioinformatics" case study of Baggerly and Coombes (2010) critique of Potti et al (2006)
+in the introduction.
+
+> For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
+> generated a great deal of interest as it was an early application of high throughput transcriptomics
+> in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
+> After observing some unusual features of the patient group, Baggerly and Coombes attempted
+> reproduction of some of the key findings @Baggerly2010-nr.
+> Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
+> techniques to piece together how the dataset was originally analysed.
+> Their investigations found a litany of issues.
+> Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
+> in the original analysis.
+> Secondly, some of the patients were included more than once (some up to four times) in the analysis,
+> likely to cause major distortions in results.
+> Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
+> both sensitive and resistant.
+> Additional errors include two cases where results (charts) were ascribed to the wrong drug.
+> Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
+> data analysis such as using spreadsheets, and such problems can be obscured by a lack of documentation.
+> The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
+> able to reproduce the findings themselves due to "corruption of several validation data sets"
+> @Potti2011-ut.
+> As the array findings were the basis for clinical trials where patients were allocated to treatments,
+> the flawed data analysis may have harmed patients given the wrong drug in the
+> period 2007-2010.
+> In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
+> later resigned.
+> Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
+> harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
+> This worst-case scenario emphasizes that computational reproducibility is crucial for
+> translating bioinformatics research into real-world outcomes.
 
@@
 
-> 3.
-Focus on Implementation Challenges: While the article acknowledges the importance of certain practices, an increased focus on potential obstacles researchers may face when trying to implement these suggestions, as well as strategies to overcome these challenges, would be insightful.
+In the introduction we have spoken to the consequences of poor computational research, and emphasize
+the urgency of this issue (which complements the case study above).
+
+> The ramifications of irreproducible and unreliable research includes misleading the community,
+> wasting research funds, slowing scientific progress, eroding public confidence in science and
+> tarnishing the reputation of associated institutions and colleagues.
+> In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.
+
+> 3. Focus on Implementation Challenges: While the article acknowledges the importance of certain
+> practices, an increased focus on potential obstacles researchers may face when trying to implement
+> these suggestions, as well as strategies to overcome these challenges, would be insightful.
 

### 2023-08-13 manuscript/5pillars.Rmd
progress on rev2

--- 2023-08-11 manuscript/5pillars.Rmd
@@
 with 26% of scripts completing without errors [@Trisovic2022-dq].
-Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
-earlier case study @Baggerly2010-nr.
-Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
-in practice this is rarely achieved.
 
-(TODO: discuss the ramifications of such shortcomings on scientific progress.
-This will highlight the urgency and significance of the proposed measures.)
+The ramifications of irreproducible and unreliable research includes misleading the community,
+wasting research funds, slowing scientific progress, eroding public confidence in science and
+tarnishing the reputation of associated institutions and colleagues.
+In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.
 
-(Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
-research directions.)
+For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
+generated a great deal of interest as it was an early application of high throughput transcriptomics
+in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
+After observing some unusual features of the patient group, Baggerly and Coombes attempted
+reproduction of some of the key findings @Baggerly2010-nr.
+Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
+techniques to piece together how the dataset was originally analysed.
+Their investigations found a litany of issues.
+Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
+in the original analysis.
+Secondly, some of the patients were included more than once (some up to four times) in the analysis,
+likely to cause major distortions in results.
+Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
+both sensitive and resistant.
+Additional errors include two cases where results (charts) were ascribed to the wrong drug.
+Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
+data analysis such as using spreadsheets, and such problems can be obscured by a lack of documentation.
+The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
+able to reproduce the findings themselves due to "corruption of several validation data sets"
+@Potti2011-ut.
+As the array findings were the basis for clinical trials where patients were allocated to treatments,
+the flawed data analysis may have harmed patients given the wrong drug in the
+period 2007-2010.
+In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
+later resigned.
+Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
+harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
+This worst-case scenario emphasizes that computational reproducibility is crucial for
+translating bioinformatics research into real-world outcomes.
 

### 2023-08-15 manuscript/rebuttal.Rmd
challenges and strategies

@@
 "forensic bioinformatics" case study of Baggerly and Coombes (2010) critique of Potti et al (2006)
-in the introduction.
+in the introduction:
 
@@
 In the introduction we have spoken to the consequences of poor computational research, and emphasize
-the urgency of this issue (which complements the case study above).
+the urgency of this issue (which complements the case study above):
 
-> The ramifications of irreproducible and unreliable research includes misleading the community,
+> The ramifications of irreproducible and unreliable research include misleading the community,
 > wasting research funds, slowing scientific progress, eroding public confidence in science and

### 2023-08-15 manuscript/5pillars.Rmd
challenges and strategies

@@
 
-Due to the constantly developing software landscape, this review is not intended to give strong
-recommendations on the exact combination of tools or resources to use in computational research,
-but to set out key principles and highlight some useful solutions that users can incorporate
-into their toolkit to improve the overall reproducibility of their work.
+Progress on implementing these best practices into computational research publications has been minimal, as
+researchers face a range of obstacles.
+
+These include:
+
+* The "publish-or-perish" incentive structures in academia reward prolific rather than scrupulous
+authors.
+Part of the problem is the incentive structures in academia, with funding bodies and journals
+largely focused on novel research directions rather than confirming previous findings
+[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
+
+* Journal standards are slow to progress.
+To illustrate this, many journals still allow publication without deposition of code or data to public
+repositories.
+
+* A lack of systematic reproducibility/replication studies in bioinformatics remains small
+[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
+well understood.
+
+* A lack of documentated protocols and tutorial materials.
+
+* An overall lack of coding/data science expertise among life science researchers and new graduates.
+
+Overcoming these challenges will require a multi-pronged strategy:
+
+* Incentive structures will need to change.
+How we measure and value research achievement.
+A change in culture or research assessment.
+How we fund research data analytics and infrastructure.
+Value rigour as well as prolificness.
+
+* Journals need to progress towards publishing reproducible workflows.
+Peng 2011 speaks about the lack of infrastructure for disseminating reproducible research.
+
+* Enhanced training.
+A lack of technical computer skills in many research groups is another culprit.
+Specific workshop-based training in these areas could help [@Williams2023-ic].
+The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
+[@Wilson2014-mh].
+Undergraduate and postgraduate courses in the life sciences would benefit from enhanced training in data
+science.
+
+* Enhanced documentation, protocols and tutorials.
+The most scalable solution would be to provide comprehensive beginner-friendly step-by-step
+guides for frequently used routines.
+For example, our group has just released such a protocol that walks novice users through the process
+of implementing functional enrichment analysis of gene expression data with the five pillars principles
+@Ziemann2023-nm.
+More such protocols are needed to make extreme reproducibility mainstream.
+A compendium of publicly available learning materials around the five pillar concepts are
+provided in the supplementary information.
+
+* What individuals can do.
+
+* Overall change in culture.
+Recognition that life sciences is a data intensive discipline.
+
+## Future directions
 
@@
 "temple" of reproducible research.
-
-The key technological advances enabling the five pillars approach have been around for about a decade,
-yet progress on integrating these into computational research publications has been minimal.
-Indeed the number of systematic replication studies in bioinformatics remains small
-[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
-well understood.
-Part of the problem is the incentive structures in academia, with funding bodies and journals
-largely focused on novel research directions rather than confirming previous findings
-[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
-
-(TODO: reviewer 2 point 3: focus on implementation challenges and strategies to overcome.
-Expand this section articulating the difficulties with making gold standard reproducibility the norm.
-Detail some strategies that can be used.
-Case in point: spreadsheet causing gene name errors
-
-1. described in 2004
-
-2. followed by large scale screens in 2016,2022
-
-3. HUGO finally takes action
-
-4. Public awareness
-
-5. Numbers of affected articles diminishes
-)
-
-A lack of technical computer skills in many research groups is another culprit.
-Specific workshop-based training in these areas could help [@Williams2023-ic].
-The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
-[@Wilson2014-mh].
-However, the most scalable solution would be to provide comprehensive beginner-friendly step-by-step
-guides for frequently used routines.
-For example, our group has just released such a protocol that walks novice users through the process
-of implementing functional enrichment analysis of gene expression data with the five pillars principles
-@Ziemann2023-nm.
-More such protocols are needed to make extreme reproducibility mainstream.
-A compendium of publicly available learning materials around the five pillar concepts are
-provided in the supplementary information.
 

### 2023-08-17 manuscript/5pillars.Rmd
progress

@@
 Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
-data analysis such as using spreadsheets, and such problems can be obscured by a lack of documentation.
+data analysis such as using spreadsheets, and these problems can be obscured by a lack of documentation.
 The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
@@
 @Gruning2018-sg; @Akalin2018-eg; @Brito2020-ld; @Peikert2021-eq;
-@Rodrigues2023-fw; @DuPre2022-bk].
+@Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv; @Stodden2016-aj].
 Inspired by these principles, this article presents the
@@
 This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
-reproducibility [@Mangul2019-ei;@Mangul2019-xp].
+reproducibility [@Mangul2019-ei;@Mangul2019-xp;@Krafczyk2021-xv].
 
@@
 
-Progress on implementing these best practices into computational research publications has been minimal, as
-researchers face a range of obstacles.
+Although the technological tools underlying these best practices have existed for a decade,
+progress on implementing them into computational research publications has been minimal.
 
-These include:
+Researchers face several hurdles in adopting these.
 
@@
 authors.
-Part of the problem is the incentive structures in academia, with funding bodies and journals
-largely focused on novel research directions rather than confirming previous findings
-[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].
+Funding bodies, institutions and journals generally prefer to focus on novel research directions rather
+than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
+@Munafo2017-ty; @Diong2021-is].
+
+* A lack of systematic reproducibility/replication studies in bioinformatics remains small
+[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the reproducibility
+problem isn't well appreciated.
 
 * Journal standards are slow to progress.
-To illustrate this, many journals still allow publication without deposition of code or data to public
-repositories.
+Our observations indicate that sharing of data and code remains optional for most medical
+journals in PubMed, but in the field of computational biology, providing these resources is
+becoming standard practice.
 
-* A lack of systematic reproducibility/replication studies in bioinformatics remains small
-[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
-well understood.
-
-* A lack of documentated protocols and tutorial materials.
+* A lack of documentated protocols and tutorial materials for highly-reproducible analysis makes it
+difficult for researchers to confidently update their code development practices.
 
 * An overall lack of coding/data science expertise among life science researchers and new graduates.
+Due to strong demand from industry and government, academia has a deficit of workers fluent in data
+science [@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
+This makes it challenging for organisations to recruit capable data analysts and cultivate a
+community of practice.
 
-Overcoming these challenges will require a multi-pronged strategy:
+Stodden et al have outlined some practical recommendations for tackling these problems [@Stodden2016-aj].
+
+Overcoming these obstacles will require a multi-pronged strategy and effort from different stakeholders.
+
+* What researchers can do.
+
+* What journals can do.
+
+* What insitutions can do.
+
+* What universities can do.
+
+* What funders can do.
 

### 2023-08-17 manuscript/rebuttal.Rmd
reviewer 2 comments addressed I think

@@
 In the introduction we have spoken to the consequences of poor computational research, and emphasize
-the urgency of this issue (which complements the case study above):
+the urgency of this issue (which complements the case study summary above):
 
@@
 
-Just before the paragraph "A lack of technical computer skills in many research groups is another culprit."
-We should explicitly discuss implementation challenges.
-Part of the preceeding paragraph can also be amended.
+We have re-written the Challenges section to highlight some of the barriers slowing the
+adoption of best practices, with reference to previous contributions on the topic.
+These are structured around what different stakeholders can do to improve the situation (including
+funders, journals, institutions and individual researchers).
 
-> 4.
-Clear Structure: Consider organizing the article into clearly defined sections such as 'Introduction', 'Background', 'Recommendations', 'Challenges', and 'Conclusion'.
-This could significantly enhance readability and comprehensibility.
+> 4. Clear Structure: Consider organizing the article into clearly defined sections such as
+> 'Introduction', 'Background', 'Recommendations', 'Challenges', and 'Conclusion'.
+> This could significantly enhance readability and comprehensibility.
 
-The subheading structure has been amended in line with these recommendations.
+The subheading structure has been amended accordingly.
 
@@
 
->Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
->of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
->These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
->significant expertise to analyze and distill key observations.
->The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
->Such reuse reduces waste and enhances the cost-effectiveness of research.
->The transparency provided by sharing data, code and documentation makes it easier for problems to be
->identified, so that erroneous findings are less likely to pollute the scientific literature.
->Researchers, their institutions, publishers and funders each have a major role to play in fighting the
->reproducibility crisis by encouraging highly reproducible research practices.
+> Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
+> of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
+> These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
+> significant expertise to analyze and distill key observations.
+> The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
+> reducing research waste.
+> Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
+> pollute the scientific literature.
+> Researchers, their institutions, publishers and funders each have a major role to play in fighting the
+> reproducibility crisis by encouraging highly reproducible research practices.

### 2023-08-17 manuscript/5pillars.Rmd
reviewer 2 comments addressed I think

@@
 translating bioinformatics research into real-world outcomes.
+
+<br>
 
@@
 This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
-reproducibility [@Mangul2019-ei;@Mangul2019-xp;@Krafczyk2021-xv].
+reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].
 
@@
 progress on implementing them into computational research publications has been minimal.
+There are very many challenges, some of which have been discussed before
+[ @Peng2011-ia; @Stodden2016-aj; @Munafo2017-gf ; @Diaba-Nuhoho2021-nt ].
 
-Researchers face several hurdles in adopting these.
+Due to the relative lack of systematic reproducibility/replication studies in bioinformatics,
+the severity of the reproducibility problem isn't well appreciated.
+The studies that have been conducted point to reproducibility rates of ~10% in computational
+life sciences [@Ioannidis2009-hv;@Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
+are required to bring attention to the issue.
 
-* The "publish-or-perish" incentive structures in academia reward prolific rather than scrupulous
-authors.
-Funding bodies, institutions and journals generally prefer to focus on novel research directions rather
+Journals are partly responsible, as their editorial policies influence author behaviours.
+Except for a few outliers, verification of computational reproducibility is not in the scope of peer
+review, but this should be considered.
+As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
+[@Peng2011-ia].
+
+Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
 than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
 @Munafo2017-ty; @Diong2021-is].
+This is a direct result of research assessment policies that incentivize journal-level prestige
+metrics and number of publications rather than the rigour [@Munafo2017-gf].
+A change in research assessment ("eg: track-record") that recognises and credits contributions toward
+reproducibility and rigor would help decrease funds wasted on sloppy research.
 
-* A lack of systematic reproducibility/replication studies in bioinformatics remains small
-[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the reproducibility
-problem isn't well appreciated.
-
-* Journal standards are slow to progress.
-Our observations indicate that sharing of data and code remains optional for most medical
-journals in PubMed, but in the field of computational biology, providing these resources is
-becoming standard practice.
-
-* A lack of documentated protocols and tutorial materials for highly-reproducible analysis makes it
-difficult for researchers to confidently update their code development practices.
-
-* An overall lack of coding/data science expertise among life science researchers and new graduates.
-Due to strong demand from industry and government, academia has a deficit of workers fluent in data
-science [@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
-This makes it challenging for organisations to recruit capable data analysts and cultivate a
-community of practice.
-
-Stodden et al have outlined some practical recommendations for tackling these problems [@Stodden2016-aj].
-
-Overcoming these obstacles will require a multi-pronged strategy and effort from different stakeholders.
-
-* What researchers can do.
-
-* What journals can do.
-
-* What insitutions can do.
-
-* What universities can do.
-
-* What funders can do.
-
-* Incentive structures will need to change.
-How we measure and value research achievement.
-A change in culture or research assessment.
-How we fund research data analytics and infrastructure.
-Value rigour as well as prolificness.
-
-* Journals need to progress towards publishing reproducible workflows.
-Peng 2011 speaks about the lack of infrastructure for disseminating reproducible research.
-
-* Enhanced training.
-A lack of technical computer skills in many research groups is another culprit.
-Specific workshop-based training in these areas could help [@Williams2023-ic].
+Individual researchers also face difficulties.
+Life science graduates often receive inadequate training in data science that places them in a
+disadvantage considering the data intensive nature of contemporary biology.
+This is compounded by the strong demand for data science experts in industry and government which
+leads to a deficit of academics fluent in data science
+[@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
+This makes it challenging for organisations to recruit and retain capable data analysts and cultivate
+a community of practice.
+Addressing this problem is not straightforward.
+It requires ehancement of data science skills in undergraduate life science courses, investment in
+upskilling the current workforce and offering working conditions that are competitive with other sectors.
+Specific workshop-based training for current researchers could help [@Williams2023-ic].
 The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
 [@Wilson2014-mh].
-Undergraduate and postgraduate courses in the life sciences would benefit from enhanced training in data
-science.
 
-* Enhanced documentation, protocols and tutorials.
-The most scalable solution would be to provide comprehensive beginner-friendly step-by-step
+A lack of documentated protocols and tutorial materials for highly reproducible analysis makes it
+difficult for researchers to confidently update their code development practices.
+Expert bioinformaticians can help by publishing comprehensive beginner-friendly step-by-step
 guides for frequently used routines.
+This approach is highly scalable and the reach of open-access publishing ensures these materials are
+broadly accesible.
 For example, our group has just released such a protocol that walks novice users through the process
@@
 @Ziemann2023-nm.
-More such protocols are needed to make extreme reproducibility mainstream.
+More guides like this are needed to make extreme reproducibility mainstream.
 A compendium of publicly available learning materials around the five pillar concepts are
 provided in the supplementary information.
-
-* What individuals can do.
-
-* Overall change in culture.
-Recognition that life sciences is a data intensive discipline.
 
@@
 significant expertise to analyze and distill key observations.
-The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
-Such reuse reduces waste and enhances the cost-effectiveness of research.
-The transparency provided by sharing data, code and documentation makes it easier for problems to be
-identified, so that erroneous findings are less likely to pollute the scientific literature.
+The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
+reducing research waste.
+Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
+pollute the scientific literature.
 Researchers, their institutions, publishers and funders each have a major role to play in fighting the
 reproducibility crisis by encouraging highly reproducible research practices.
-
-https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-021-05875-3#Sec2
 

### 2023-08-18 manuscript/rebuttal.Rmd
rev3

@@
 > reproducibility crisis by encouraging highly reproducible research practices.
+
+## Reviewer 3
+
+> The paper is mostly well written and addresses an important topic: computational reproducibility and
+> tools to achieve it. The authors describe five "pillars" under which various tools and techniques can
+> be categorized. The paper is pretty comprehensive in what it covers and has useful supplementary
+> material. My main issue with the paper is that it reiterates many things that have been covered in
+> other papers. The authors cite 11 such papers. Many of the topics covered in submitted paper have
+> already been covered well in other papers. I am most familiar with reference 13, which covers many of
+> the same topics, although this paper goes into more detail on many issues. The authors could do more
+> to differentiate their paper from previous ones and perhaps remove some topics that are already
+> covered well elsewhere.
+
+RESPONSE: XX
+
+> Aside from that, I have listed below some relatively minor issues that, if addressed, would improve the paper.
+When I list page numbers, I am using the PDF page numbers rather than the numbers shown in the top-left corner of the manuscript.
+
+> 2. Page 3, Line 36: "bioinformatics data analysts (not tool developers)".
+> The authors state these individuals as the primary audience, but some parts of the paper seem to be
+> targeted at a more technical audience. Or maybe I am misunderstanding the intent. If the audience
+> is "bioinformatics data analysts," that would imply people who are bioinformaticians but are
+> analyzing data rather than creating tools. However, a much bigger audience (and perhaps more
+> important audience) are non-bioinformaticians who analyze data.
+
+RESPONSE: The passage has been amended for clarity:
+
+> The intended audience is anyone who analyzes biological data.
+
+> 3. Page 3, Line 43: "enshrined by code" (this language is awkward)
+
+RESPONSE: The passage has been amended for clarity:
+
+> "... it needs to be formalised in code and ... "
+
+> 4. Page 3, Lines 45-47: What about tasks that cannot be automated? U see that this topic is addressed later. But this part implies that everything can be automated.
+
+> 5. Page 3, Line 52: It says that spreadsheets are "overused and misused." This is subjective and not backed by evidence, other than the well-known examples of gene symbols being formatted as dates.
+
+> 6.
+Page 3, Line 56: It is not necessarily true that analyses performed using web tools are not reproducible.
+Although rare, some web tools facilitate reproducibility by providing code or configuration files and/or allowing the apps to be executed locally.
+
+> 7. Page 5, Line 15, type-o = "authors provided along the"
+
+> 8. Page 6, Line 21: "quantum leap" (this term is overly optimistic in this context)
+
+> 9.
+How do you get from a notebook to an actual paper submission if you have to do custom formatting of the document, including references?
+My understanding is that this is still not possible, but please correct me if I'm wrong.
+
+> 10. What about when your data files are too large to fit on a personal computer?
+
+> 11. What about computationally intensive tasks that must be performed using specialized computing environments like the cloud or clusters?
+
+> 12. Page 6, line 32: The master script idea was already mentioned earlier.
+
+> 13.
+Section on version control: This section focuses mostly on using VC for software development (that is my interpretation).
+To be consistent with the introduction, it should focus more on data analyses.
+Although I use VC for analyses, I feel that simpler approaches are better in many cases.
+For example Dropbox and Google Drive provide some version-control and backup functionality and do not require the same level of knowledge as git.
+
+> 14. Page 7: Many data analysts will not know what JupyterLab or VS Code are. References are also needed.
+
+> 15. Page 9, line 7: References are missing for these other tools.
+
+> 16. Page 9, lines 7-8: I believe you, but I am not aware of evidence that supports this claim.
+
+> 17. The Biocontainers project should be mentioned.
+
+> 18. Page 10, lines 51-52: I disagree that the risk is small. There are many instances of using genomic data to identify individuals who have committed crimes.
+
+> 19. Figures 2 and 4 are very similar to figures used in reference 13.
+
+> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles. People can go to the source article for that.
+
+> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR. There are lots of problems with FAIRness in these repositories.
+
+> 22. Page 11, lines 13-15: That's not true for some of these disciplines. Ecology has NEON, evo bio has NCBI.
+
+> 23. Page 11, line 34: Need evidence to back this up.
+
+> 24.
+Page 11, line 39: It is not necessarily true that CSV files are better than Excel.
+Excel can retain information about data types, for example, whereas CSVs do not.
+It depends on what you are trying to accomplish.
+
+> 25.
+One thing that could be added is something about Common Workflow Language.
+It's a community-supported specification for accomplishing many of the objectives described here.
+There are some recent papers about this.
+
+> 26.
+Page 14, lines 9-10: The big question is how to make more progress.
+We have the tools to achieve reproducibility, but why are we rarely achieving it?
+The paper mentions incentives and lack of training, which are true.
+You might consider elaborating a bit.
+My cynical view is that writing more papers and tutorials will do little without strong incentives and more automation.

### 2023-08-18 manuscript/5pillars.Rmd
rev3

@@
 transparent reporting and the role of computational environments.
-The intended audience is bioinformatics data analysts (not tool developers), however the
+The intended audience is anyone who analyzes biological data (not tool developers), however the
 principles described here could equally apply to other domains of study.
@@
 
-For a bioinformatics project to be reproducible, it needs to be enshrined by code and include all
+For a bioinformatics project to be reproducible, it needs to be formalised in code and include all
 instructions, from inspecting the raw data to generating the outputs that form the conclusions of the
@@
 There are very many challenges, some of which have been discussed before
-[ @Peng2011-ia; @Stodden2016-aj; @Munafo2017-gf ; @Diaba-Nuhoho2021-nt ].
+[ @Peng2011-ia; @Stodden2016-aj; @Munafo2017-gf ; @Diaba-Nuhoho2021-nt].
 

### 2023-08-21 manuscript/rebuttal.Rmd
working on reviewer comments

@@
 
-> 4. Page 3, Lines 45-47: What about tasks that cannot be automated? U see that this topic is addressed later. But this part implies that everything can be automated.
+> 4. Page 3, Lines 45-47: What about tasks that cannot be automated? I see that this topic is addressed
+later. But this part implies that everything can be automated.
 
-> 5. Page 3, Line 52: It says that spreadsheets are "overused and misused." This is subjective and not backed by evidence, other than the well-known examples of gene symbols being formatted as dates.
+RESPONSE: The section has been amended to acknowledge that not all steps can be automated.
 
-> 6.
-Page 3, Line 56: It is not necessarily true that analyses performed using web tools are not reproducible.
-Although rare, some web tools facilitate reproducibility by providing code or configuration files and/or allowing the apps to be executed locally.
+> To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
+> possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
+
+> 5. Page 3, Line 52: It says that spreadsheets are "overused and misused." This is subjective and not
+> backed by evidence, other than the well-known examples of gene symbols being formatted as dates.
+
+RESPONSE: The passage in question has been replaced, with reference to evidence from a large
+survey that found that ~69% of university academics used MS Excel in their research activities.
+
+> Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
+> researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.
+
+> 6. Page 3, Line 56: It is not necessarily true that analyses performed using web tools are not
+> reproducible. Although rare, some web tools facilitate reproducibility by providing code or
+> configuration files and/or allowing the apps to be executed locally.
 

### 2023-08-21 manuscript/5pillars.Rmd
working on reviewer comments

@@
 
-For a bioinformatics project to be reproducible, it needs to be formalised in code and include all
-instructions, from inspecting the raw data to generating the outputs that form the conclusions of the
-study.
-This automated process removes the need for any manual steps, which are time-consuming and
+To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
+possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
+Automated processes remove the need for manual steps, which are time-consuming and
 prone to errors.
@@
 which would be difficult for graphical tools like spreadsheets or webtools.
-Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
-somewhat overused and misused in the life sciences, leading to problems like inadvertent conversion
-of gene names to dates in databases and supplementary Excel files [@Zeeberg2004-js;
-@Ziemann2016-pv; @Abeysooriya2021-uz].
+Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
+leading to problems like inadvertent conversion of gene names to dates in databases and supplementary
+Excel files [@Zeeberg2004-js; @Ziemann2016-pv; @Abeysooriya2021-uz].
+Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
+researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.
+
 While webtools are valuable for data exploration, there are worries that they undermine

### 2023-08-21 manuscript/rebuttal.Rmd
progress on reviewer 3

@@
 
-RESPONSE: The passage has been amended for clarity:
+RESPONSE: The passage has been amended:
 
@@
 
+RESPONSE: Thank you for raising this point. We have amended the section in question:
+
+> While webtools are valuable for data exploration, there are worries that they undermine
+> reproducibility for the sake of convenience [@Wijesooriya2022-ly].
+> Transferring data between compute platforms is also discouraged.
+> For example, having workflows that involve combinations of web-based and scripted tools require
+> data transfer steps which are inefficient and error-prone.
+> On the other hand, some web based tools excel at reproducibility.
+> The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
+> bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
+> [@The_Galaxy_Community2022-dt;@Reich2006-na].
+> Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
+> or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).
+
 > 7. Page 5, Line 15, type-o = "authors provided along the"
+
+RESPONSE: The passage in question has been amended:
+
+> In this work, authors provided alongside the traditional research paper a Jupyter notebook ..
 
@@
 
-> 9.
-How do you get from a notebook to an actual paper submission if you have to do custom formatting of the document, including references?
-My understanding is that this is still not possible, but please correct me if I'm wrong.
+RESPONSE: The passage in question has been amended:
+
+> These features make literate programming a useful tool for science communication in a range of
+> situations.
+
+> 9. How do you get from a notebook to an actual paper submission if you have to do custom formatting
+> of the document, including references?
+> My understanding is that this is still not possible, but please correct me if I'm wrong.
+
+RESPONSE: It is possible to achieve accurate journal-specific formatting of references with a Rmarkdown
+document.
+It requires a few things:
+
+* A .bib file that contains the bibliography in bibtex format.
+
+* A .csl file that specifies the citation style.
+These are avaialble from [GitHub](https://github.com/citation-style-language/styles).
+
+* Modifcation to the Rmd header section that specifies the path of the bib and csl files.
+
+This enables convenient reformatting of the bibliography for different venues.
+In terms of overall document formatting, the `rticles` package provides a suite of
+custom R Markdown LaTeX formats suitable for many journals.
+
+We have amended the Literate Programming section to mention these features briefly:
+
+> Quarto and R Markdown have the ability to automatically generate a bibliography in many different
+> journal styles [@noauthor_undated-pg].
+> Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
+> freely available @Xie2018-rf.
 
@@
 
-> 11. What about computationally intensive tasks that must be performed using specialized computing environments like the cloud or clusters?
+RESPONSE: When data files are too large for a PC, it is common to use high performance computers, shared
+clusters or cloud computers.
+These computers are typically accessed via SSH (headless mode), so Quarto and Rstudio graphical
+interfaces might not be available.
+In those cases, the commands `quarto render document1.qmd` and `rmarkdown::render("document2.Rmd")` will
+work for Quarto and R Markdown respectively.
+
+> 11. What about computationally intensive tasks that must be performed using specialized computing
+> environments like the cloud or clusters?
+
+RESPONSE: We have amended the relevant section to suggest a specific workflow manager for compute
+intensive applications.
+
+> Projects involving computationally intensive tasks might benefit from a build/workflow automation
+> solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
+> `nextflow` [@Di_Tommaso2017-yo].
 
@@
 
-> 13.
-Section on version control: This section focuses mostly on using VC for software development (that is my interpretation).
-To be consistent with the introduction, it should focus more on data analyses.
-Although I use VC for analyses, I feel that simpler approaches are better in many cases.
-For example Dropbox and Google Drive provide some version-control and backup functionality and do not require the same level of knowledge as git.
+RESPONSE: True. We have amended the "Documentation" section to remove the unnecessary reiteration:
+
+> As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
+> minimising the number of commands required.
+
+> 13. Section on version control: This section focuses mostly on using VC for software development
+> (that is my interpretation).
+> To be consistent with the introduction, it should focus more on data analyses.
+> Although I use VC for analyses, I feel that simpler approaches are better in many cases.
+> For example Dropbox and Google Drive provide some version-control and backup functionality and do
+> not require the same level of knowledge as git.
 

### 2023-08-21 manuscript/5pillars.Rmd
progress on reviewer 3

@@
 data transfer steps which are inefficient and error-prone.
+On the other hand, some web based tools excel at reproducibility.
+The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
+bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
+[@The_Galaxy_Community2022-dt;@Reich2006-na].
+Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
+or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).
 
-To facilitate an end-to-end automated process, the code and data need to be "linked", which means the
+To facilitate an end-to-end automated process, code and data need to be "linked", which means the
 code is aware of the location of the data and how to get it automatically @Peng2011-ia.
@@
 Unix Shell.
+These approaches work equally well on personal computers, cloud servers and high performance clusters.
 Using free and open source software ensures that the research is accessible to the greatest audience
@@
 the scientific community.
-Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
-`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
+Projects involving computationally intensive tasks might benefit from a build/workflow automation
+solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
 `nextflow` [@Di_Tommaso2017-yo].
@@
 has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
-In this work, authors provided along the traditional research paper a
+In this work, authors provided alongside the traditional research paper a
 [Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
@@
 Outputs include HTML documents, but also slideshows and even ebooks.
+Quarto and R Markdown have the ability to automatically generate a bibliography in many different
+journal styles [@noauthor_undated-pg].
+Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
+freely available @Xie2018-rf.
 
-These features make literate programming a quantum leap for science communication in a range of
+These features make literate programming a useful tool for science communication in a range of
 situations.
@@
 
-As outlined above, one of the goals is to reduce the complexity of reproduction, and using a master
-script is a simple way to do this.
-It also makes the `README` documentation much simpler.
+As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
+minimising the number of commands required.
+This also makes the `README` documentation much simpler.
 For example, the pathway analysis workflow we previously developed could be reproduced and inspected

### 2023-08-22 manuscript/rebuttal.Rmd
working on rev3

@@
 
-> Aside from that, I have listed below some relatively minor issues that, if addressed, would improve the paper.
-When I list page numbers, I am using the PDF page numbers rather than the numbers shown in the top-left corner of the manuscript.
+> Aside from that, I have listed below some relatively minor issues that, if addressed, would improve
+> the paper. When I list page numbers, I am using the PDF page numbers rather than the numbers shown in
+> the top-left corner of the manuscript.
 
@@
 
-> 14. Page 7: Many data analysts will not know what JupyterLab or VS Code are. References are also needed.
+RESPONSE: In accordance with this suggestion, we have rephrased some of the text to highlight the
+utility of DVCS for data analysts.
+
+> Although DVCSs come from the world of software engineering, data analysts can significantly
+> benefit from their use in the life sciences [@ram2013; @blischak2016; @perez-riverol2016].
+
+The bullet points following this passage outline the benefits of using DVCS.
+While cloud storage like dropbox could protect against code loss, they cannot other five points cannot
+be satisfied.
+
+> 14. Page 7: Many data analysts will not know what JupyterLab or VS Code are.
+References are also needed.
+
+RESPONSE: The phrase has been updated accordingly:
+
+> `git` is typically used at the command line, however it is also incorporated into integrated
+> development environments commonly used in bioinformatics including Rstudio, JupyterLab [68,69] and
+> VS Code [70,71].
 
@@
 
+TODO: Podman, Rkt or Apptainer/Singularity, Docker need refs.
+
 > 16. Page 9, lines 7-8: I believe you, but I am not aware of evidence that supports this claim.
+
+TODO: "Docker remains the most widely used containerisation system for web development and research"
 
@@
 
-> 18. Page 10, lines 51-52: I disagree that the risk is small. There are many instances of using genomic data to identify individuals who have committed crimes.
+RESPONSE: Biocontainers is mentioned in the last paragraph of the "Compute environment control" section
+and has reference (Ref #91).
+
+> 18. Page 10, lines 51-52: I disagree that the risk is small.
+> There are many instances of using genomic data to identify individuals who have committed crimes.
+
+RESPONSE: In light of this, we have changed the section in question:
+
+> While deidentification of data has been widely used to safeguard participant privacy, this may not be
+> suitable for genomics data due to the availability of existing public genetic data sets that heighten
+> reidentification risks [105].
+> For example, the surnames of some male participants can be inferred based on the Y chromosome
+> variants seen in deidentified genomic data together with public geneology web sites [106].
+> To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
+> Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
+> frameworks designed to protect personal data in a way that perserves reproducibility [107,108].
 
@@
 
-> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles. People can go to the source article for that.
+Figure 4 is adapted from the diagram on the Docker official website, which was originally published in
+2015 or earlier (https://web.archive.org/web/20151023003338/https://www.docker.com/what-docker).
+We have cited the original source.
 
-> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR. There are lots of problems with FAIRness in these repositories.
+Figure 2 is somewhat different from reference 13, as it shows an Rstudio window with windows for raw code
+in text format beside the output document.
+The figure in reference 13 shows a Jupyter notebook window, which has a different arrangement of code
+and outputs.
 
-> 22. Page 11, lines 13-15: That's not true for some of these disciplines. Ecology has NEON, evo bio has NCBI.
+> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles.
+> People can go to the source article for that.
+
+RESPONSE: We acknowledge that it can sometimes be tiresome to reiterate things like the FAIR principles.
+However summarising the key points succinctly will be in the interests of readers with limited time.
+We note that the original description of the FAIR principles by Wilkinson and colleagues is ~5,000 words,
+while this entire manuscript is ~6,500. We note that other reviewers did not insist on removing this
+section.
+
+> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
+There are lots of problems with FAIRness in these repositories.
+
+RESPONSE: While there may be problems with FAIRness with NCBI GEO and SRA, these repositries mostly
+satisfy FAIR principles (eg: https://www.biorxiv.org/content/10.1101/739334v1.full.pdf).
+From the perspective of a researcher sharing genomic data, they are "good enough" as compared to the
+well known alternatives (eg: general purpose repository, institutional data storage, supplementary file,
+dropbox, etc.).
+
+> 22. Page 11, lines 13-15: That's not true for some of these disciplines. Ecology has NEON, evo bio
+has NCBI.
+
+RESPONSE: We acknowledge that NCBI is a well known repository for evolutionary biology.
+We have altered the section for clarity:
+
+> The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
+> FAIR repositories that accepts data from other domains of study like ecology,
+> physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
 
@@
 
-> 24.
-Page 11, line 39: It is not necessarily true that CSV files are better than Excel.
+TODO: "Supplementary data files are not FAIR."
+
+> 24. Page 11, line 39: It is not necessarily true that CSV files are better than Excel.
 Excel can retain information about data types, for example, whereas CSVs do not.
@@
 
-> 25.
-One thing that could be added is something about Common Workflow Language.
+TODO:
+
+> 25. One thing that could be added is something about Common Workflow Language.
 It's a community-supported specification for accomplishing many of the objectives described here.
@@
 
-> 26.
-Page 14, lines 9-10: The big question is how to make more progress.
+TODO:
+
+> 26. Page 14, lines 9-10: The big question is how to make more progress.
 We have the tools to achieve reproducibility, but why are we rarely achieving it?
@@
 You might consider elaborating a bit.
-My cynical view is that writing more papers and tutorials will do little without strong incentives and more automation.
+My cynical view is that writing more papers and tutorials will do little without strong incentives
+and more automation.

### 2023-08-22 manuscript/5pillars.Rmd
working on rev3

@@
 
-Computational researchers can benefit significantly by adopting DVCS into their routine
-practices.
-Below we outline some of the benefits.
+Although DVCSs come from the world of software engineering, data analysts can significantly
+benefit from their use in the life sciences [@ram2013; @blischak2016; @perez-riverol2016].
+Below we outline some of these advantages.
 
@@
 
-`git` is typically used at the command line, however it is also integrated into Rstudio, JupyterLab and
-VS Code, and there are also several `git` clients with graphical interfaces which better allow inspection
-of code changes.
-
-Finally, although DVCSs come from the world of software engineering,
-their use has become a standard in computational biology, as witnessed by the numerous guides
-and best practices on the subject [@ram2013; @blischak2016; @perez-riverol2016].
+`git` is typically used at the command line, however it is also incorporated into integrated development
+environments commonly used in bioinformatics including Rstudio, JupyterLab
+[@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].
+There are also several `git` clients with graphical interfaces which better allow inspection
+of code changes (e.g.: @noauthor_undated-rh, @noauthor_undated-kj).
 
@@
 Some may keep data private due to concerns over participant privacy.
-This is a well-founded concern for identifiable data, however once identifying features are removed,
-it is nearly impossible to re-identify genomic data @Venkatesaramani2021-tn unless the attackers have
-pre-existing genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
-Therefore, sharing de-identified genomic data and metadata is likely to carry little risk to participant
-privacy.
-Still, much human genomic data and metadata resides in restricted databases due to the scope of
-data sharing that was agreed to in the consent process.
-For example, the consent form may only allow sharing of data to researchers investigating certain
-diseases, which means a vetting process must be undertaken before sharing data with external
-investigators.
-In the interests of reuse and reproducibility, future studies could consider a consent form that allows
-more broad sharing of de-identified data.
+
+While deidentification of data has been widely used to safeguard participant privacy, this may not
+be suitable for genomics data due to the availability of existing public genetic data sets that heighten
+reidentification risks @Heeney2011-ri.
+For example, the surnames of some male participants can be inferred based on the Y chromosome variants
+seen in deidentified genomic data together with public geneology web sites @Gymrek2013-fz.
+To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
+Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
+frameworks designed to protect personal data in a way that perserves reproducibility
+[@Rehm2021-kt;@Thorogood2021-kt].
 
@@
 
-High throughput biology has already built a number of data repositories for specific data types
-conforming to FAIR principles (e.g.: GEO, SRA, ENA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
-However other domains lack well known specialized organized data repositories, such as evolutionary
-biology, ecology, cell biology, molecular simulation, social sciences and computing.
-The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
-repository that accepts data from these fields [@Pampel2013-ep].
+High throughput biology has already built a number of data repositories for specific data types mostly
+conforming to FAIR principles.
+Repositories like GEO, SRA, ENA and PRIDE were developed in the aftermath of the draft human
+genome assembly and have become well known to omics researchers
+[@Edgar2002-hs; @Kodama2012-bn; Martens2005-wo]).
+The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
+FAIR repositories that accepts data from other domains of study like ecology,
+physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
 If no specialized repository exists, then a general-purpose repository such as

### 2023-08-22 manuscript/5pillars.Rmd
pillars

@@
 
-```{r,5pillars,fig.cap="Figure 1. The five pillars of computational reproducibility.",echo=FALSE,fig.height=6}
+```{r,5pillars,fig.cap="Figure 1. The five pillars of computational reproducibility.",echo=FALSE,fig.height=5}
+
+par(mar = c(0.1,0.1,0.1,0.1))
 
@@
 cols2 <- adjustcolor( cols, alpha.f = 0.7)
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 9))
 polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
@@
 points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
-text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)
-#mtext("Five pillars of reproducible computational research", cex=1.5)
+text(5,9,lab="Five pillars of reproducible computational research", cex=1.5)
 
@@
 text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
-text( 5 , 0.2 , lab="End-to-end automated process")
-#text(5,7,lab="Code testing")
-#text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
+text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
+text(5,7,lab="Code testing",cex=1.2)
 
@@
 
-```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}
+```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=8}
 
+library(png)
+library(grid)
+
+cols <- brewer.pal(5, "Set2")
 cols2 <- adjustcolor( cols, alpha.f = 0.4)
 
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
+par(mar = c(0.1,0.1,0.1,0.1))
+
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(-5, 8.5))
 polygon(c(-0.6,3,3,-0.6),c(0.8,0.8,5.2,5.2), col=cols2[1], border = NA)
@@
 text(c(8.2) ,c(6.2), labels="pull",srt=-60)
+
+mypng = readPNG('../img/version_control.png')
+grid.raster(mypng, x=.55, y=.2, width=0.85)
 
@@
 library("RColorBrewer")
+par(mar = c(0.1,0.1,0.1,0.1))
 cols <- brewer.pal(5, "Set2")

### 2023-08-23 manuscript/rebuttal.Rmd
rev3 nearly done

@@
 
-### Reviewer 1
+## Reviewer 1
 
@@
 
-### Reviewer 2
+## Reviewer 2
 
@@
 
+### R2 Point 1
+
 > 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
@@
 
-To demonstrate the importance of computational reprodcibility, we elaborate on the
+RESPONSE: To demonstrate the importance of computational reprodcibility, we elaborate on the
 "forensic bioinformatics" case study of Baggerly and Coombes (2010) critique of Potti et al (2006)
@@
 
+### R2 Point 2
+
 > 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the
@@
 
+### R2 Point 3
+
 > 3. Focus on Implementation Challenges: While the article acknowledges the importance of certain
@@
 funders, journals, institutions and individual researchers).
+
+### R2 Point 4
 
@@
 
+### R2 Point 5
+
 > 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of
@@
 
+### R3 Point 1
+
 > The paper is mostly well written and addresses an important topic: computational reproducibility and
@@
 
+### R3 Point 2
+
 > 2. Page 3, Line 36: "bioinformatics data analysts (not tool developers)".
@@
 
+### R3 Point 3
+
 > 3. Page 3, Line 43: "enshrined by code" (this language is awkward)
@@
 > "... it needs to be formalised in code and ... "
+
+### R3 Point 4
 
@@
 
+### R3 Point 5
+
 > 5. Page 3, Line 52: It says that spreadsheets are "overused and misused." This is subjective and not
@@
 > researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.
+
+### R3 Point 6
 
@@
 
+### R3 Point 7
+
 > 7. Page 5, Line 15, type-o = "authors provided along the"
@@
 > In this work, authors provided alongside the traditional research paper a Jupyter notebook ..
+
+### R3 Point 8
 
@@
 > situations.
+
+### R3 Point 9
 
@@
 
+### R3 Point 10
+
 > 10. What about when your data files are too large to fit on a personal computer?
@@
 work for Quarto and R Markdown respectively.
+
+### R3 Point 11
 
@@
 
+### R3 Point 12
+
 > 12. Page 6, line 32: The master script idea was already mentioned earlier.
@@
 > minimising the number of commands required.
+
+### R3 Point 13
 
@@
 
+### R3 Point 14
+
 > 14. Page 7: Many data analysts will not know what JupyterLab or VS Code are.
@@
 
+### R3 Point 15
+
 > 15. Page 9, line 7: References are missing for these other tools.
 
-TODO: Podman, Rkt or Apptainer/Singularity, Docker need refs.
+RESPONSE: Citations have been added for Podman and Apptainer/Singularity.
+Rkt was removed, as it appears that development of the project has ceased.
+
+### R3 Point 16
 
@@
 
+### R3 Point 17
+
 > 17. The Biocontainers project should be mentioned.
@@
 and has reference (Ref #91).
+
+### R3 Point 18
 
@@
 
+### R3 Point 19
+
 > 19. Figures 2 and 4 are very similar to figures used in reference 13.
@@
 
+### R3 Point 20
+
 > 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles.
@@
 
+### R3 Point 21
+
 > 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
@@
 dropbox, etc.).
+
+### R3 Point 22
 
@@
 
+### R3 Point 23
+
 > 23. Page 11, line 34: Need evidence to back this up.
 
-TODO: "Supplementary data files are not FAIR."
+RESPONSE: Reviewer 3 commented on the passage saying supplementary data files are less
+findable and accessible as compared to FAIR data repositories.
+We note that Roche and colleagues previously stated that [@Roche2015-aq]:
+
+> There are no standards for organizing supplementary data both within and across journals
+> and such data are often not readily discoverable or openly accessible
+> (to those without a relevant journal subscription, for example).
+
+They also cite an article by Caetano and Aisenberg that speaks directly to the findability and
+accessibility @Caetano2014-bi:
+
+> Many journals provide the option for authors to
+> include supplementary data that are made available in the digital version of the publication.
+> However, repositories increase data discoverability, are more reliable for long-term storage and
+> make it possible to make data sets available under open access even
+> when authors transfer copyright ownership to publishers.
+
+In light of this argument, we have added a citation to Caetano and Aisenberg in the updated text.
+
+### R3 Point 24
 
@@
 
-TODO:
+RESPONSE: It is true that in some cases, Excel files are more convenient to view.
+However, in the context of large data sets with thousands of rows and their analytical
+reproducibility, Excel is a non-starter.
+For example, it is well known that the behaviour of Excel changes between versions.
+Earlier versions have limitations around the number of allowed rows and columns.
+Small changes to Excel files such as merging cells can break machine readibility.
+Excel users often use cell highlighting or text colour to encode some information which isn't
+easily machine readable.
+The simplicity of TSV and CSV formats means they are readily and rapidly ingested by commonly used
+programming languages without needing to install extra packages.
+In line with this, the point in question has been amended:
+
+> 5. Use file formats that are machine-readable and compatible with many different types of software.
+> For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
+> specifications that are suitable for very large data.
+
+### R3 Point 25
 
@@
 
+### R3 Point 26
+
 > 26. Page 14, lines 9-10: The big question is how to make more progress.
-We have the tools to achieve reproducibility, but why are we rarely achieving it?
-The paper mentions incentives and lack of training, which are true.
-You might consider elaborating a bit.
-My cynical view is that writing more papers and tutorials will do little without strong incentives
-and more automation.
+> We have the tools to achieve reproducibility, but why are we rarely achieving it?
+> The paper mentions incentives and lack of training, which are true.
+> You might consider elaborating a bit.
+> My cynical view is that writing more papers and tutorials will do little without strong incentives
+> and more automation.

### 2023-08-23 manuscript/5pillars.Rmd
rev3 nearly done

@@
 promise of reproducibility.
-While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
-containerisation system for web development and research.
+While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
+[@Kurtzer2017-zq], Docker remains the most widely used containerisation system for web development and research.
 Docker now has a large community of users, extensive documentation and
@@
 3. Avoid large supplementary data files accompanying journal articles as these are less findable and
-accessible.
+accessible [@Caetano2014-bi].
 
@@
 
-5. Use file formats that are compatible with many different types of software; e.g.: comma-separated
-values (CSV) files are preferred over Excel files.
+5. Use file formats that are machine-readable and compatible with many different types of software.
+For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
+specifications that are suitable for very large data.
 

### 2023-08-24 manuscript/rebuttal.Rmd
rev3 again

@@
 
-RESPONSE: XX
+RESPONSE: We thank reviewer 3 for their careful assessment of our article and appreciate the suggestions
+which have improved the manuscript.
+Indeed, this article does reiterate a great deal of what has already been said, it is a review after
+all.
+We note that each of those recommendation articles mentioned have different emphases.
+For example, some of them don't speak to the benefits of literate programming,
+while others do not mention the importance of documentation.
+Some are focused on specific concepts like infrastructure, containerisation or
+continuous analysis.
+Some are proof of principle workflows, which present a rigid combination of tools.
+Here, we describe the ways computational researchers can make their own work more reproducible.
+In addition to the Five Pillars infographic, the specific combination of recommendations
+is unique.
+And ours is the only such article that provides a handy list of recommended tutorials as a supplementary
+resource.
 

### 2023-08-24 manuscript/5pillars.Rmd
rev3 again

@@
 [ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
-@Gruning2018-sg; @Akalin2018-eg; @Brito2020-ld; @Peikert2021-eq;
-@Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv; @Stodden2016-aj].
-Inspired by these principles, this article presents the
-*five pillars of reproducible computational research* (**Figure 1**).
-There are clear parallels with the established three pillars of the open science framework,
-(open data, code and papers) [@Gorgolewski2016-hv].
+@Stodden2016-aj; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Akalin2018-eg;
+@Brito2020-ld; @Peikert2021-eq; @Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv].
+Inspired by these principles, we presents the *five pillars of reproducible computational research*
+(**Figure 1**).
 Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
@@
 principles described here could equally apply to other domains of study.
+There are clear parallels with the established three pillars of the open science framework,
+(open data, code and papers) [@Gorgolewski2016-hv].
 
-```{r,5pillars,fig.cap="Figure 1. The five pillars of computational reproducibility.",echo=FALSE,fig.height=5}
+```{r,5pillars,fig.cap="Figure 1. The five pillars of computational reproducibility.",echo=FALSE,fig.height=4.5,fig.width=6}
 
@@
 text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
-text(5,7,lab="Code testing",cex=1.2)
+text(5,7,lab="Process validation & code testing",cex=1.2)
 
@@
 
-```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=8}
+```{r,dvcs,fig.cap="Figure 3.
+Distributed version control.
+(A) Each author has a copy of the repository where their contributions are committed before being pushed to the main central repository.
+Adapted from @Chacon2014-ys.
+(B) An example hypothetical git workflow history for a research project involving a team of three authors.
+Circles represent code commits.
+Path divergences create separate branches for independent development.
+Horizontal paths indicate code changes for a particular branch.
+Path convergences indicate where branch specific differences are incorporated into the main branch.
+From @ram2013.",echo=FALSE,fig.height=5.2}
+
+suppressMessages({
 
@@
 
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(-5, 8.5))
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(-3.5, 13.5), ylim=c(-3.5, 7.5))
 polygon(c(-0.6,3,3,-0.6),c(0.8,0.8,5.2,5.2), col=cols2[1], border = NA)
@@
 
-polygon(c(2.5,7.3,7.3,2.5),c(7,7,8.5,8.5), col=cols[4], border = NA)
+polygon(c(2.5,7.3,7.3,2.5),c(7,7,7.7,7.7), col=cols[4], border = NA)
 
-labs <- c("Developer 1",
+labs <- c("Author 1",
   "Working Files",
   "Local Repository",
-  "Developer 2",
+  "Author 2",
   "Working Files",
   "Local Repository",
-  "Developer 3",
+  "Author 3",
   "Working Files",
@@
 
-text( c(1.2,1.2,1.2,4.9,4.9,4.9,8.5,8.5,8.5), c(0,1.5,4.5,0,1.5,4.5,0,1.5,4.5), labels=labs )
-text(4.9,7.7,labels="Main Repository",cex=1.5)
+text( c(1.3,1.3,1.3,4.9,4.9,4.9,8.5,8.5,8.5), c(0.5,1.5,4.5,0.5,1.5,4.5,0.5,1.5,4.5), labels=labs )
+text(4.9,7.3,labels="Main Repository",cex=1.3)
 
-arrows( 1, 2, 1, 4 ,lwd=3)
-arrows( 1.4, 4, 1.4, 2 ,lwd=3)
-arrows( 4.7, 2, 4.7, 4 ,lwd=3)
-arrows( 5.1, 4, 5.1, 2 ,lwd=3)
-arrows( 8.3, 2, 8.3, 4 ,lwd=3)
-arrows( 8.7, 4, 8.7, 2 ,lwd=3)
-arrows( 1, 5, 2.5, 7 ,lwd=3)
-arrows( 2.9, 7 , 1.4, 5 ,lwd=3)
-arrows( 4.7, 5, 4.7, 7 ,lwd=3)
-arrows( 5.1, 7 , 5.1, 5 ,lwd=3)
-arrows( 7.3, 7, 8.7, 5 ,lwd=3)
-arrows( 8.3, 5, 6.9, 7 ,lwd=3)
+library("shape")
+Arrows( 1, 2, 1, 4 ,lwd=3, arr.type="simple")
+Arrows( 1.4, 4, 1.4, 2 ,lwd=3,arr.type="simple")
+Arrows( 4.7, 2, 4.7, 4 ,lwd=3, arr.type="simple")
+Arrows( 5.1, 4, 5.1, 2 ,lwd=3, arr.type="simple")
+Arrows( 8.3, 2, 8.3, 4 ,lwd=3, arr.type="simple")
+Arrows( 8.7, 4, 8.7, 2 ,lwd=3, arr.type="simple")
+Arrows( 1, 5, 2.5, 7 ,lwd=3, arr.type="simple")
+Arrows( 2.9, 7 , 1.4, 5 ,lwd=3, arr.type="simple")
+Arrows( 4.7, 5, 4.7, 7 ,lwd=3, arr.type="simple")
+Arrows( 5.1, 7 , 5.1, 5 ,lwd=3, arr.type="simple")
+Arrows( 7.3, 7, 8.7, 5 ,lwd=3, arr.type="simple")
+Arrows( 8.3, 5, 6.9, 7 ,lwd=3, arr.type="simple")
 
@@
 text(c(1.7,5.4,9) ,c(3,3,3), labels="update",srt=-90)
-text(c(1.5) ,c(6.2), labels="push",srt=50)
-text(c(2.4) ,c(5.9), labels="pull",srt=50)
+text(c(1.5) ,c(6.2), labels="push",srt=45)
+text(c(2.4) ,c(5.9), labels="pull",srt=45)
 text(c(4.4) ,c(6), labels="push",srt=90)
 text(c(5.4) ,c(6), labels="pull",srt=-90)
-text(c(7.3) ,c(5.9), labels="push",srt=-60)
-text(c(8.2) ,c(6.2), labels="pull",srt=-60)
+text(c(7.3) ,c(5.9), labels="push",srt=-55)
+text(c(8.2) ,c(6.2), labels="pull",srt=-55)
+
+text(c(1) ,c(7.5), labels="A",cex=1.7)
+text(c(-3.5) ,c(0.5), labels="B",cex=1.7)
 
 mypng = readPNG('../img/version_control.png')
-grid.raster(mypng, x=.55, y=.2, width=0.85)
+grid.raster(mypng, x=.5, y=.16, width=1)
+
+})
 
@@
 
-```{r,containers,fig.cap="Figure 4. Architecture of containers (left) compared to VMs (right). Adapted from @noauthor_2021-lt.",echo=FALSE,fig.height=6}
+```{r,containers,fig.cap="Figure 4. Architecture of containers (left) compared to VMs (right). Adapted from @noauthor_2021-lt.",echo=FALSE,fig.height=5}
 
@@
 cols2 <- adjustcolor( cols, alpha.f = 0.7)
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 12), ylim=c(0, 8))
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 11.5), ylim=c(0, 7))
 
@@
 lines(c(6,6,11.5,11.5),c(6.2,6.5,6.5,6.2)  )
-text(8.5,6.7,labels="Virtualized applications")
+text(9,6.7,labels="Virtualized applications")
 

### 2023-08-25 manuscript/rebuttal.Rmd
rev3 done

@@
 
-Response: We thank reviewer 1 for their positive feedback.
+RESPONSE: We thank reviewer 1 for their positive feedback.
 With regards to the ordering of the five pillars, we have amended Figure 1 so that the pillars now
@@
 Response: We have completely rewritten the conclusion.
-Instead of reiterating the five pillars, we focus on the motivation for adoption:
+Instead of reiterating the five pillars, we provide the following:
 
-> Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
-> of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
-> These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
-> significant expertise to analyze and distill key observations.
-> The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
-> reducing research waste.
-> Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
-> pollute the scientific literature.
+> As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
+> bioinformatics research, and current standards are falling short.
+> If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
+> such disasters from happening again.
 > Researchers, their institutions, publishers and funders each have a major role to play in fighting the
@@
 Here, we describe the ways computational researchers can make their own work more reproducible.
-In addition to the Five Pillars infographic, the specific combination of recommendations
-is unique.
-And ours is the only such article that provides a handy list of recommended tutorials as a supplementary
-resource.
+We are assuming that the readers are not advanced analysts, so some topics like pipelining systems
+are covered in less detail and we refer the reader to those relevant reviews.
+What does make this article unique is the Five Pillars infographic, and specific combination of
+recommendations.
+Our article also provides a handy list of recommended tutorials as a supplementary resource.
 
@@
 
-TODO: "Docker remains the most widely used containerisation system for web development and research"
+RESPONSE: The claim made in the manuscript is that "Docker remains the most widely used
+containerisation system for web development and research".
+We do not have exact usage data on this, however we can look at other metrics like GitHub stars.
+
+| Container System | GitHub stars* |
+| --- | --- |
+| Docker | 72.2k |
+| Podman | 18.9k |
+| Singularity | 3.1k |
+
+* 25th August.
+
+Based on this we have altered the text:
+
+> While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
+> [@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
+> stars).
 
@@
 
-TODO:
+RESPONSE: Thanks for raising this. We have added CWL to the section mentioning pipeline solutions, and
+refered readers to dedicated reviews on the topic.
+
+> Projects involving computationally intensive tasks would benefit from a build/workflow automation
+> solution.
+> Guides for selecting and using such systems have been already reviewed by others [@Leipzig2016-bf;
+> @Reiter2021-lh].
+> But it is worth mentioning some of the most used solutions in bioinformatics, which include
+> `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021], `CWL` [@Crusoe2022-cb],
+> `WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
+> The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
+> for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
+> need to be repeated.
+> After fixing the issue, re-running the analysis would pick up at step 8 again,
+> which saves labor and compute time.
 
@@
 > and more automation.
+
+RESPONSE: Thanks for raising this, and I note that similar sentiments were shared by reviewer 2 about the
+challenges that are holding back change.
+Please refer to our response to Reviewer 2's Point #3.
+The Challenges section now includes brief mentions of incentives and automation.

### 2023-08-25 manuscript/5pillars.Rmd
rev3 done

@@
 cols2 <- adjustcolor( cols, alpha.f = 0.7)
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 9))
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 9.7), ylim=c(0, 9))
 polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
@@
 polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
-polygon(c(-0.2,5,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
-points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
-text(5,9,lab="Five pillars of reproducible computational research", cex=1.5)
+polygon(c(-0.2,4.9,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
+points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=16,col=cols )
+text(4.8,9,lab="Five pillars of reproducible computational research", cex=1.5)
 
@@
 the scientific community.
-Projects involving computationally intensive tasks might benefit from a build/workflow automation
-solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
-`nextflow` [@Di_Tommaso2017-yo].
+Projects involving computationally intensive tasks would benefit from a build/workflow automation
+solution.
+Guides for selecting and using such systems have been already reviewed by others [@Leipzig2016-bf;
+@Reiter2021-lh].
+But it is worth mentioning some of the most used solutions in bioinformatics, which include
+`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021], `CWL` [@Crusoe2022-cb],
+`WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
 The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
@@
 need to be repeated.
-After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
+After fixing the issue, re-running the analysis would pick up at step 8 again,
 which saves labor and compute time.
@@
 While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
-[@Kurtzer2017-zq], Docker remains the most widely used containerisation system for web development and research.
-Docker now has a large community of users, extensive documentation and
+[@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
+stars as of August 2023).
+Docker has a large community of users, extensive documentation and
 a vast collection of pre-built container images in the DockerHub registry.
@@
 license.
+A recently described set of recommendations for making software FAIR describes further measures
+that data analysts could use to enhance their code documentation.
 
@@
 relative to the overall research project.
+
+<br>
+
+### Process validation and testing
+
+When these principles are not integrated correctly and checked, problems readily emerge
+[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
+This is why regular code testing is another recommended best practice, especially after making updates
+to the code or data [@Rule2019-eh; @The_Turing_Way_Community2019-ig].
+It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
+scientific community would fare reproducing such a workflow.
+
+Moreover, some have suggested adapting software development
+principles of unit tests and continuous integration/continuous delivery.
+This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
+made to a project's code base [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
+If the workflow executes successfully and tests are passed, the repository can be decorated with a
+"badge" to inform users of the lack of errors in the code.
+If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
+article can be assessed computationally, then such tests could be used to demonstrate computational
+reproducibility without the need for individuals to execute the workflow themselves.
+Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
+"temple" of reproducible research.
 
@@
 The studies that have been conducted point to reproducibility rates of ~10% in computational
-life sciences [@Ioannidis2009-hv;@Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
+life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
 are required to bring attention to the issue.
 
-Journals are partly responsible, as their editorial policies influence author behaviours.
+Journals are partly responsible, as their editorial policies influence author behaviours [@Han2017-pz].
 Except for a few outliers, verification of computational reproducibility is not in the scope of peer
-review, but this should be considered.
+review, but automating this in the publication process should be considered.
 As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
@@
 This is a direct result of research assessment policies that incentivize journal-level prestige
-metrics and number of publications rather than the rigour [@Munafo2017-gf].
+metrics and number of publications rather than the rigor [@Munafo2017-gf; @Oransky2023-zb].
 A change in research assessment ("eg: track-record") that recognises and credits contributions toward
@@
 
-## Future directions
-
-When these principles are not integrated correctly and checked, problems readily emerge
-[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
-This is why regular code testing is another recommended best practice, especially after making updates
-to the code or data [@Rule2019-eh; @The_Turing_Way_Community2019-ig].
-It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
-scientific community would fare reproducing such a workflow.
-
-Moreover, some have suggested adapting software development
-principles of unit tests and continuous integration/continuous delivery.
-This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
-made to a project's code base [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
-If the workflow executes successfully and tests are passed, the repository can be decorated with a
-"badge" to inform users of the lack of errors in the code.
-If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
-article can be assessed computationally, then such tests could be used to demonstrate computational
-reproducibility without the need for individuals to execute the workflow themselves.
-Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
-"temple" of reproducible research.
-
 <br>
@@
 
-Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
-of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
-These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
-significant expertise to analyze and distill key observations.
-The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
-reducing research waste.
-Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
-pollute the scientific literature.
+As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
+bioinformatics research; and current standards are falling short.
+If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
+such disasters from happening again.
 Researchers, their institutions, publishers and funders each have a major role to play in fighting the

### 2023-08-28 manuscript/rebuttal.Rmd
process validation

@@
 
-Response: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.
+RESPONSE: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.
 
@@
 RESPONSE: To demonstrate the importance of computational reprodcibility, we elaborate on the
-"forensic bioinformatics" case study of Baggerly and Coombes (2010) critique of Potti et al (2006)
-in the introduction:
+"forensic bioinformatics" case study of Baggerly and Coombes (2010)@Baggerly2010-nr critique of Potti
+et al (2006) @Potti2006-yk in the introduction:
 
@@
 
-In the introduction we have spoken to the consequences of poor computational research, and emphasize
-the urgency of this issue (which complements the case study summary above):
+RESPONSE: In the introduction we have spoken to the consequences of poor computational research, and
+emphasize the urgency of this issue (which complements the case study summary above):
 
@@
 
-We have re-written the Challenges section to highlight some of the barriers slowing the
+RESPONSE: We have re-written the Challenges section to highlight some of the barriers slowing the
 adoption of best practices, with reference to previous contributions on the topic.
@@
 funders, journals, institutions and individual researchers).
+We specifically note the effect of incentives on motivation, and the role of automation for
+routine reproducibility checking.
 
@@
 
-The subheading structure has been amended accordingly.
+RESPONSE: The subheading structure has been amended accordingly.
 
@@
   * Documentation
+
+  * Process validation & code testing
 
@@
 are covered in less detail and we refer the reader to those relevant reviews.
-What does make this article unique is the Five Pillars infographic, and specific combination of
-recommendations.
+What does make this article unique is the five pillars infographic, and specific combination of
+recommendations that incorporates new material (eg: Software Heritage, Guix, Continuous Analysis).
 Our article also provides a handy list of recommended tutorials as a supplementary resource.
@@
 
-> "... it needs to be formalised in code and ... "
+> ... it needs to be formalised in code and ...
 
@@
 > To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
-> possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
+> possible, from inspecting the raw data to generating the outputs that form the conclusions of the
+study.
 
@@
 > bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
-> [@The_Galaxy_Community2022-dt;@Reich2006-na].
+> [@The_Galaxy_Community2022-dt; @Reich2006-na].
 > Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
@@
 
-> Projects involving computationally intensive tasks might benefit from a build/workflow automation
-> solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
-> `nextflow` [@Di_Tommaso2017-yo].
+> Projects involving computationally intensive tasks would benefit from a build/workflow automation
+> solution.
+> Guides for selecting and using such systems have been already reviewed by others [@Leipzig2016-bf; @Reiter2021-lh].
+> But it is worth mentioning some of the most used solutions in bioinformatics, which include
+> `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021], `CWL` [@Crusoe2022-cb],
+> `WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
+> The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
+> for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
+> need to be repeated.
+> After fixing the issue, re-running the analysis would pick up at step 8 again,
+> which saves labor and compute time.
 
@@
 The bullet points following this passage outline the benefits of using DVCS.
-While cloud storage like dropbox could protect against code loss, they cannot other five points cannot
+While cloud storage like dropbox could protect against code loss, the other five points cannot
 be satisfied.
@@
 > `git` is typically used at the command line, however it is also incorporated into integrated
-> development environments commonly used in bioinformatics including Rstudio, JupyterLab [68,69] and
-> VS Code [70,71].
+> development environments commonly used in bioinformatics including Rstudio, JupyterLab
+> [@Kluyver2016jupyter] and VS Code [@noauthor_undated-fn].
 
@@
 
-RESPONSE: Biocontainers is mentioned in the last paragraph of the "Compute environment control" section
-and has reference (Ref #91).
+RESPONSE: Biocontainers is mentioned in the last paragraph of the "Compute environment control" section.
 
@@
 > suitable for genomics data due to the availability of existing public genetic data sets that heighten
-> reidentification risks [105].
+> reidentification risks [@Heeney2011-ri].
 > For example, the surnames of some male participants can be inferred based on the Y chromosome
-> variants seen in deidentified genomic data together with public geneology web sites [106].
+> variants seen in deidentified genomic data together with public geneology web sites [@Gymrek2013-fz].
 > To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
 > Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
-> frameworks designed to protect personal data in a way that perserves reproducibility [107,108].
+> frameworks designed to protect personal data in a way that perserves reproducibility [@Rehm2021-kt;@Thorogood2021-kt].
 
@@
 and outputs.
+
+We do not believe any changes to Figure 2 or 4 are necessary.
 
@@
 RESPONSE: While there may be problems with FAIRness with NCBI GEO and SRA, these repositries mostly
-satisfy FAIR principles (eg: https://www.biorxiv.org/content/10.1101/739334v1.full.pdf).
+satisfy FAIR principles (eg: @Bonaretti2019-zs).
 From the perspective of a researcher sharing genomic data, they are "good enough" as compared to the
 well known alternatives (eg: general purpose repository, institutional data storage, supplementary file,
-dropbox, etc.).
+dropbox, on request, etc.).
 
@@
 
-In light of this argument, we have added a citation to Caetano and Aisenberg in the updated text.
+In light of this argument, we have added citations to Caetano and Aisenberg in the updated text.
 
@@
 programming languages without needing to install extra packages.
-In line with this, the point in question has been amended:
+In line with this, the passage been amended:
 
@@
 > `WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
-> The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
-> for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
-> need to be repeated.
-> After fixing the issue, re-running the analysis would pick up at step 8 again,
-> which saves labor and compute time.
 

### 2023-08-28 manuscript/5pillars.Rmd
process validation

@@
 text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
-text(5,7,lab="Process validation & code testing",cex=1.2)
+text(5,7,lab="Process validation & testing",cex=1.2)
 
@@
 
+In bioinformatics, sharing code is becoming standard practice for reproducibility and transparency,
+and is a requirement for many specialized journals [@Cadwallader2022-dq].
+Code sharing appears to improve the rate of article citations [@Maitner2023-pv].
+One of the most popular ways to share research code is through online software repositories
+[@Cadwallader2022-od] with integrated version control.
 A version control system (sometimes called "source control") is a type of program that "tracks changes"
@@
 A recently described set of recommendations for making software FAIR describes further measures
-that data analysts could use to enhance their code documentation.
+that data analysts could use to enhance their code documentation @Patel2023-qr.
 
@@
 
-When these principles are not integrated correctly and checked, problems readily emerge
+When these principles are not integrated correctly and checked, problems can readily emerge
 [@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
-This is why regular code testing is another recommended best practice, especially after making updates
-to the code or data [@Rule2019-eh; @The_Turing_Way_Community2019-ig].
-It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
-scientific community would fare reproducing such a workflow.
-
-Moreover, some have suggested adapting software development
-principles of unit tests and continuous integration/continuous delivery.
-This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
-made to a project's code base [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
+This is why regular code testing after making updates to the code or data is considered best practice
+[@Rule2019-eh; @The_Turing_Way_Community2019-ig; @Noor2022-uc].
+For bioinformatics software, testing may involve a suite of unit tests of each function, and another
+integration test whether the functions are working properly together [@Sarma2016-ol].
+
+For data analysts, the situation is somewhat different as a typical bioinformatics workflow might
+involve hundreds of lines of code, so conducting tests at each step in the process would be impractical.
+However incorporating such tests at key points in a workflow such as during quality control of input
+data, after data cleaning, before statistical analysis and a summary of the findings.
+While `if/else` statements can be used for testing, it is recommended to use a dedicated testing
+package such as `testthat` for R [@Wickham2011-tt] or `pytest` for Python [@pytest7.4].
+In the example in Box 1, `testthat` raises an error to halt the script if the test fails, while
+the `if/else` statement requires an extra line to do this.
+
+```{r,test,class.source = 'fold-show'}
+
+data(iris)
+numspecies <- length(unique(iris$Species))
+
+# ifelse test
+if ( numspecies == 3) {
+  print("Iris data describes three species")
+} else {
+  stop("Iris data doesn't describe three species")
+}
+
+# testthat test
+library(testthat)
+data(iris)
+testthat::expect_equal(numspecies, 3)
+
+```
+
+*Box 1: An example of a test in R using an if/else statement and with the `testthat` package.
+The test is checking that the `iris` dataset describes three species.*
+
+Literate programming also provides an opportunity to record sanity checks in an output report.
+In R, commands like `dim()`, `length()`, `head()`, `str()` are a simple way to show the features of
+a dataset to validate that data transformation steps are working as expected.
+
+Successful execution of all literate scripts in a project is an important validation step
+that proves that the scripts are free from code errors.
+It is worthwhile to conduct such testing on a separate computer to understand how others would fare
+reproducing such a workflow.
+Testing extends to documentation as well, so it is important to update it so it remains in accordance
+with the codebase, and actively seek feedback from others on its suitability.
+
+In software development, continuous integration/continuous delivery (CI/CD) is a process of automated
+software compilation and testing.
+A derivation of this for data analysis has been termed "continuous analysis", and involves automatic
+execution and testing whenever a change is made to a project's code base
+[@Beaulieu-Jones2017-vj; @Blischak2019-dq].
 If the workflow executes successfully and tests are passed, the repository can be decorated with a

### 2023-08-28 manuscript/rebuttal.Rmd
close #29 reviewer comments addressed now

@@
 RESPONSE: To demonstrate the importance of computational reprodcibility, we elaborate on the
-"forensic bioinformatics" case study of Baggerly and Coombes (2010)@Baggerly2010-nr critique of Potti
+"forensic bioinformatics" case study of Baggerly and Coombes (2010) @Baggerly2010-nr critique of Potti
 et al (2006) @Potti2006-yk in the introduction:
@@
 
-  * Process validation & code testing
+  * Process validation and testing
 
@@
 recommendations that incorporates new material (eg: Software Heritage, Guix, Continuous Analysis).
+This is useful, because bioinformatics analysts can use the infographic to benchmark their projects
+and identify areas where changes are needed.
 Our article also provides a handy list of recommended tutorials as a supplementary resource.
@@
 
-> In this work, authors provided alongside the traditional research paper a Jupyter notebook ..
+> In this work, authors provided alongside the traditional research paper a Jupyter notebook ...
 
@@
 
-Figure 4 is adapted from the diagram on the Docker official website, which was originally published in
+RESPONSE: Figure 4 is adapted from the diagram on the Docker official website, which was originally published in
 2015 or earlier (https://web.archive.org/web/20151023003338/https://www.docker.com/what-docker).
@@
 We note that the original description of the FAIR principles by Wilkinson and colleagues is ~5,000 words,
-while this entire manuscript is ~6,500. We note that other reviewers did not insist on removing this
+while this entire manuscript is ~7,000. We note that other reviewers did not insist on removing this
 section.
@@
 
-RESPONSE: While there may be problems with FAIRness with NCBI GEO and SRA, these repositries mostly
+RESPONSE: While there may be problems with FAIRness with NCBI GEO and SRA, these repositories mostly
 satisfy FAIR principles (eg: @Bonaretti2019-zs).
@@
 
-RESPONSE: We acknowledge that NCBI is a well known repository for evolutionary biology.
+RESPONSE: We acknowledge that NCBI is a well known repository for molecular evolution.
 We have altered the section for clarity:
@@
 
-They also cite an article by Caetano and Aisenberg that speaks directly to the findability and
+They also cite an article by Caetano and Aisenberg that speaks directly to findability and
 accessibility @Caetano2014-bi:
@@
 RESPONSE: Thanks for raising this, and I note that similar sentiments were shared by reviewer 2 about the
-challenges that are holding back change.
+challenges that are holding back progress.
 Please refer to our response to Reviewer 2's Point #3.

### 2023-08-28 manuscript/5pillars.Rmd
close #29 reviewer comments addressed now

@@
 
+*Box 1: An example of a test in R using an if/else statement and with the `testthat` package.
+The test is checking that the `iris` dataset describes three species.*
+
+<hr>
+
 ```{r,test,class.source = 'fold-show'}
@@
 
-*Box 1: An example of a test in R using an if/else statement and with the `testthat` package.
-The test is checking that the `iris` dataset describes three species.*
+<hr>
 
@@
 [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
-If the workflow executes successfully and tests are passed, the repository can be decorated with a
-"badge" to inform users of the lack of errors in the code.
+If the workflow executes successfully and tests are passed, the repository and linked research article
+can be decorated with a "badge" to inform users of the lack of errors in the code,
+similar to badges used for other Open Science practices [@Kidwell2016-bb].
 If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
 article can be assessed computationally, then such tests could be used to demonstrate computational
-reproducibility without the need for individuals to execute the workflow themselves.
-Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
-"temple" of reproducible research.
+reproducibility without the need for individuals to execute the workflow themselves [@Suetake2022-zq].
 
@@
 The studies that have been conducted point to reproducibility rates of ~10% in computational
-life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
+life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further studies
 are required to bring attention to the issue.
@@
 This is a direct result of research assessment policies that incentivize journal-level prestige
-metrics and number of publications rather than the rigor [@Munafo2017-gf; @Oransky2023-zb].
+metrics and number of publications rather than rigor [@Munafo2017-gf; @Oransky2023-zb].
 A change in research assessment ("eg: track-record") that recognises and credits contributions toward

### 2023-08-28 manuscript/rebuttal.Rmd
minor spelling

@@
 
-> ... it needs to be formalised in code and ...
+> ... it needs to be formalized in code and ...
 
@@
 
-> To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
+> To ensure reproducibility of bioinformatics workflows, they need to be formalized in code wherever
 > possible, from inspecting the raw data to generating the outputs that form the conclusions of the
@@
 > As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
-> minimising the number of commands required.
+> minimizing the number of commands required.
 
@@
 
-| Container System | GitHub stars* |
-| --- | --- |
-| Docker | 72.2k |
-| Podman | 18.9k |
-| Singularity | 3.1k |
+Docker: 72.2k
 
-* 25th August.
+Podman: 18.9k
+
+Singularity: 3.1k
+
+(25th August).
 
@@
 > Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
-> frameworks designed to protect personal data in a way that perserves reproducibility [@Rehm2021-kt;@Thorogood2021-kt].
+> frameworks designed to protect personal data in a way that preserves reproducibility [@Rehm2021-kt;@Thorogood2021-kt].
 
@@
 RESPONSE: We acknowledge that it can sometimes be tiresome to reiterate things like the FAIR principles.
-However summarising the key points succinctly will be in the interests of readers with limited time.
+However summarizing the key points succinctly will be in the interests of readers with limited time.
 We note that the original description of the FAIR principles by Wilkinson and colleagues is ~5,000 words,

### 2023-08-28 manuscript/5pillars.Rmd
minor spelling

@@
 
-To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
+To ensure reproducibility of bioinformatics workflows, they need to be formalized in code wherever
 possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
@@
 Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
-frameworks designed to protect personal data in a way that perserves reproducibility
+frameworks designed to protect personal data in a way that preserves reproducibility
 [@Rehm2021-kt;@Thorogood2021-kt].
@@
 As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
-minimising the number of commands required.
+minimizing the number of commands required.
 This also makes the `README` documentation much simpler.
@@
 metrics and number of publications rather than rigor [@Munafo2017-gf; @Oransky2023-zb].
-A change in research assessment ("eg: track-record") that recognises and credits contributions toward
+A change in research assessment ("eg: track-record") that recognizes and credits contributions toward
 reproducibility and rigor would help decrease funds wasted on sloppy research.

### 2023-08-29 main.sh
rocker not working due to dependancy issue

@@ -4,6 +4,6 @@ git clone https://github.com/markziemann/5pillars.git
 
 cd 5pillars
 
-Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer"))'
+Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer","png"))'
 
 Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'

### 2023-08-29 test_build.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-docker run -it rocker/rstudio:4.3 bash -c "wget https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
+docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
 docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html . \
   && echo Build successful \
   || echo Error Build Failed

### 2023-08-29 main.sh
update dependencies

@@ -4,6 +4,6 @@ git clone https://github.com/markziemann/5pillars.git
 
 cd 5pillars
 
-Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer","png"))'
+Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer","png","shape","testthat"))'
 
 Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'

### 2023-08-29 test_build.sh
cache problem?

@@ -1,5 +1,5 @@
 #!/bin/bash
-docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
+docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget --no-check-certificate --no-cache --no-cookies https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
 docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html . \
   && echo Build successful \
   || echo Error Build Failed

### 2023-08-29 helper.sh
updating and simplifying docs

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+git clone https://github.com/markziemann/5pillars.git
+
+cd 5pillars
+
+Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer","png","shape","testthat"))'
+
+Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'

### 2023-08-29 main.sh

@@ -1,9 +1,6 @@
 #!/bin/bash
-
-git clone https://github.com/markziemann/5pillars.git
-
-cd 5pillars
-
-Rscript -e 'install.packages(c("rmarkdown","rmdformats","RColorBrewer","png","shape","testthat"))'
-
-Rscript -e 'rmarkdown::render("manuscript/5pillars.Rmd")'
+docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget --no-check-certificate --no-cache --no-cookies https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
+docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html . \
+  && echo Build successful \
+  || echo Error Build Failed
+[[ $(hostname) == "ziemann-01" ]] && cp 5pillars.html ~/public/5pillars/

### 2023-08-29 test_build.sh

@@ -1,6 +1,0 @@
-#!/bin/bash
-docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget --no-check-certificate --no-cache --no-cookies https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
-docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html . \
-  && echo Build successful \
-  || echo Error Build Failed
-[[ $(hostname) == "ziemann-01" ]] && cp 5pillars.html ~/public/5pillars/

### 2023-08-29 main.sh
update docs and main script

@@ -1,5 +1,5 @@
 #!/bin/bash
-docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget --no-check-certificate --no-cache --no-cookies https://raw.githubusercontent.com/markziemann/5pillars/main/main.sh ; bash main.sh"
+docker run -it bioconductor/bioconductor:RELEASE_3_17 bash -c "wget --no-check-certificate --no-cache --no-cookies https://raw.githubusercontent.com/markziemann/5pillars/main/helper.sh ; bash helper.sh"
 docker cp $(docker ps -aql):/5pillars/manuscript/5pillars.html . \
   && echo Build successful \
   || echo Error Build Failed

### 2023-08-30 manuscript/5pillars.Rmd
improvements to fig 1, fig 4

@@
 research.
-These include (1) code version control and sharing, (2) compute environment control,
-(3) literate programming, (4) documentation and (5) FAIR data sharing.
-These best practices will ensure that computational research work can be
+These include (1) literate programming, (2) code version control and sharing,
+(3) compute environment control, (4) FAIR and persistent data sharing and (5) documentation.
+These practices will ensure that computational research work can be
 reproduced quickly and easily, long into the future.
@@
 tarnishing the reputation of associated institutions and colleagues.
-In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.
+In clinical research, irreproducible bioinformatics has the potential to place patient
+safety at risk.
 
@@
 text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
-text(5,7,lab="Process validation & testing",cex=1.2)
+text(5,7,lab="Continuous validation",cex=1.2)
 
@@
 polygon(c(6,11.5,11.5,6),c(0,0,0.9,0.9), col=cols2[1], border = NA)
-polygon(c(6,11.5,11.5,6),c(1,1,1.9,1.9), col=cols2[3], border = NA)
-
-labs3 <- c("Hardware","Hypervisor")
-text(c(8.5,8.5),c(0.4,1.4),labels=labs3)
-
-polygon(c(6,7.8,7.8,6),c(2,2,4.5,4.5), col=cols2[2], border = NA)
-polygon(c(8,9.6,9.6,8),c(2,2,4.5,4.5), col=cols2[2], border = NA)
-polygon(c(9.8,11.5,11.5,9.8),c(2,2,4.5,4.5), col=cols2[2], border = NA)
-
-text(c(6.85,8.8,10.65),c(3.5,3.5,3.5),labels="Guest\nOperating\nSystem")
+polygon(c(6,11.5,11.5,6),c(1,1,1.9,1.9), col=cols2[2], border = NA)
+polygon(c(6,11.5,11.5,6),c(2,2,2.9,2.9), col=cols2[3], border = NA)
+
+labs3 <- c("Hardware","Host Operating System","Hypervisor")
+text(c(8.5,8.5,8.5),c(0.4,1.4,2.4),labels=labs3)
+
+polygon(c(6,7.8,7.8,6),c(3,3,4.5,4.5), col=cols2[2], border = NA)
+polygon(c(8,9.6,9.6,8),c(3,3,4.5,4.5), col=cols2[2], border = NA)
+polygon(c(9.8,11.5,11.5,9.8),c(3,3,4.5,4.5), col=cols2[2], border = NA)
+
+text(c(6.85,8.8,10.65),c(3.8,3.8,3.8),labels="Guest\nOperating\nSystem")
 
@@
 
-### Process validation and testing
+### Continuous validation
 
@@
 a dataset to validate that data transformation steps are working as expected.
+Histograms, boxplots, scatterplots and other simple charts are similarly effective.
 
@@
 [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
-If the workflow executes successfully and tests are passed, the repository and linked research article
+We prefer the term "continuous validation" as it places more emphasis on the correctness of analyses
+to ensure they are fit to share.
+If a workflow is executed successfully and tests are passed, the repository and linked research article
 can be decorated with a "badge" to inform users of the lack of errors in the code,
@@
 reproducibility and rigor would help decrease funds wasted on sloppy research.
+Such changes would be welcomed by industry as they would face fewer replication failures
+[@Rosenblatt2016-cq], which could speed development of research into useful products.
 

### 2023-08-31 manuscript/rebuttal.Rmd
close #37 thanks Pierre

@@
 
-RESPONSE: To demonstrate the importance of computational reprodcibility, we elaborate on the
+RESPONSE: To demonstrate the importance of computational reproducibility, we elaborate on the
 "forensic bioinformatics" case study of Baggerly and Coombes (2010) @Baggerly2010-nr critique of Potti
@@
 
-* Modifcation to the Rmd header section that specifies the path of the bib and csl files.
+* Modification to the Rmd header section that specifies the path of the bib and csl files.
 
@@
 > journal styles [@noauthor_undated-pg].
-> Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
-> freely available @Xie2018-rf.
+> Further, dozens of journal-style Markdown document templates are freely available, and integrate
+> with MyST notebook system or the `rticles` R package @Xie2018-rf.
 
@@
 programming languages without needing to install extra packages.
-In line with this, the passage been amended:
+In line with this and suggestions from others, the passage been amended:
 
 > 5. Use file formats that are machine-readable and compatible with many different types of software.
-> For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
-> specifications that are suitable for very large data.
+> Some examples include comma- and tab-separated values (CSV/TSV) formats, eXtensible Markup Language (XML),
+> JavaScript Object Notation (JSON), Hierarchical Data Format version 5 (HDF5) and Apache Parquet.
 

### 2023-08-31 manuscript/5pillars.Rmd
close #37 thanks Pierre

@@
 [ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
-@Stodden2016-aj; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Akalin2018-eg;
+@Stodden2016-aj; @Wilson2017-dv; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Kim2018-nv; @Akalin2018-eg;
 @Brito2020-ld; @Peikert2021-eq; @Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv].
@@
 that performed and documented the entire analysis procedure.
-The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
+The MyST (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
 and has some notable advantages around flexibility of content that make it an attractive choice for executable
@@
 journal styles [@noauthor_undated-pg].
-Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
-freely available @Xie2018-rf.
+Further, dozens of journal-style Markdown document templates are freely available, and integrate
+with MyST notebook system or the `rticles` R package @Xie2018-rf.
 
@@
 There are also several `git` clients with graphical interfaces which better allow inspection
-of code changes (e.g.: @noauthor_undated-rh, @noauthor_undated-kj).
+of code changes (e.g.: [@noauthor_undated-rh; @noauthor_undated-kj]).
 
@@
 genome assembly and have become well known to omics researchers
-[@Edgar2002-hs; @Kodama2012-bn; Martens2005-wo]).
+[@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
 The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
@@
 5. Use file formats that are machine-readable and compatible with many different types of software.
-For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
-specifications that are suitable for very large data.
+Some examples include comma- and tab-separated values (CSV/TSV) formats, eXtensible Markup Language (XML),
+JavaScript Object Notation (JSON), Hierarchical Data Format version 5 (HDF5) and Apache Parquet.
 
@@
 Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
-their reproducibility [@Nekrutenko2012; @Garijo2013-zl; @Simoneau2021-kp].
+their reproducibility [@Nekrutenko2012; @Garijo2013-zl; @Kim2018-nv; @Simoneau2021-kp].
 Commonly missing information includes versions of software and packages as well as any parameter setting
@@
 In addition, the `README` should outline the contents of the repository, how users can contribute to
-the codebase and how to report issues, such as code errors and discrepancies.
+the project and how to report issues, such as code errors and discrepancies.
 The `README` should also list the contributors to the codebase and acknowledge any financial support.
@@
 This is why regular code testing after making updates to the code or data is considered best practice
-[@Rule2019-eh; @The_Turing_Way_Community2019-ig; @Noor2022-uc].
+[@Kim2018-nv; @Rule2019-eh; @The_Turing_Way_Community2019-ig; @Noor2022-uc].
 For bioinformatics software, testing may involve a suite of unit tests of each function, and another
@@
 If a workflow is executed successfully and tests are passed, the repository and linked research article
-can be decorated with a "badge" to inform users of the lack of errors in the code,
+can be decorated with a "badge" to indicate computational reproducibility,
 similar to badges used for other Open Science practices [@Kidwell2016-bb].
@@
 the severity of the reproducibility problem isn't well appreciated.
-The studies that have been conducted point to reproducibility rates of ~10% in computational
+The studies that have been conducted point to reproducibility rates of ~20% in computational
 life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further studies
@@
 Addressing this problem is not straightforward.
-It requires ehancement of data science skills in undergraduate life science courses, investment in
-upskilling the current workforce and offering working conditions that are competitive with other sectors.
+It requires ehancement of reproducible data science skills in undergraduate life science courses,
+investment in upskilling the current workforce and offering working conditions that are competitive
+with other sectors.
 Specific workshop-based training for current researchers could help [@Williams2023-ic].
@@
 
-**Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
-He has recently been recruited to Burnet Institute as Head of Bioinformatics.
+**Mark Ziemann** is Head of Bioinformatics at Burnet Institute and former Lecturer in
+Biotechnology and Bioinformatics at Deakin University.
 His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects

### 2023-08-31 manuscript/5pillars.Rmd
spellcheck american english

@@
 
-* Correspondance: m.ziemann@deakin.edu.au
-
-| Author | ORCID |
+* Correspondence: m.ziemann@deakin.edu.au
+
+| `Author      ` |  `ORCID             ` |
 | --- | --- |
@@
 results.
-Robustness refers to the ability for the findings to be generalisable to other situations
+Robustness refers to the ability for the findings to be generalizable to other situations
 @Nosek2022-zc.
@@
 Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
-techniques to piece together how the dataset was originally analysed.
+techniques to piece together how the dataset was originally analyzed.
 Their investigations found a litany of issues.
@@
 data analysis such as using spreadsheets, and these problems can be obscured by a lack of documentation.
-The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
+The article underwent two corrigenda, but was ultimately retracted in 2011, as the authors were not
 able to reproduce the findings themselves due to "corruption of several validation data sets"
@@
 share parts of the host operating system [@Nust2020-cr].
-In the example shown in **Figure 4**, the five containerised applications share the same operating
+In the example shown in **Figure 4**, the five containerized applications share the same operating
 system, while the three VM applications each involve their own operating system which incurs a
@@
 across different systems.
-For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
+For example, the most popular containerization system, Docker [@Merkel2014-ek], makes it
 possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
@@
 While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
-[@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
+[@Kurtzer2017-zq], Docker remains the most widely used containerization system (according to GitHub
 stars as of August 2023).
@@
 a dataset to validate that data transformation steps are working as expected.
-Histograms, boxplots, scatterplots and other simple charts are similarly effective.
+Histograms, box plots, scatter plots and other simple charts are similarly effective.
 
@@
 
-Journals are partly responsible, as their editorial policies influence author behaviours [@Han2017-pz].
+Journals are partly responsible, as their editorial policies influence author behaviors [@Han2017-pz].
 Except for a few outliers, verification of computational reproducibility is not in the scope of peer
@@
 [@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
-This makes it challenging for organisations to recruit and retain capable data analysts and cultivate
+This makes it challenging for organizations to recruit and retain capable data analysts and cultivate
 a community of practice.
 Addressing this problem is not straightforward.
-It requires ehancement of reproducible data science skills in undergraduate life science courses,
-investment in upskilling the current workforce and offering working conditions that are competitive
+It requires enhancement of reproducible data science skills in undergraduate life science courses,
+investment in up-skilling the current workforce and offering working conditions that are competitive
 with other sectors.
@@
 
-A lack of documentated protocols and tutorial materials for highly reproducible analysis makes it
+A lack of documented protocols and tutorial materials for highly reproducible analysis makes it
 difficult for researchers to confidently update their code development practices.
@@
 This approach is highly scalable and the reach of open-access publishing ensures these materials are
-broadly accesible.
+broadly accessible.
 For example, our group has just released such a protocol that walks novice users through the process
@@
 bioinformatics research; and current standards are falling short.
-If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
+If adopted widely, the transparency and reproducibility measures proposed here will reduce the chance of
 such disasters from happening again.
@@
 
-## Acknowledgements
+## Acknowledgments
 

### 2023-09-21 manuscript/rebuttal.Rmd
2nd round comments

@@
 > Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
-> techniques to piece together how the dataset was originally analysed.
+> techniques to piece together how the dataset was originally analyzed.
 > Their investigations found a litany of issues.
@@
 > data analysis such as using spreadsheets, and such problems can be obscured by a lack of documentation.
-> The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
+> The article underwent two corrigenda, but was ultimately retracted in 2011, as the authors were not
 > able to reproduce the findings themselves due to "corruption of several validation data sets"
@@
 
-  * Process validation and testing
+  * Continuous validation
 
@@
 > bioinformatics research, and current standards are falling short.
-> If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
+> If adopted widely, the transparency and reproducibility measures proposed here will reduce the chance of
 > such disasters from happening again.
@@
 while others do not mention the importance of documentation.
-Some are focused on specific concepts like infrastructure, containerisation or
+Some are focused on specific concepts like infrastructure, containerization or
 continuous analysis.
@@
 * A .csl file that specifies the citation style.
-These are avaialble from [GitHub](https://github.com/citation-style-language/styles).
+These are available from [GitHub](https://github.com/citation-style-language/styles).
 
@@
 RESPONSE: The claim made in the manuscript is that "Docker remains the most widely used
-containerisation system for web development and research".
+containerization system for web development and research".
 We do not have exact usage data on this, however we can look at other metrics like GitHub stars.
@@
 > While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
-> [@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
+> [@Kurtzer2017-zq], Docker remains the most widely used containerization system (according to GitHub
 > stars).
@@
 RESPONSE: Thanks for raising this. We have added CWL to the section mentioning pipeline solutions, and
-refered readers to dedicated reviews on the topic.
+referred readers to dedicated reviews on the topic.
 
@@
 The Challenges section now includes brief mentions of incentives and automation.
+
+## Bibliography

### 2023-09-21 manuscript/rebuttal2.Rmd
2nd round comments

---
title: "Response to reviewer comments, part II (The five pillars)"
author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
date: "`r Sys.Date()`"
output: word_document
code_folding: hide
fig_width: 7
fig_height: 6
bibliography: 5pillars.bib
csl: plos-computational-biology.csl
---

Dear Editor,

Again, many thanks for your consideration of our manuscript.
Below is our point-by-point response to Reviewer 3's remining comments.

### Point 2

> In my view, the paper is not suitable for "anyone who analyzes biological data." After all, the
> journal is Briefings in Bioinformatics.
> Much of the material is too technical for the stated audience.
> The authors have provided "Practical guides for implementing ‘Five Pillar’ principles."
> This may be helpful to some, but it is not realistic to assume that people will consume all of
> those materials before reading the paper or that by doing so they will be able to understand all of
> the technical details in the manuscript.
> In my view, the intended audience should be clarified in the manuscript.

RESPONSE: Indeed much of the manuscript is oriented towards gold-standard practices and this could be
out of the immediate reach for many biological data analysts, but this is not a futile exercise.
Firstly, papers such as this one set out a road map of self-education which may take years to
accomplish.
Secondly, where self-education reaches its limits, they can seek workshops and other training to
strengthen these priority areas.

Therefore, we have altered the phrase to recognise the journey of bioinformaticians in training.

> The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
> principles described here could equally apply to other domains of study.

If the reviewer has a specific suggestion, then it would be helpful to provide this.

### Point 10

> 10. What about when your data files are too large to fit on a personal computer? Please address in
the manuscript.

RESPONSE: In accordance with this suggestion we have added the following passage to the literate
programming section.

> While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
> they can also be executed in "headless" mode on high performance computing clusters or cloud
> servers using command-line interfaces to take advantage of greater computational resources.
> Headless mode is important for integrating literate scripts into larger workflows and for
> automated testing.

### Point 17

> 17. The Biocontainers project should be mentioned.

RESPONSE: We have added an additional mention of BioContainers which appears just after where DockerHub
is mentioned.

> In bioinformatics, containers are already extensively used.
> For example, BioContainers is a registry for sharing bioinformatics software containers
> [@Da_Veiga_Leprevost2017-yw].
> Containers are equally useful to data analysts by encapsulating the environment in which analytical
> scripts are executed.

### Point 20

> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles.
> People can go to the source article for that.
> I don't mean to be snarky, but everyone has limited time. I still believe that spelling out these
> points only serves to make the manuscript longer than necessary and thus make it less likely for
> people with "limited time" to read it.
> FAIRness is not the main thrust of this manuscript.

RESPONSE: We appreciate the reviewer's point of view.
We have removed mention of FAIR and we refer readers to the literature on this topic.
FAIR has also been removed from the pillars infographic and Figure 5 is also gone.

### R3 Point 21

> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
There are lots of problems with FAIRness in these repositories.

RESPONSE: Possibly the reviewer hasn't seen the rebuttal letter had it's own bibliography.
Anyway, since "FAIR" is not described anymore, it simplifies the description of these databases,
which satify most best practices. We cite the work of Bonaretti et al 2019.
We also cite Hassenruck et al 2021, who report deficiencies in these databases, but spread the blame
between database administrators, researchers and other stakeholders.
The two important passages we have amended are provided below:

> In the wake of the Human Genome Project, a number of repositories for specific biological data types
> were established including GEO, SRA, ENA and PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
> Although these repositories suffer issues around interoperability and reusability, they do support
> data reuse through findability and accessibility [@Bonaretti2019-zs; @Hassenruck2021-mi].

...

> Researchers should be aware, however, that these repositories are often not moderated,
> and that it is their responsibility to provide sufficient metadata.

## Bibliography

### 2023-09-21 guides/practical_guides.Rmd
documentation edits
2nd round comments

@@
 
-Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>,
-Anusuiya Bora<sup>1</sup>
-
-Author information
-
-1. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.
-
-2. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.
-
-Correspondance: m.ziemann@deakin.edu.au
-
-| Author | ORCID |
-| --- | --- |
-| Mark Ziemann | 0000-0002-7688-6974 |
-| Pierre Poulain | 0000-0003-4177-3619 |
-| Anusuiya Bora | 0009-0006-2908-1352 |
-
-</details>
-</p>
-
 Source: https://github.com/markziemann/5pillars/blob/main/guides/practical_guides.Rmd
 
-The five pillar approach is a synthesis of over a decade of learnings around computational
-reproducibility, and contains six key recommendaions and many best practices.
+The five pillar approach is a synthesis of over a decade of learning around computational
+reproducibility, and contains seven key recommendations and many best practices.
 These may be overwhelming for any data scientists, especially beginners and novices,
@@
 
-### Introduction to Unix and Shell programming
+### Introduction to Unix and shell programming
 
@@
 <br>
+
+### Introduction to Galaxy: bioinformatics in the browser
+
+* [Galaxy: A very short introduction](https://www.youtube.com/watch?v=VZoz3k5EehI&ab_channel=GalaxyProject)
 
@@
 * [Tutorial: The Ultimate Beginner’s Guide to Jupyter Notebooks](https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af)
+
+* [Tutorial: Write your next paper in MyST Markdown with data, code & Jupyter notebooks](https://www.youtube.com/watch?v=1BWYLzQd8Ds&t=4s&ab_channel=curvenote)
 
@@
 
+## Practical guides for Conda, Guix and Docker
+
+Control the environment.
+Record the version of programming languages and packages.
+Become familiar and experiment with Conda, Guix and Docker.
+Select the option that works best for your project.
+
+* [Tutorial: Getting started with conda](https://conda.io/projects/conda/en/latest/user-guide/getting-started.html)
+
+* [Video tutorial: Anaconda Beginners Guide for Linux and Windows](https://www.youtube.com/watch?v=MUZtVEDKXsk)
+
+* [Tutorial: Running RStudio Server with Docker](https://davetang.org/muse/2021/04/24/running-rstudio-server-with-docker/)
+
+* [Tutorial: Creating a Docker Image for your Application](https://www.stereolabs.com/docs/docker/creating-your-image/)
+
+* [Documentation: Getting started with GNU Guix](https://guix.gnu.org/manual/en/html_node/Getting-Started.html)
+
+<br>
+
 ## Practical guides to data sharing
@@
 * [Documentation: FAIRsharing.org](https://fairsharing.org/)
+
+<br>
+
+## Practical guides for documenting computational research
+
+Document the process of reproduction as if you were an external researcher.
+Test the instructions.
+
+* [Guide: How to write an Awesome README](https://towardsdatascience.com/how-to-write-an-awesome-readme-68bf4be91f8b)
+
+* [Ebook: Bioinformatics Data Skills Chapter 2](https://vincebuffalo.com/book/)
+
+* [Example: README for "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker"](https://github.com/aaronpeikert/reproducible-research/blob/master/README.md)
+
+* [Guide: A Quick Guide to Organizing Computational Biology Projects](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1000424)
 
@@
 
-## Practical guides for Conda, Guix and Docker
-
-Control the environment.
-Record the version of programming languages and packages.
-Become familiar and experiment with Conda, Guix and Docker.
-Select the option that works best for your project.
-
-* [Tutorial: Getting started with conda](https://conda.io/projects/conda/en/latest/user-guide/getting-started.html)
-
-* [Video tutorial: Anaconda Beginners Guide for Linux and Windows](https://www.youtube.com/watch?v=MUZtVEDKXsk)
-
-* [Tutorial: Running RStudio Server with Docker](https://davetang.org/muse/2021/04/24/running-rstudio-server-with-docker/)
-
-* [Tutorial: Creating a Docker Image for your Application](https://www.stereolabs.com/docs/docker/creating-your-image/)
-
-* [Documentation: Getting started with GNU Guix](https://guix.gnu.org/manual/en/html_node/Getting-Started.html)
-
-<br>
-
-## Practical guides for documenting computational research
-
-Document the process of reproduction as if you were an external researcher.
-Test the instructions.
-
-* [Guide: How to write an Awesome README](https://towardsdatascience.com/how-to-write-an-awesome-readme-68bf4be91f8b)
-
-* [Ebook: Bioinformatics Data Skills Chapter 2](https://vincebuffalo.com/book/)
-
-* [Example: README for "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker"](https://github.com/aaronpeikert/reproducible-research/blob/master/README.md)
-
-<br>
-
-## Moving from "dev" to "prod": testing and continuous analysis
+## Continuous analysis/continuous validation
 
@@
 errors.
+Transparency and reproducibility help validation which can be continously maintained.
 
@@
 
+* [Tool: Drone: System for continuous delivery](https://www.drone.io/)
+
 * [Documentation: GitHub Actions](https://docs.github.com/en/actions)

### 2023-09-21 manuscript/5pillars.Rmd
2nd round comments

@@
 
-* Correspondence: m.ziemann@deakin.edu.au
-
-| `Author      ` |  `ORCID             ` |
+(*) Correspondence: m.ziemann@deakin.edu.au
+
+| Author | ORCID |
 | --- | --- |
@@
 These include (1) literate programming, (2) code version control and sharing,
-(3) compute environment control, (4) FAIR and persistent data sharing and (5) documentation.
+(3) compute environment control, (4) persistent data sharing and (5) documentation.
 These practices will ensure that computational research work can be
@@
 transparent reporting and the role of computational environments.
-The intended audience is anyone who analyzes biological data (not tool developers), however the
+The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
 principles described here could equally apply to other domains of study.
@@
   "Compute \nenvironment \ncontrol",
-  "FAIR &\npersistent\ndata sharing",
+  "Persistent\ndata sharing",
   "Documentation")
@@
 [@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
-A typical genomics/bioinformatics study could involve one large literate script, or be broken down into
-smaller scripts, where each one contains the code for generating a part of the article.
+A typical genomics/bioinformatics study could involve one large literate script, or be broken down intosmaller scripts, where each one contains the code for generating a part of the article.
 A multi-script approach would benefit from a master script that executes each component.
 This further makes the job of reproducibility easier.
+
+While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
+they can also be executed in "headless" mode on high performance computing clusters or cloud
+servers using command-line interfaces to take advantage of greater computational resources.
+Headless mode is important for integrating literate scripts into larger workflows and for
+automated testing.
 
@@
 One of the most popular ways to share research code is through online software repositories
-[@Cadwallader2022-od] with integrated version control.
+with integrated version control [@Cadwallader2022-od].
 A version control system (sometimes called "source control") is a type of program that "tracks changes"
 made to sets of files, typically other computer program source code, scripts and documentation.
-A set of files under version control is called a "repository" and represents a project or sub-project.
+A repository is simply a set of files under version control that represents a project or sub-project.
 Version control is used extensively by software developers and is considered one of the key best
@@
 reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].
+In bioinformatics, containers are already extensively used.
+For example, BioContainers is a registry for sharing bioinformatics software containers
+[@Da_Veiga_Leprevost2017-yw].
+Containers are equally useful to data analysts by encapsulating the environment in which analytical
+scripts are executed.
 
@@
 
-### FAIR and persistent data sharing
+### Persistent data sharing
 
@@
 @Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
-(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]), although there are some challenges.
-Some researchers would rather keep data for their private use.
-Some may keep data private due to concerns over participant privacy.
-
-While deidentification of data has been widely used to safeguard participant privacy, this may not
+(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]).
+Best practices for research data sharing have been proposed [@Wilkinson2016-qq], as have
+recommendations around sharing human genetic/genomic data [@Byrd2020-ri].
+While deidentification of data has long been used to safeguard participant privacy, this may not
 be suitable for genomics data due to the availability of existing public genetic data sets that heighten
@@
 frameworks designed to protect personal data in a way that preserves reproducibility
-[@Rehm2021-kt;@Thorogood2021-kt].
-
-In order to maximize the value of shared data for reuse and reproducibility, it needs to be Findable,
-Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
-These principles are summarized in **Figure 5**.
-
-![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb.](../img/gr2.jpg)
-
-High throughput biology has already built a number of data repositories for specific data types mostly
-conforming to FAIR principles.
-Repositories like GEO, SRA, ENA and PRIDE were developed in the aftermath of the draft human
-genome assembly and have become well known to omics researchers
-[@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
+[@Rehm2021-kt; @Thorogood2021-kt].
+
+In the wake of the Human Genome Project, a number of repositories for specific biological data types
+were established including GEO, SRA, ENA and PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
+Although these repositories suffer issues around interoperability and reusability, they do support
+data reuse through findability and accessibility [@Bonaretti2019-zs; @Hassenruck2021-mi].
 The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
-FAIR repositories that accepts data from other domains of study like ecology,
+repositories that accepts data from other domains of study like ecology,
 physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
@@
 [Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
-should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
-and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
-To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
+should be considered.
+Researchers should be aware, however, that these repositories are often not moderated,
+and that it is their responsibility to provide sufficient metadata.
+To this aim, general recommendations have been proposed to enhance data reuse in ecology, but which
 can also be applied to other disciplines [@Roche2015-aq].
@@
 license.
-A recently described set of recommendations for making software FAIR describes further measures
+A recently described set of recommendations for documenting software describes further measures
 that data analysts could use to enhance their code documentation @Patel2023-qr.

### 2023-09-21 manuscript/rebuttal2.Rmd
addressed rev3 comments

@@
 Again, many thanks for your consideration of our manuscript.
-Below is our point-by-point response to Reviewer 3's remining comments.
+Below is our point-by-point response to Reviewer 3's remaining comments.
 
@@
 
-Therefore, we have altered the phrase to recognise the journey of bioinformaticians in training.
+Therefore, we have altered the phrase in the abstract and body to recognize the journey of
+bioinformaticians in training.
 
+ABSTRACT:
+> This guide is designed for bioinformatics data analysts and bioinformaticians in training,
+> but should be relevant to other domain of study.
+
+RECOMMENDATIONS:
 > The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
@@
 
-> While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
+> While it is common to run literate programs on a PC with graphical interfaces such as RStudio,
 > they can also be executed in "headless" mode on high performance computing clusters or cloud
 > servers using command-line interfaces to take advantage of greater computational resources.
-> Headless mode is important for integrating literate scripts into larger workflows and for
+> Headless mode is also important for integrating literate scripts into larger workflows and for
 > automated testing.
@@
 Anyway, since "FAIR" is not described anymore, it simplifies the description of these databases,
-which satify most best practices. We cite the work of Bonaretti et al 2019.
+which satisfy most best practices. We cite the work of Bonaretti et al 2019.
 We also cite Hassenruck et al 2021, who report deficiencies in these databases, but spread the blame

### 2023-09-21 manuscript/5pillars.Rmd
addressed rev3 comments

@@
 reproduced quickly and easily, long into the future.
-This guide is designed for bioinformatic data analysts, but should be relevant for other
-domains.
+This guide is designed for bioinformatics data analysts and bioinformaticians in training,
+but should be relevant to other domains of study.
 
@@
 Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
-which would be difficult for graphical tools like spreadsheets or webtools.
+which would be difficult for graphical tools like spreadsheets or web tools.
 Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
@@
 
-While webtools are valuable for data exploration, there are worries that they undermine
+While web tools are valuable for data exploration, there are worries that they undermine
 reproducibility for the sake of convenience [@Wijesooriya2022-ly].
@@
 [@The_Galaxy_Community2022-dt;@Reich2006-na].
-Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
+Some web tools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
 or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).
@@
 [@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
-A typical genomics/bioinformatics study could involve one large literate script, or be broken down intosmaller scripts, where each one contains the code for generating a part of the article.
+A typical genomics/bioinformatics study could involve one large literate script, or be broken down
+into smaller scripts, where each one contains the code for generating a part of the article.
 A multi-script approach would benefit from a master script that executes each component.
@@
 
-While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
+While it is common to run literate programs on a PC with graphical interfaces such as RStudio,
 they can also be executed in "headless" mode on high performance computing clusters or cloud
 servers using command-line interfaces to take advantage of greater computational resources.
-Headless mode is important for integrating literate scripts into larger workflows and for
+Headless mode is also important for integrating literate scripts into larger workflows and for
 automated testing.
@@
 practices in software engineering.
-Distributed version control systems (DVCS) involve a central web-accessible server hosting a
+Distributed Version Control Systems (DVCS) involve a central web-accessible server hosting a
 repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
@@
 `git` is typically used at the command line, however it is also incorporated into integrated development
-environments commonly used in bioinformatics including Rstudio, JupyterLab
+environments commonly used in bioinformatics including RStudio, JupyterLab
 [@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].

### 2023-09-22 manuscript/rebuttal2.Rmd
typofix

@@
 strengthen these priority areas.
-
 Therefore, we have altered the phrase in the abstract and body to recognize the journey of
@@
 ABSTRACT:
+
 > This guide is designed for bioinformatics data analysts and bioinformaticians in training,
@@
 RECOMMENDATIONS:
+
 > The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
@@
 
-If the reviewer has a specific suggestion, then it would be helpful to provide this.
+If the reviewer has a specific suggestion, then it would be helpful to provide it.
 

### 2023-09-26 manuscript/rebuttal2.Rmd
acceptance

@@
 Below is our point-by-point response to Reviewer 3's remaining comments.
+Please note that this rebuttal letter has its own bibliography and the numbering will differ to the
+main text.
 
@@
 > This guide is designed for bioinformatics data analysts and bioinformaticians in training,
-> but should be relevant to other domain of study.
+> but should be relevant to other domains of study.
 
@@
 RESPONSE: We appreciate the reviewer's point of view.
-We have removed mention of FAIR and we refer readers to the literature on this topic.
+We have reduced the discussion of FAIR principles to just one sentence.
+
+> "In order to maximize the value of shared data for reuse and reproducibility, it needs to be Findable,
+> Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq."
+
 FAIR has also been removed from the pillars infographic and Figure 5 is also gone.
@@
 
-RESPONSE: Possibly the reviewer hasn't seen the rebuttal letter had it's own bibliography.
-Anyway, since "FAIR" is not described anymore, it simplifies the description of these databases,
+RESPONSE: Possibly the reviewer hasn't seen that the rebuttal letter had it's own bibliography.
+Anyway, since "FAIR" is not described in depth anymore, it simplifies the description of these databases,
 which satisfy most best practices. We cite the work of Bonaretti et al 2019.

### 2023-09-27 manuscript/5pillars.Rmd
acceptance
slight fix

@@
 ```
+
 Another potential solution to this problem is to use a package/environment management system such as
@@
 
+In order to maximize the value of shared data for reuse and reproducibility, it needs to be Findable,
+Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
 In the wake of the Human Genome Project, a number of repositories for specific biological data types
